### 2008-03-09 proposal.tex
Initial commit of research proposal.

\documentclass{acm_proc_article-sp}

\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{appendix}
\usepackage{float}

\floatstyle{ruled}
\newfloat{program}{thp}{lop}
\floatname{program}{Program}

\begin{document}

\title{Evaluation of Cache-Oblivious Data Structures\\Draft research proposal}

\numberofauthors{1}
\author{Maks Verver\\ \email{m.verver@student.utwente.nl}}

% Obligatory permission block
\toappear{
Permission to make digital or hard copies of all or part of this work
for personal or classroom use is granted without fee provided that
copies are not made or distributed for profit or commercial advantage
and that copies bear this notice and the full citation on the first
page. To copy otherwise, or republish, to post on servers or to
redistribute to lists, requires prior specific permission.

\textit{\small 8$^{th}$ Twente Student Conference on IT, Enschede, June, 2008}

Copyright 2008, University of Twente, Faculty of Electrical Engineering,
Mathematics and Computer Science}
% End of obligatory permission block

\maketitle

\keywords{cache efficiency,locality of reference,algorithms}

\section{Introduction}
A fundamental part of theoretical computer science is the study of algorithms: formal descriptions of how computations may be performed. In order for algorithms to be able to process information efficiently, it must be stored in a systematic way, called a data structure. Algorithms and data structures therefore complement each other: while algorithms describe how a computer may process information, data structures describe how information is stored in a (model of a) computer's memory.

Traditionally, algorithms have been evaluated in a simplified model of computation. In this model it is assumed that a computer executes an algorithm in discrete steps, and at each step it performs one elementary operation (e.g. comparing two numbers, adding one to another, storing a value in memory, et cetera). Each elementary operation is performed within a constant time. In this model, both storing and retrieving data values in memory is considered to be an elementary operation.

This model is close enough to the way computers work to be extremely useful in the development and analysis of data structures and algorithms that work well in practice. However, like every model, it is a simplification of reality. One of the simplifications of the model that has become more detached from reality, is the uniform memory model: the assumption that storing and retrieving data in memory is a constant-time operation.

The main reasons this simplification does not hold in practice is that processors can process data much faster than it can be read from (or written to) memory. To overcome the memory bottleneck, system designers have added faster cache memory at various points in the computer system. This has resulted in hierarchical ordering of various types of memory, each faster but smaller than the next, such as the processor's registers, L1 cache on the processor die, L2 cache, main memory, magnetic disk storage and sometimes network storage.

[TODO: add picture of memory hierarchy]

[TODO: more detailed discussion of how fast memory acts as a cache for slow memory, data is transfered and blocks, and how this leads to algorithms that have good temporal and spatial locality to have superior performance.]

Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. To be able to analyze these algorithms and data structures formally, a different model of computation and especially of the cost of accessing data in memory is required.

\subsection{External memory model}
In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited). The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transfered between the two memories. The advantage of using such a model is that it more accurately reflects reality yet still allows formal analysis of the performance of various algorithms and data structures. Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy. In that case, the possibility of transfering several data blocks at once is usually dropped.

\subsection{Cache-oblivious memory model}
The external memory model can be simplified and extended to a cache-oblivious model (\cite{prokop1999coa}) which recognizes the same distinction between internal and external memory, but does not assume the relevant parameters (size of the cache and size of the data blocks) are known. Therefore, cache-oblivious data structures and algorithms must be defined without explicitly referencing these parameters, yet still perform well when evaluated in the cache-oblivious model. Algorithms that perform well in the cache-oblivious model are more general (since they have no parameters that must be specified) and by design tuned to all levels in the memory hierarchy, which means they should perform well in practice without any tuning.

Erik D. Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms (\cite{demaine2002coa}). He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.

\subsection{Categories}
The development of the different memory models means data structures and algorithms can be classified depending on the assumptions about their environment used in their definitions.
\begin{list}{}{}
\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache lay-out (such as block size or cache memory size).
\item \textbf{Cache-unaware} [does a better term exist?] data structures and algorithms  are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotical sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
\end{list}

\section{Related work}
Many data structures and algorithms have been analyzed in the cache-oblivious model. In many cases, new data structures and algorithms have been introduced that perform optimally in this model. Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting (\cite{prokop1999coa}). Bender, Demaine and Farach-Colton designed a cache-oblivious variant of a B-tree (\cite{bender2005cob}, \cite{bender2004lpc}).

Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model (\cite{vitter2001ema}), which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.

[TODO: add more recent related work]

[TODO: find empirical results if they exists, or note lack of empirical data]

\section{Problem statement}
Several cache-oblivious data structures and algorithms have been proposed. In most cases, theoretical complexity analysis shows that the proposed solutions are asymptotically optimal, or equivalent to alternative non-oblivious solutions (up to a constant factor). However, in software engineering practice we are not only interesting in asymptotic computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have undesirable theoretical properties are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).

This raises the question wether cache-oblivious data structures are actually preferable to traditional data structures in practice. The theory suggests that cache-oblivious algorithms should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set. However, such empirical data is scarce when it comes to cache-oblivious data structures. The goal of the proposed research is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).

From the application programmer's point of view, datastructures and algorithms can be grouped together based on the functionality they provide. For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different. Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the research to data structures that provide similar functions. For the proposed research, we will focus on dictionary-like data structures. Such data structures support storage and retrieval of data values by associating them with a particular data value, called a key.

[TODO: discuss many variations in more detail? E.g. in-order traversal, range queries, etc.]

By disregarding the values in a dictionary, it can be simplified to a dynamic set. A set data structure supports (at the very least) insertion of elements and querying wether the set contains a particular element. This is an important class of data structures, because their functionality is often required to implement efficient combinatorial search algorithms. Since these search algorithms are often used to process large large data sets, the performance of underlying data structures is important.

\subsection{Research questions}
Following from the above, the main question to be answered is:
\begin{list}{}{}
\item Do cache-oblivious data structures have practical merit?
\end{list}

More specifically, these questions will be answered:
\newcounter{cnt}
\begin{list}{\arabic{cnt}.}{\usecounter{cnt}}
\item How do cache-oblivious dictionary data structures perform when compared to cache-aware data structures?
\item How do cache-oblivious dictionary data structures perform when compared to cache-unaware data structures?
\item How does the memory use of cache-oblivious data structures relate to the memory used by the other data structures?
\end{list}
Because we are primarily interested in practical performance, the measure of performance will be the actual run-time of the experiments. Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures.

\section{Research approach}

% To add:
% - language of implementation
% - platform used
% - data used
% - how to test

The goal of the research is to evaluate a variety of dictionary-like data structures in a practical setting. For this, we need a realistic scenario that makes heavy use of dictionaries, and that allows us to easily change the implementation used. The scenario that will be used, is that of a state-space search. The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state. State search algorithms are often used to solve reachability problems, which occur for example in formal model checking. The main data structures used are a queue (either a regular queue or, if search heuristics are used, a priority queue) and a dynamic set. An example of a simple state search algorithm is given in program \ref{prog-search}.

\begin{program}
\begin{verbatim}
Queue queue = new Queue;
Set visited = new Set;
queue.insert(initial_state);
while (!queue.empty()) {
    State state = queue.extract();
    for (State s : successors(state)) {
        if (isGoal(s)) break;
        if (visited.contains(s) == false) {
            visited.insert(s);
            queue.insert(s);
        }
    }
}
\end{verbatim}
\caption{Pseudo-code for a simple state search algorithm.}
\label{prog-search}
\end{program}

Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the runtime of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient). The main focus of the research will therefore be on alternative implementations of the set data structure.

The situation changes, of course, when instead of a first-in-first-out queue a priority queue is used. This is often the case when we are searching for a shortest path (instead of any path that reaches the goal) or when heuristics are used in the search. In this case, the complexity of the queue increases. If time allows it, the research could be extended to evaluating different implementations of priority queues as well.

\subsection{Set operations}
A minimal set data structure should be able to determine whether or not a given element is a member of the set. A dynamic set differs from a static set in that it support operations to add and/or remove elements from the set dynamically. In our experiments, we will start out with an empty set and only allow for insertion of new elements. This means the our dynamic set must support two basic operations:
\begin{itemize}
\item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
\item \verb#contains(S,X)# -- Returns wether set $S$ contains $X$.
\end{itemize}
In practice, it useful to be able to test if a set contains an element, and add this element if it does not. This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation. We will therefore require that the \verb#insert# operation also returns wether or not the element was newly inserted.

\subsection{Set implementations}
Many different data structures have been developed that can be used to efficiently implement the operations described above. Several data structures will be selected from the three categories listed in the introduction:
\begin{itemize}
\item \textbf{Cache-unaware}: e.g. a hash table.
\item \textbf{Cache-aware}: e.g. a traditional B-tree.
\item \textbf{Cache-obvlious}: e.g. a cache-oblivious B-tree.
\end{itemize}
[To be determined: which data structures to use?]

The cache-oblivious data structures have been developed recently and are not commonly used in practice. These will have to be implemented from scratch, using existing publications as a guideline. The traditional data structures are commonly used in practice and therefore there are high-quality implementations already available as software libraries. For this research, however, it is undesirable to use existing libraries, for two reasons. First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations. Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not. This may give the well-optimized data structures an unfair advantage.

In an attempt to reduce the bias caused by differences in functionality and quality between existing libraries, all data structures used in the performance evaluation will be implemented from scratch. Existing libraries will only be used to verify that the newly implemented data structures have appropriate performance.

\subsection{Measurement}
As described in the problem statement, the main metric of interest is the total runtime of our algorithm, since the focus of the research is to determine the practical merit of the data structures used.

[To be determined: which hardware and software platform used, et cetera.]

Apart from the framework in which the various data structures are evaluated, an important decision is the choice of the data set used to perform the experiments. At least three variables must be taken into account:
\begin{itemize}
\item The size of the data set: Cache-oblivious data structures usually rely on asymptotic optimality, which means that they should perform well on large data sets. It would be interesting to see what the effect of various data sizes is on their relative performance.
\item The ratio between queries and insertions: for a particular data structure, it may be much more efficient to query the existence of a value in the set than to add a new value to it. We should either ensure our ratio between queries and insertions is realistic, or run experiments with varying ratios to see how this affects the performance of different data structures.
\item Data access patterns: some data structures perform better when accessing elements that have recently been accessed before. Again, we should either ensure our access patterns are realistic, or test with varying access patterns and see how data structures are affected.
\end{itemize}

Finally, we must estimate the overhead caused by our testing framework and other outside influences, to be able to quantify the results obtained.

\section{Planning}
See table \ref{tab-planning}.

\begin{table*}
\begin{center}
\begin{tabular}{ | l | l | l | }
\hline
\textbf{Task} & \textbf{Result} & \textbf{Deadline} \\
\hline
Finish research proposal
    & Final reseach proposal
    & March 27th \\
Study literature
    & "Related work" section
    & April 4th \\
Implement algorithms
    & Working code
    & April 28th \\
Finish paper contents
    & Complete paper for peer review
    & May 6th \\
Revise paper
    & Revised paper
    & May 23rd \\
Final paper formatting
    & Finished paper
    & June 11th \\
Prepare presentation
    & Presentation
    & June 20th \\
\hline
\end{tabular}
\caption{Planning}
\label{tab-planning}
\end{center}
\end{table*}

\bibliographystyle{alpha}
\bibliography{proposal}
\end{document}

### 2008-03-09 proposal.tex
Spell checked draft proposal.

@@
 
-\keywords{cache efficiency,locality of reference,algorithms}
+\keywords{cache efficiency, locality of reference, algorithms}
 
@@
 The main reasons this simplification does not hold in practice is that processors can process data much faster than it can be read from (or written to) memory.
-To overcome the memory bottleneck, system designers have added faster cache memory at various points in the computer system.
-This has resulted in hierarchical ordering of various types of memory, each faster but smaller than the next, such as the processor's registers, L1 cache on the processor die, L2 cache, main memory, magnetic disk storage and sometimes network storage.
+To overcome the memory bottleneck, computer system designers have added faster cache memory at various points in the computer architecture.
+This has resulted in a hierarchical ordering of various types of memory, each faster but smaller than the next, such as the processor's registers, L1 cache on the processor die, L2 cache, main memory, magnetic disk storage and sometimes network storage.
 
@@
 
-[TODO: more detailed discussion of how fast memory acts as a cache for slow memory, data is transfered and blocks, and how this leads to algorithms that have good temporal and spatial locality to have superior performance.]
+[TODO: more detailed discussion of how fast memory acts as a cache for slow memory, data is transferred and blocks, and how this leads to algorithms that have good temporal and spatial locality to have superior performance.]
 
@@
 In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited).
-The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transfered between the two memories.
+The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories.
 The advantage of using such a model is that it more accurately reflects reality yet still allows formal analysis of the performance of various algorithms and data structures.
 Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy.
-In that case, the possibility of transfering several data blocks at once is usually dropped.
+In that case, the possibility of transferring several data blocks at once is usually dropped.
 
@@
 \item \textbf{Cache-unaware} [does a better term exist?] data structures and algorithms  are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
-\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotical sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
+\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotic sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
 \end{list}
@@
 
-This raises the question wether cache-oblivious data structures are actually preferable to traditional data structures in practice.
+This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice.
 The theory suggests that cache-oblivious algorithms should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set.
@@
 
-From the application programmer's point of view, datastructures and algorithms can be grouped together based on the functionality they provide.
+From the application programmer's point of view, data structures and algorithms can be grouped together based on the functionality they provide.
 For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different.
@@
 By disregarding the values in a dictionary, it can be simplified to a dynamic set.
-A set data structure supports (at the very least) insertion of elements and querying wether the set contains a particular element.
+A set data structure supports (at the very least) insertion of elements and querying whether the set contains a particular element.
 This is an important class of data structures, because their functionality is often required to implement efficient combinatorial search algorithms.
@@
 
-Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the runtime of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient).
+Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient).
 The main focus of the research will therefore be on alternative implementations of the set data structure.
@@
 \item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
-\item \verb#contains(S,X)# -- Returns wether set $S$ contains $X$.
+\item \verb#contains(S,X)# -- Returns whether set $S$ contains $X$.
 \end{itemize}
@@
 This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation.
-We will therefore require that the \verb#insert# operation also returns wether or not the element was newly inserted.
+We will therefore require that the \verb#insert# operation also returns whether or not the element was newly inserted.
 
@@
 \item \textbf{Cache-aware}: e.g. a traditional B-tree.
-\item \textbf{Cache-obvlious}: e.g. a cache-oblivious B-tree.
+\item \textbf{Cache-oblivious}: e.g. a cache-oblivious B-tree.
 \end{itemize}
@@
 \subsection{Measurement}
-As described in the problem statement, the main metric of interest is the total runtime of our algorithm, since the focus of the research is to determine the practical merit of the data structures used.
+As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the research is to determine the practical merit of the data structures used.
 
@@
 Finish research proposal
-    & Final reseach proposal
+    & Final research proposal
     & March 27th \\

### 2008-03-10 proposal.tex

@@
 
-\subsection{External memory model}
+\subsection{External Memory Model}
 In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited).
@@
 
-\subsection{Cache-oblivious memory model}
+\subsection{Cache-Oblivious Memory Model}
 The external memory model can be simplified and extended to a cache-oblivious model (\cite{prokop1999coa}) which recognizes the same distinction between internal and external memory, but does not assume the relevant parameters (size of the cache and size of the data blocks) are known.
@@
 \begin{list}{}{}
-\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache lay-out (such as block size or cache memory size).
+\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
 \item \textbf{Cache-unaware} [does a better term exist?] data structures and algorithms  are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
@@
 
-\section{Related work}
+\section{Related Work}
 Many data structures and algorithms have been analyzed in the cache-oblivious model.
@@
 
-\section{Problem statement}
-Several cache-oblivious data structures and algorithms have been proposed.
-In most cases, theoretical complexity analysis shows that the proposed solutions are asymptotically optimal, or equivalent to alternative non-oblivious solutions (up to a constant factor).
-However, in software engineering practice we are not only interesting in asymptotic computational complexity, but also in the practical performance of data structures and algorithms.
-Indeed, many algorithms that have undesirable theoretical properties are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).
+\section{Problem Statement}
+Several cache-oblivious data structures and algorithms have been described above.
+Theoretical complexity analysis shows that the proposed solutions are asymptotically optimal.
+However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms.
+Indeed, many algorithms that have undesirable theoretical complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).
 
@@
 For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different.
-Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the research to data structures that provide similar functions.
+Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions.
 For the proposed research, we will focus on dictionary-like data structures.
@@
 
-\subsection{Research questions}
+\subsection{Research Questions}
 Following from the above, the main question to be answered is:
@@
 
-\section{Research approach}
+\section{Research Approach}
 
@@
 
-The goal of the research is to evaluate a variety of dictionary-like data structures in a practical setting.
+The goal of the project is to evaluate a variety of dictionary-like data structures in a practical setting.
 For this, we need a realistic scenario that makes heavy use of dictionaries, and that allows us to easily change the implementation used.
@@
 Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient).
-The main focus of the research will therefore be on alternative implementations of the set data structure.
+The main focus of the research project will therefore be on alternative implementations of the set data structure.
 
@@
 In this case, the complexity of the queue increases.
-If time allows it, the research could be extended to evaluating different implementations of priority queues as well.
+If time allows it, the project could be extended to evaluating different implementations of priority queues as well.
 
-\subsection{Set operations}
+\subsection{Set Operations}
 A minimal set data structure should be able to determine whether or not a given element is a member of the set.
@@
 
-\subsection{Set implementations}
+\subsection{Set Implementations}
 Many different data structures have been developed that can be used to efficiently implement the operations described above.
@@
 \subsection{Measurement}
-As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the research is to determine the practical merit of the data structures used.
+As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the project is to determine the practical merit of the data structures used.
 
@@
 Study literature
-    & "Related work" section
+    & "Related Work" section
     & April 4th \\

### 2008-03-12 proposal/proposal.tex
Lots of updates to the proposal.

--- 2008-03-10 proposal.tex
@@
 
+% TODO: abstract
+%\begin{abstract}
+%The cache-oblivious memory model is a model that allows algorithms and data structures to be evaluated in terms of memory transfers in a two-level memory hierarchy.
+Efficient data structurs and algorithms that operate on them are of interest because they are used in more complex, domain-specific algorithms.
+This proposal suggests that emperical evidence is necessary to determine wether cache-oblivious data structures that have been described previously, perform well enough in practice so that they may replace traditional data structures.
+%\end{abstract}
+
 \keywords{cache efficiency, locality of reference, algorithms}
@@
 However, like every model, it is a simplification of reality.
-One of the simplifications of the model that has become more detached from reality, is the uniform memory model: the assumption that storing and retrieving data in memory is a constant-time operation.
+One of the simplifications is the assumptions of a uniform memory model: storing and retrieving data in memory is a constant-time operation.
+Advancements in software and hardware design over the last two decades have caused this assumption to be increasingly detached from reality.
 
-The main reasons this simplification does not hold in practice is that processors can process data much faster than it can be read from (or written to) memory.
-To overcome the memory bottleneck, computer system designers have added faster cache memory at various points in the computer architecture.
-This has resulted in a hierarchical ordering of various types of memory, each faster but smaller than the next, such as the processor's registers, L1 cache on the processor die, L2 cache, main memory, magnetic disk storage and sometimes network storage.
+One of the reasons this assumption does not hold in reality, is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (faster) main memory when the latter is in short supply.
+Another is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
+Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem.
 
-[TODO: add picture of memory hierarchy]
+It should be clear by now that a modern computer system lacks a single, central memory storage.
+Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage.
+At each of these levels, the next storage type is both significantly faster and significantly larger than the previous type.
+Typically, the processor can only directly manipulate data contained in its registers.
+To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy.
+Memory transfer takes place in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
 
-[TODO: more detailed discussion of how fast memory acts as a cache for slow memory, data is transferred and blocks, and how this leads to algorithms that have good temporal and spatial locality to have superior performance.]
+% TODO: add picture of memory hierarchy
 
+Apparently, the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed.
 Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account.
-To be able to analyze these algorithms and data structures formally, a different model of computation and especially of the cost of accessing data in memory is required.
+To be able to analyze these algorithms and data structures formally, a different, more realistic model of the cost of accessing data in memory is required.
 
-\subsection{External Memory Model}
+\subsection{Alternative Memory Models}
 In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited).
 The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories.
-The advantage of using such a model is that it more accurately reflects reality yet still allows formal analysis of the performance of various algorithms and data structures.
 Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy.
-In that case, the possibility of transferring several data blocks at once is usually dropped.
+In that case, the possibility of transferring several data blocks at once (bulk transfer) may have to be dropped.
+
+The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice there is a larger memory hierarchy consisting of multiple levels.
+Aggarwal, Alpern, Chandra and Snir (\cite{aggarwal1987mhm}) introduced a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
 
 \subsection{Cache-Oblivious Memory Model}
-The external memory model can be simplified and extended to a cache-oblivious model (\cite{prokop1999coa}) which recognizes the same distinction between internal and external memory, but does not assume the relevant parameters (size of the cache and size of the data blocks) are known.
-Therefore, cache-oblivious data structures and algorithms must be defined without explicitly referencing these parameters, yet still perform well when evaluated in the cache-oblivious model.
-Algorithms that perform well in the cache-oblivious model are more general (since they have no parameters that must be specified) and by design tuned to all levels in the memory hierarchy, which means they should perform well in practice without any tuning.
+The external memory model was generalized and simplified into the cache-oblivious model (\cite{prokop1999coa}) which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
+Data is transferred between the two in aligned data blocks of size B.
+In this model, the values of parameters like M and B are not known to the algorithms.
+Therefore, in the cache-oblivious model data structures and algorithms must be defined without explicitly referencing these parameters.
+Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
+Like the external memory model, the cache-oblivious model can be used to describe every pair of adjacent levels in the memory hierarchy.
+
+It should be noted that the cache-oblivious model is somewhat similar to the hierarchical model discussed above if the cost function of the hierarchical model is unknown.
+The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
+
+The key advantage of the cache-oblivious model is that algorithms that perform optimally when evaluated in this model are automatically tuned to all levels in the real memory hierarchy.
+Therefore, these algorithms might perform better than algorithms that are tuned to a specific level in the hierarchy.
 
@@
 \item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
-\item \textbf{Cache-unaware} [does a better term exist?] data structures and algorithms  are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
+\item \textbf{Cache-unaware} data structures and algorithms are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
 \item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotic sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
@@
 Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting (\cite{prokop1999coa}).
-Bender, Demaine and Farach-Colton designed a cache-oblivious variant of a B-tree (\cite{bender2005cob}, \cite{bender2004lpc}).
+
+Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree (\cite{bender2005cob}) achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
+Later, Bender, Duan, Iacono and Wu simplified the data structure (\cite{bender2004lpc}) while preserving the supported operations and complexity bounds and adding finger searches.
+The latter data structure consists of sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout used to find elements.
 
 Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model (\cite{vitter2001ema}), which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.
+Unfortunately, his survey lacks empirical results.
 
-[TODO: add more recent related work]
+% [TODO: add more recent related work]
 
-[TODO: find empirical results if they exists, or note lack of empirical data]
+% [TODO: find empirical results if they exists, or note lack of empirical data]
 
 \section{Problem Statement}
-Several cache-oblivious data structures and algorithms have been described above.
-Theoretical complexity analysis shows that the proposed solutions are asymptotically optimal.
+As described above, several cache-oblivious data structures and algorithms have been proposed.
+Complexity analysis shows that the proposed solutions are asymptotically optimal.
 However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms.
-Indeed, many algorithms that have undesirable theoretical complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).
+Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).
+The converse is true as well: some algorithms that are better than widely used alternatives in terms of computational complexity have bad performance on practical data sets, use too much memory, or are too hard to implement correctly (for example, linear time construction of suffix arrays).
 
@@
 Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions.
-For the proposed research, we will focus on dictionary-like data structures.
-Such data structures support storage and retrieval of data values by associating them with a particular data value, called a key.
-
-[TODO: discuss many variations in more detail? E.g. in-order traversal, range queries, etc.]
-
-By disregarding the values in a dictionary, it can be simplified to a dynamic set.
-A set data structure supports (at the very least) insertion of elements and querying whether the set contains a particular element.
-This is an important class of data structures, because their functionality is often required to implement efficient combinatorial search algorithms.
-Since these search algorithms are often used to process large large data sets, the performance of underlying data structures is important.
+As will described below, we will limit ourselves to a one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
 
@@
 \begin{list}{\arabic{cnt}.}{\usecounter{cnt}}
-\item How do cache-oblivious dictionary data structures perform when compared to cache-aware data structures?
-\item How do cache-oblivious dictionary data structures perform when compared to cache-unaware data structures?
-\item How does the memory use of cache-oblivious data structures relate to the memory used by the other data structures?
+\item How do cache-oblivious data structures perform when compared to cache-aware data structures?
+\item How do cache-oblivious data structures implementations perform when compared to cache-unaware data structures?
+\item How does the memory use of cache-oblivious data structures relate to the memory used by other data structures?
 \end{list}
@@
 
-The goal of the project is to evaluate a variety of dictionary-like data structures in a practical setting.
+The goal of the project is to evaluate a variety of data structures in a practical setting.
 For this, we need a realistic scenario that makes heavy use of dictionaries, and that allows us to easily change the implementation used.
-The scenario that will be used, is that of a state-space search.
+The scenario that will be used, is that of a state space search (or, in short: state search).
 The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state.
@@
 
-Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient).
+In this program, a queue is a data structure where elements can be inserted and extracted on a last-in-first-out basis.
+A dynamic set is a data structure that models a set to which elements may be added (and optionally, from which they may be removed) and which may be queried to see if the set contains a particular element.
+
+Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient).
 The main focus of the research project will therefore be on alternative implementations of the set data structure.
 
-The situation changes, of course, when instead of a first-in-first-out queue a priority queue is used.
+%The situation changes, of course, when instead of a first-in-first-out queue a priority queue is used.
 This is often the case when we are searching for a shortest path (instead of any path that reaches the goal) or when heuristics are used in the search.
-In this case, the complexity of the queue increases.
+In this case, the complexity of the queue implementation increases.
 If time allows it, the project could be extended to evaluating different implementations of priority queues as well.
@@
 A minimal set data structure should be able to determine whether or not a given element is a member of the set.
-A dynamic set differs from a static set in that it support operations to add and/or remove elements from the set dynamically.
+A dynamic set differs from a static set in that it supports operations to add and/or remove elements from the set dynamically.
 In our experiments, we will start out with an empty set and only allow for insertion of new elements.
@@
 \end{itemize}
-In practice, it useful to be able to test if a set contains an element, and add this element if it does not.
+In practice, it is useful to be able to test if a set contains an element and add this element if it does not.
 This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation.
-We will therefore require that the \verb#insert# operation also returns whether or not the element was newly inserted.
+To take advantage of this possibility, we will require that the \verb#insert# operation also returns whether or not the element was newly inserted.
 
@@
 Many different data structures have been developed that can be used to efficiently implement the operations described above.
-Several data structures will be selected from the three categories listed in the introduction:
+For each of the categories identified earlier, at least one data structure will be implemented:
 \begin{itemize}
-\item \textbf{Cache-unaware}: e.g. a hash table.
-\item \textbf{Cache-aware}: e.g. a traditional B-tree.
-\item \textbf{Cache-oblivious}: e.g. a cache-oblivious B-tree.
+\item \textbf{Cache-unaware}: hash tables.
+Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
+Other possibilities are Patricia trees (\cite{morrison1968ppa}), Splay trees (\cite{sleator1985sab}) or one of the many other variations on search trees.
+\item \textbf{Cache-aware}: B-trees.
+The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
+\item \textbf{Cache-oblivious}: the B-tree-like data structure proposed by Bender, Duan, Iacono and Wu.
+Since it provides functionality comparable to that of a normal B-tree, this seems like a fair candidate for a comparison.
+An alternative is a the cache-oblivious exponential search tree proposed by Rahman, Cole and Raman.
+\cite{rahman2001opd}.
+% The second, if I can find it! (If so, add description to related work)
 \end{itemize}
-[To be determined: which data structures to use?]
 
@@
 These will have to be implemented from scratch, using existing publications as a guideline.
-The traditional data structures are commonly used in practice and therefore there are high-quality implementations already available as software libraries.
+The traditional data structures are commonly used and therefore have high-quality implementations already available as software libraries.
 For this research, however, it is undesirable to use existing libraries, for two reasons.
@@
 In an attempt to reduce the bias caused by differences in functionality and quality between existing libraries, all data structures used in the performance evaluation will be implemented from scratch.
-Existing libraries will only be used to verify that the newly implemented data structures have appropriate performance.
 
 \subsection{Measurement}
+% TODO: which hardware and software platform used, et cetera. Do we discuss this?
 As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the project is to determine the practical merit of the data structures used.
+To be able to quantify the obtained results, we also need to be able to estimate the overhead incurred by the use of our testing framework and other outside influences.
+Two methods will be employed:
+\begin{itemize}
+\item Profiling tools such as OProfile (\cite{oprofile}) can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
+\item A mock set implementation will be written that does not implement set functionality itself, but can reads the answers to queries from a file.
+The execution time using this implementation is an estimation of the overhead caused by the testing framework, as the set implementation will do almost no work at all.
+\end{itemize}
 
-[To be determined: which hardware and software platform used, et cetera.]
+As a secondary metric, we are interested in the memory use of the data structures used. This is harder to quantify, but the following methods may be employed:
+\begin{itemize}
+\item The maximum resident set size of a process is a lower bound on the amount of memory a process used.
+\item The number of times a memory page has to be read from or written to disk is an indication of how often the application exceeds the available main memory.
+\end{itemize}
+Note that these metrics only given an indication of memory use on the boundary between main memory and disk storage.
 
-Apart from the framework in which the various data structures are evaluated, an important decision is the choice of the data set used to perform the experiments.
-At least three variables must be taken into account:
+\subsection{Test Data}
+The test data will be generated with state space generation software that generates synthetic test data based on parameters supplied by the user.
+This allows us to assess the effect of varying parameters on the performance of the data structures being evaluated.
+Since the same software is used in state search related research, the generated test cases are expected to be representative of realistic data sets.
+
+The results may depend on the properties of the generated test data. At least three variables can be identified:
 \begin{itemize}
@@
 We should either ensure our ratio between queries and insertions is realistic, or run experiments with varying ratios to see how this affects the performance of different data structures.
+In the case of state search algorithms, a sparse state space (few transitions between states) results in a high number of insertions.
 \item Data access patterns: some data structures perform better when accessing elements that have recently been accessed before.
+For example, splay trees explicitly support this principle.
 Again, we should either ensure our access patterns are realistic, or test with varying access patterns and see how data structures are affected.
 \end{itemize}
-
-Finally, we must estimate the overhead caused by our testing framework and other outside influences, to be able to quantify the results obtained.
 

### 2008-03-12 proposal/proposal.tex
More updates.

@@
 \section{Introduction}
-A fundamental part of theoretical computer science is the study of algorithms: formal descriptions of how computations may be performed.
+%[A fundamental part of theoretical computer science is the study of algorithms: formal descriptions of how computations may be performed.
 In order for algorithms to be able to process information efficiently, it must be stored in a systematic way, called a data structure.
-Algorithms and data structures therefore complement each other: while algorithms describe how a computer may process information, data structures describe how information is stored in a (model of a) computer's memory.
+Algorithms and data structures therefore complement each other: while algorithms describe how a computer may process information, data structures describe how information is stored in a (model of a) computer's memory.]
 
+A fundamental part of theoretical computer science is the study of algorithms (formal descriptions of how computations may be performed) and data structures (descriptions of how information is organized and stored in computers).
 Traditionally, algorithms have been evaluated in a simplified model of computation.
-In this model it is assumed that a computer executes an algorithm in discrete steps, and at each step it performs one elementary operation (e.g.
+In this model it is assumed that a computer executes an algorithm in discrete steps.
+At each step it performs one elementary operation (e.g.
 comparing two numbers, adding one to another, storing a value in memory, et cetera).
@@
 However, like every model, it is a simplification of reality.
-One of the simplifications is the assumptions of a uniform memory model: storing and retrieving data in memory is a constant-time operation.
+One of the simplifications is the assumption that storing and retrieving data in memory is a constant-time operation (i.e.
+the uniform memory model).
 Advancements in software and hardware design over the last two decades have caused this assumption to be increasingly detached from reality.
 
-One of the reasons this assumption does not hold in reality, is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (faster) main memory when the latter is in short supply.
+One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (faster) main memory when the latter is in short supply.
 Another is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
-Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem.
+Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem, but this increases the differences in access time between different types of memory.
 
@@
 At each of these levels, the next storage type is both significantly faster and significantly larger than the previous type.
-Typically, the processor can only directly manipulate data contained in its registers.
+Typically, the processor can directly manipulate data contained in its registers only.
 To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy.
-Memory transfer takes place in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
+Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
 
@@
 The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice there is a larger memory hierarchy consisting of multiple levels.
+Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
 Aggarwal, Alpern, Chandra and Snir (\cite{aggarwal1987mhm}) introduced a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
+This more closely resembles a real memory hierarchy.
 
@@
 Data is transferred between the two in aligned data blocks of size B.
-In this model, the values of parameters like M and B are not known to the algorithms.
-Therefore, in the cache-oblivious model data structures and algorithms must be defined without explicitly referencing these parameters.
+In this model (and unlike the external memory model) the values of parameters like M and B are not known to algorithms and data structures, so they cannot be referenced explicitly in their definition.
 Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
@@
 
-It should be noted that the cache-oblivious model is somewhat similar to the hierarchical model discussed above if the cost function of the hierarchical model is unknown.
+It should be noted that the cache-oblivious model can also be considered similar to the hierarchical model discussed above.
 The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
+% This makes it easier to design algorithms in the oblivious model, because in the HMM the algorithm has the responsibility to use the cache correctly.
 
-The key advantage of the cache-oblivious model is that algorithms that perform optimally when evaluated in this model are automatically tuned to all levels in the real memory hierarchy.
-Therefore, these algorithms might perform better than algorithms that are tuned to a specific level in the hierarchy.
+The key advantage of the cache-oblivious model is that algorithms that perform optimally when evaluated in this model are implicitly tuned to all levels in the real memory hierarchy at once.
+Therefore, these algorithms might perform better than algorithms that are tuned to a specific level in the hierarchy only.
 
@@
 Later, Bender, Duan, Iacono and Wu simplified the data structure (\cite{bender2004lpc}) while preserving the supported operations and complexity bounds and adding finger searches.
-The latter data structure consists of sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout used to find elements.
+The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout used to find elements.
 
@@
 Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).
-The converse is true as well: some algorithms that are better than widely used alternatives in terms of computational complexity have bad performance on practical data sets, use too much memory, or are too hard to implement correctly (for example, linear time construction of suffix arrays).
+%The converse is true as well: some algorithms that are better than widely used alternatives in terms of computational complexity have bad performance on practical data sets, use too much memory, or are too hard to implement correctly (for example, linear time construction of suffix arrays).
 
@@
 The theory suggests that cache-oblivious algorithms should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set.
-However, such empirical data is scarce when it comes to cache-oblivious data structures.
-The goal of the proposed research is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).
+However, such empirical data is scarce when it comes to cache-oblivious data structures, as existing research has focused mainly on theoretical analysis.
+The goal of the proposed research project is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).
 
@@
 Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions.
-As will described below, we will limit ourselves to a one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
+As will described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
 
@@
 \item How do cache-oblivious data structures perform when compared to cache-aware data structures?
-\item How do cache-oblivious data structures implementations perform when compared to cache-unaware data structures?
+\item How do cache-oblivious data structures perform when compared to cache-unaware data structures?
 \item How does the memory use of cache-oblivious data structures relate to the memory used by other data structures?
@@
 The goal of the project is to evaluate a variety of data structures in a practical setting.
-For this, we need a realistic scenario that makes heavy use of dictionaries, and that allows us to easily change the implementation used.
-The scenario that will be used, is that of a state space search (or, in short: state search).
+For this, we need a realistic scenario that makes heavy use of data structures of which cache-oblivious implementations are available.
+A scenario that suits these needs, is that of a state space search algorithm (or, in short: state search).
 The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state.
@@
 
-In this program, a queue is a data structure where elements can be inserted and extracted on a last-in-first-out basis.
+In this program, a queue is a data structure where elements can be inserted and extracted on a first-in-first-out basis.
 A dynamic set is a data structure that models a set to which elements may be added (and optionally, from which they may be removed) and which may be queried to see if the set contains a particular element.
 
-Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient).
+Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient.
 The main focus of the research project will therefore be on alternative implementations of the set data structure.
@@
 Since it provides functionality comparable to that of a normal B-tree, this seems like a fair candidate for a comparison.
-An alternative is a the cache-oblivious exponential search tree proposed by Rahman, Cole and Raman.
+An alternative is the cache-oblivious exponential search tree proposed by Rahman, Cole and Raman.
 \cite{rahman2001opd}.
@@
 Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not.
-This may give the well-optimized data structures an unfair advantage.
+This may give the existing libraries an unfair advantage.
 
-In an attempt to reduce the bias caused by differences in functionality and quality between existing libraries, all data structures used in the performance evaluation will be implemented from scratch.
+In an attempt to reduce the bias caused by differences in functionality and quality between existing and newly developed libraries, all data structures used in the performance evaluation will be implemented from scratch.
 
@@
 % TODO: which hardware and software platform used, et cetera. Do we discuss this?
-As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the project is to determine the practical merit of the data structures used.
+As described in the problem statement, the main metric of interest is the total execution time of our algorithm.
 To be able to quantify the obtained results, we also need to be able to estimate the overhead incurred by the use of our testing framework and other outside influences.
@@
 \item Profiling tools such as OProfile (\cite{oprofile}) can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
-\item A mock set implementation will be written that does not implement set functionality itself, but can reads the answers to queries from a file.
-The execution time using this implementation is an estimation of the overhead caused by the testing framework, as the set implementation will do almost no work at all.
+\item A mock set implementation will be written that does not implement set functionality itself, but reads the answers to queries from a file.
+This can be used to give an upper bound on the overhead caused by the testing framework.
 \end{itemize}
 
-As a secondary metric, we are interested in the memory use of the data structures used. This is harder to quantify, but the following methods may be employed:
+As a secondary metric, we are interested in the memory use of the different data structures. This is harder to quantify, but the following methods may be employed:
 \begin{itemize}
@@
 This allows us to assess the effect of varying parameters on the performance of the data structures being evaluated.
-Since the same software is used in state search related research, the generated test cases are expected to be representative of realistic data sets.
+Since the same software is already used in state search related research, the generated test cases are expected to be representative of realistic data sets.
 
@@
 \begin{itemize}
-\item The size of the data set: Cache-oblivious data structures usually rely on asymptotic optimality, which means that they should perform well on large data sets.
+\item The size of the data set: Cache-oblivious data structures usually rely on asymptotic optimality, which means that they may not perform well on small data sets.
 It would be interesting to see what the effect of various data sizes is on their relative performance.

### 2008-03-13 proposal/proposal.tex
Moved text "Research Proposal" to subtitle.
Fixed bibliography style.

@@
 
-\title{Evaluation of Cache-Oblivious Data Structures\\Draft research proposal}
+\title{Evaluation of Cache-Oblivious Data Structures}
+\subtitle{Research Proposal (draft)}
 
@@
 \subsection{Alternative Memory Models}
-In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited).
+In the external memory model, as described by Aggarwal and Vitter \cite{aggarwal1988ioc}, a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited).
 The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories.
@@
 Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
-Aggarwal, Alpern, Chandra and Snir (\cite{aggarwal1987mhm}) introduced a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
+Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} introduced a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
 This more closely resembles a real memory hierarchy.
@@
 \subsection{Cache-Oblivious Memory Model}
-The external memory model was generalized and simplified into the cache-oblivious model (\cite{prokop1999coa}) which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
+The external memory model was generalized and simplified into the cache-oblivious model \cite{prokop1999coa} which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
 Data is transferred between the two in aligned data blocks of size B.
@@
 Erik D.
-Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms (\cite{demaine2002coa}).
+Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms \cite{demaine2002coa}.
 He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.
@@
 In many cases, new data structures and algorithms have been introduced that perform optimally in this model.
-Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting (\cite{prokop1999coa}).
+Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.
 
-Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree (\cite{bender2005cob}) achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
-Later, Bender, Duan, Iacono and Wu simplified the data structure (\cite{bender2004lpc}) while preserving the supported operations and complexity bounds and adding finger searches.
+Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
+Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding finger searches.
 The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout used to find elements.
 
-Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model (\cite{vitter2001ema}), which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.
+Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model \cite{vitter2001ema}, which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.
 Unfortunately, his survey lacks empirical results.
@@
 Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
-Other possibilities are Patricia trees (\cite{morrison1968ppa}), Splay trees (\cite{sleator1985sab}) or one of the many other variations on search trees.
+Other possibilities are Patricia trees \cite{morrison1968ppa}, Splay trees \cite{sleator1985sab} or one of the many other variations on search trees.
 \item \textbf{Cache-aware}: B-trees.
@@
 \begin{itemize}
-\item Profiling tools such as OProfile (\cite{oprofile}) can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
+\item Profiling tools such as OProfile \cite{oprofile} can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
 \item A mock set implementation will be written that does not implement set functionality itself, but reads the answers to queries from a file.
@@
 
-\bibliographystyle{alpha}
+\bibliographystyle{plain}
 \bibliography{proposal}

### 2008-03-13 btree/Btree_Set.c
Code clean-up: added some curly braces.

@@ -56,6 +56,8 @@ struct PageEntry
     char   data[1];         /* Data */
 };
 
+/* Prints the contents of the given page in a human-readable format.
+   Useful for debugging. */
 static void debug_print_page(Btree_Set *set, char *page, FILE *fp)
 {
     int n, N;
@@ -72,7 +74,9 @@ static void debug_print_page(Btree_Set *set, char *page, FILE *fp)
         if (n > 0)
         {
             if (n%16 == 0)
+            {
                 putc('\n', fp);
+            }
             else
             {
                 putc(' ', fp);
@@ -289,10 +293,14 @@ static struct PageEntry *find_or_insert_page( Btree_Set *set, int pageno,
         d = cmp(page + BEGIN(page, mid), SIZE(page, mid), key_data, key_size);
 
         if (d < 0)
+        {
             n = mid + 1;
+        }
         else
         if (d > 0)
+        {
             m = mid;
+        }
         else
         {
             /* Entry found! */

### 2008-03-13 btree/test.c

@@ -27,6 +27,7 @@ static void parse_args(int argc, char *argv[])
     int type = -1, pagesize;
 
     while ((opt = getopt(argc, argv, "bhc:k")) != -1)
+    {
         switch (opt)
         {
         case 'b':
@@ -53,6 +54,7 @@ static void parse_args(int argc, char *argv[])
             fprintf(stderr, "Unknown option: %c\n", opt);
             exit(1);
         }
+    }
 
     if (type == -1)
     {

### 2008-03-13 btree/Btree_Set.c
Eliminated dynamic memory allocation in B-tree operations (memory is still allocated by the constructor).

@@ -47,14 +47,19 @@ typedef struct Btree_Set
     size_t  pagesize;       /* Size of data pages */
     int     pages;          /* Number of pages */
     int     root;           /* Index of root page */
+
+    /* Temporary memory pool */
+    size_t  mem_size;       /* Total amount of memory allocated */
+    size_t  mem_used;       /* Memory used (set to zero before each operation) */
+    char    *mem;           /* Allocated memory pool */
 } Btree_Set;
 
-struct PageEntry
+typedef struct PageEntry
 {
     size_t size;            /* Size of data */
     int    child;           /* Index of page following entry */
     char   data[1];         /* Data */
-};
+} PageEntry;
 
 /* Prints the contents of the given page in a human-readable format.
    Useful for debugging. */
@@ -91,16 +96,27 @@ static void debug_print_page(Btree_Set *set, char *page, FILE *fp)
     fprintf(fp, "----\n");
 }
 
-/* Allocates a blank page.
-   The memory is zeroed to make debugging easier. */
+/*  Allocates the requested number of a bytes from the reserved space. */
+static void *alloc_mem(Btree_Set *set, size_t size)
+{
+    void *data;
+
+    /* Align to "long" while assuming sizeof(long) is a power of two! */
+    size = (size + (sizeof(long) - 1))&~(sizeof(long) - 1);
+
+    /* Allocate from the pool. */
+    assert(set->mem_size - set->mem_used >= size);
+    data = set->mem + set->mem_used;
+    set->mem_used += size;
+
+    return data;
+}
+
+/* Allocates a new page from the reserved space.
+   This page must not be freed by the caller. */
 static void *alloc_page(Btree_Set *set)
 {
-    char *page;
-
-    page = calloc(1, set->pagesize);
-    assert(page != NULL);
-
-    return page;
+    return alloc_mem(set, set->pagesize);
 }
 
 /* Reads a page with the given index.
@@ -154,12 +170,12 @@ static void set_destroy(Btree_Set *set)
 
 /* Creates an entry to be inserted in a page, consisting of a value
    (a size/data pair) and a successor page index. */
-static struct PageEntry *make_entry(size_t size, int page, const void *data)
+static PageEntry *make_entry(
+    Btree_Set *set, size_t size, int page, const void *data)
 {
-    struct PageEntry *entry;
+    PageEntry *entry;
 
[…]

### 2008-03-13 btree/Btree_Set.c
Added support for user-supplied comparison and hashing functions (not supported for BerkeleyDB)

@@ -43,6 +43,7 @@
 typedef struct Btree_Set
 {
     Set     base;
+
     int     fd;             /* Descriptor of file backing the set */
     size_t  pagesize;       /* Size of data pages */
     int     pages;          /* Number of pages */
@@ -183,23 +184,6 @@ static PageEntry *make_entry(
     return entry;
 }
 
-/* Lexicographical comparison. */
-static int cmp(const void *d1, size_t s1, const void *d2, size_t s2)
-{
-    int dif;
-
-    dif = memcmp(d1, d2, s1 < s2 ? s1 : s2);
-    if (dif == 0)
-    {
-        if (s1 < s2)
-            dif = -1;
-        else
-        if (s1 > s2)
-            dif = +1;
-    }
-    return dif;
-}
-
 /* Inserts an entry into a page.
    The 'entry' should be dynamically allocated, and is freed by this function.
    If the page has to be split, a new entry is returned, that is to be inserted
@@ -303,7 +287,8 @@ static PageEntry *find_or_insert_page( Btree_Set *set, int pageno,
         int d, mid;
 
         mid = (n + m)/2;
-        d = cmp(page + BEGIN(page, mid), SIZE(page, mid), key_data, key_size);
+        d = set->base.compare( set->base.context,
+            page + BEGIN(page, mid), SIZE(page, mid), key_data, key_size );
 
         if (d < 0)
         {
@@ -430,9 +415,13 @@ Set *Btree_Set_create(const char *filepath, size_t pagesize)
         return NULL;
     }
 
+    set->base.context  = NULL;
     set->base.destroy  = (void*)set_destroy;
     set->base.insert   = (void*)set_insert;
     set->base.contains = (void*)set_contains;
+    set->base.compare  = default_compare;
+    set->base.hash     = NULL;
+
     set->fd       = fd;
     set->pagesize = pagesize;
     set->pages    = 1;

### 2008-03-13 btree/Set.h

@@ -6,6 +6,9 @@
 
 /* Defines the basic operations on a set data structure.
 
+The "context" field may be set to a specific value by the application.
+It is not directly used by the set implementations.
+
 void destroy(Set *set)
     Frees the data structure and associated resources.
 
@@ -16,11 +19,27 @@ bool insert(Set *set, const void *key_data, size_t key_size)
 bool contains(Set *set, const void *key_data, size_t key_size)
     Returns true if the key element is present in the set; false otherwise.
 
+int compare(const void *context, const void *data1, size_t size1,
+                                 const void *data2, size_t size2)
+    Performs a three-way comparison between two keys, returning an integer
+    less than, equal to, or greater than zero if the first key is respectively
+    less than, equal to, or greater than the second.
+
+unsigned hash(const void *context, const void *key_data, size_t key_size)
+    Computes a hash value for the given key.
+
 */
 typedef struct Set {
+    void *context;
+
+    /* These functions are set by the constructor */
     void (*destroy)(struct Set *);
     bool (*insert)(struct Set *, const void *, size_t);
     bool (*contains)(struct Set *, const void *, size_t);
+
+    /* These functions may be overridden by the caller */
+    int (*compare)(const void *, const void *, size_t, const void *, size_t);
+    unsigned (*hash)(const void *, const void *, size_t);
 } Set;
 
 /* Creates a set data structure backed by a BerkeleyDB B-tree. */
@@ -31,5 +50,12 @@ Set *BDB_Hash_Set_create(const char *filepath);
 
 /* Creates a set data structure backed by a custom B-tree implementation. */
 Set *Btree_Set_create(const char *filepath, size_t pagesize);
+
+/* Default comparison function */
+int default_compare(const void *ignored,
+    const void *d1, size_t s1, const void *d2, size_t s2);
+
+/* Default hash function */
+unsigned default_hash(const void *ignored, const void *data, size_t size);
 
 #endif /* ndef SET_H_INCLUDED */

### 2008-03-17 proposal/proposal.tex
Updated research proposal.

@@
 
-% TODO: abstract
+% TODO: abstract?
 %\begin{abstract}
-%The cache-oblivious memory model is a model that allows algorithms and data structures to be evaluated in terms of memory transfers in a two-level memory hierarchy.
-Efficient data structurs and algorithms that operate on them are of interest because they are used in more complex, domain-specific algorithms.
-This proposal suggests that emperical evidence is necessary to determine wether cache-oblivious data structures that have been described previously, perform well enough in practice so that they may replace traditional data structures.
 %\end{abstract}
@@
 Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage.
-At each of these levels, the next storage type is both significantly faster and significantly larger than the previous type.
+At each of these levels, the next storage type is both significantly slower and significantly larger than the previous type.
 Typically, the processor can directly manipulate data contained in its registers only.
@@
 This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice.
-The theory suggests that cache-oblivious algorithms should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set.
-However, such empirical data is scarce when it comes to cache-oblivious data structures, as existing research has focused mainly on theoretical analysis.
+These data structures are designed to perform optimally in a model that was designed to more accurately reflect reality, which suggests they should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set.
+However, such empirical data is scarce, as existing research has focused mainly on theoretical analysis.
 The goal of the proposed research project is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).
@@
 Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions.
-As will described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
+As will be described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
 
@@
 
-More specifically, these questions will be answered:
+Because we are primarily interested in practical performance, actual execution time is the most important metric.
+Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures.
+That means that the main question will be answered by first addressing the following questions:
 \newcounter{cnt}
 \begin{list}{\arabic{cnt}.}{\usecounter{cnt}}
-\item How do cache-oblivious data structures perform when compared to cache-aware data structures?
-\item How do cache-oblivious data structures perform when compared to cache-unaware data structures?
-\item How does the memory use of cache-oblivious data structures relate to the memory used by other data structures?
+\item How fast are cache-oblivious data structures when compared to cache-aware data structures?
+\item How fast are cache-oblivious data structures when compared to cache-unaware data structures?
+\item How much memory do cache-oblivious data structures use compared to other data structures?
 \end{list}
-Because we are primarily interested in practical performance, the measure of performance will be the actual run-time of the experiments.
-Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures.
+These questions can be answered by implementing the relevant data structures and measuring their practical performance on a realistic data set.
 
 \section{Research Approach}
-
-% To add:
-% - language of implementation
-% - platform used
-% - data used
-% - how to test
-
 The goal of the project is to evaluate a variety of data structures in a practical setting.
@@
 State search algorithms are often used to solve reachability problems, which occur for example in formal model checking.
-The main data structures used are a queue (either a regular queue or, if search heuristics are used, a priority queue) and a dynamic set.
-An example of a simple state search algorithm is given in program \ref{prog-search}.
+In a simple implementation, such as given in program \ref{prog-search}, the main data structures used are a queue and a dynamic set.
 
@@
 
-Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient.
+Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient enough not to dominate the performance of the algorithm.
 The main focus of the research project will therefore be on alternative implementations of the set data structure.
@@
 \subsection{Set Operations}
-A minimal set data structure should be able to determine whether or not a given element is a member of the set.
+At a minimum, a set data structure should be able to determine whether or not a given element is a member of the set.
 A dynamic set differs from a static set in that it supports operations to add and/or remove elements from the set dynamically.
@@
 \subsection{Measurement}
-% TODO: which hardware and software platform used, et cetera. Do we discuss this?
 As described in the problem statement, the main metric of interest is the total execution time of our algorithm.

### 2008-03-17 sets/gen-test.py

@@ -14,7 +14,7 @@ while len(ls) < N:
     ls.append(s)
 
 # Output a bunch of random queries
-for _ in range(2*N):
+for _ in range(4*N):
     s = ls[rnd.randint(0,len(ls)-1)]
     if rnd.randint(0,1):
         print s

### 2008-03-24 proposal/proposal.tex
Some more changes to the proposal.

@@
 
-% TODO: abstract?
-%\begin{abstract}
-%\end{abstract}
+\begin{abstract}
+In modern computer hardware architecture, memory is organized in a hierarchy consisting of several types of memory with different sizes and access times.
+The cache-oblivious model has been proposed reflect this reality more accurately than traditional models.
+A number of algorithms and data structures have been proposed that perfom optimally in this model.
+The goal of the proposed research project is to implement a selection of cache-oblivious set-like data structures, evaluate their performance (in terms of execution time and memory use) in a realistic environment, and compare their performance to traditional data structures.
+\end{abstract}
 
@@
 \section{Introduction}
-%[A fundamental part of theoretical computer science is the study of algorithms: formal descriptions of how computations may be performed.
-In order for algorithms to be able to process information efficiently, it must be stored in a systematic way, called a data structure.
-Algorithms and data structures therefore complement each other: while algorithms describe how a computer may process information, data structures describe how information is stored in a (model of a) computer's memory.]
-
 A fundamental part of theoretical computer science is the study of algorithms (formal descriptions of how computations may be performed) and data structures (descriptions of how information is organized and stored in computers).
@@
 
-It should be clear by now that a modern computer system lacks a single, central memory storage.
+It should be clear by now that a modern computer system lacks a central memory storage with uniform performance characteristics.
 Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage.
@@
 
+We will call algorithms that are designed to minimize the number of transfers in a two-level memory hierarchy ``cache-aware'' (as opposed to traditional ``cache-unaware'' algorithms) or ``cache-concious''.
+These algorithms typically rely on knowledge of the block size to achieve optimal performance.
+
 \subsection{Cache-Oblivious Memory Model}
@@
 Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
-Like the external memory model, the cache-oblivious model can be used to describe every pair of adjacent levels in the memory hierarchy.
 
@@
 The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
-% This makes it easier to design algorithms in the oblivious model, because in the HMM the algorithm has the responsibility to use the cache correctly.
+The cache-oblivious model is more realistic in this respect.
 
-The key advantage of the cache-oblivious model is that algorithms that perform optimally when evaluated in this model are implicitly tuned to all levels in the real memory hierarchy at once.
-Therefore, these algorithms might perform better than algorithms that are tuned to a specific level in the hierarchy only.
+Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration.
+The key advantage of this class of algorithms is that they are implicitly tuned to all levels in the memory hierarchy at once.
+It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.
 
@@
 
+Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list.
+This means buckets have to resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).
+
 Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model \cite{vitter2001ema}, which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.
@@
 
-% [TODO: add more recent related work]
-
-% [TODO: find empirical results if they exists, or note lack of empirical data]
+Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-oblivious priority deque.
+Their main result is that although the cache-oblivious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
 

### 2008-03-26 proposal/proposal.tex
Updated proposal with some more figures and final changes.

@@
 \usepackage{float}
+\usepackage{url}
 
@@
 \title{Evaluation of Cache-Oblivious Data Structures}
-\subtitle{Research Proposal (draft)}
+\subtitle{Research Proposal}
 
@@
 
-\textit{\small 8$^{th}$ Twente Student Conference on IT, Enschede, June, 2008}
+\textit{\small 9$^{th}$ Twente Student Conference on IT, Enschede, June, 2008}
 
@@
 \begin{abstract}
-In modern computer hardware architecture, memory is organized in a hierarchy consisting of several types of memory with different sizes and access times.
-The cache-oblivious model has been proposed reflect this reality more accurately than traditional models.
-A number of algorithms and data structures have been proposed that perfom optimally in this model.
+In modern computer hardware architecture, memory is organized in a hierarchy consisting of several types of memory with different memory sizes, block transfer sizes and access times.
+The cache-oblivious model has been proposed to reflect this reality more accurately than traditional models.
+A number of algorithms and data structures have been proposed that perform optimally in this model.
 The goal of the proposed research project is to implement a selection of cache-oblivious set-like data structures, evaluate their performance (in terms of execution time and memory use) in a realistic environment, and compare their performance to traditional data structures.
@@
 the uniform memory model).
-Advancements in software and hardware design over the last two decades have caused this assumption to be increasingly detached from reality.
+Advancements in software and hardware design over the last two de\-cades have caused this assumption to be increasingly detached from reality.
 
-One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (faster) main memory when the latter is in short supply.
+One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (fast\-er) main memory when the latter is in short supply.
 Another is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
@@
 
-It should be clear by now that a modern computer system lacks a central memory storage with uniform performance characteristics.
-Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage.
-At each of these levels, the next storage type is both significantly slower and significantly larger than the previous type.
-Typically, the processor can directly manipulate data contained in its registers only.
+As a result, a modern computer system lacks a central memory storage with uniform performance characteristics.
+Instead, it employs a hierarchy of memory storage types.
+Figure \ref{fig-memhier} gives a typical example of such a hierarchy.
+The processor can directly manipulate data contained in its registers only.
 To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy.
-Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
+Memory is typically transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
 
-% TODO: add picture of memory hierarchy
+\begin{figure}
+\centering
+\includegraphics{memhier}
+\caption{Schematic depiction of the memory hierarchy}\label{fig-memhier}
+\end{figure}
 
-Apparently, the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed.
+In the memory hierarchy, every next level of storage is both significantly slower and significantly larger than the one above it and with increasing memory sizes, the block size increases as well.
+Table \ref{tab-memhier} gives an overview of typical memory sizes, block sizes, and access times.
+The values for the (general purpose) registers, L1 cache and L2 cache are those for the Pentium M processor as given in \cite{intel-opt}; other values are approximate.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{ l l l l }
+\hline
+\textbf{Type} & \textbf{Total Size} & \textbf{Block Size} & \textbf{Access Time} \\
+\hline
+Registers  &   32 bytes & 4 bytes & 1 cycle \\
+L1 Cache   &   32 KB    & 64 bytes & 3 cyles \\
+L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
+RAM        &   ~2 GB    & 1 KB     & 50-100 cycles \\
+Disk       & ~300 GB    & 4 KB     & 5,000-10,000 cycles \\
+\hline
+\end{tabular}
+\caption{Sizes and access times of various types of storage}
+\label{tab-memhier}
+\end{center}
+\end{table}
+
+To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed.
 Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account.
@@
 
-The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice there is a larger memory hierarchy consisting of multiple levels.
+The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice the memory hierarchy is deeper.
 Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
@@
 
-We will call algorithms that are designed to minimize the number of transfers in a two-level memory hierarchy ``cache-aware'' (as opposed to traditional ``cache-unaware'' algorithms) or ``cache-concious''.
+We will call algorithms that are designed to minimize the number of transfers in a two-level memory hierarchy ``cache-aware'' (as opposed to traditional ``cache-unaware'' algorithms) or ``cache-conscious''.
 These algorithms typically rely on knowledge of the block size to achieve optimal performance.
@@
 The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
-The cache-oblivious model is more realistic in this respect.
+The cache-oblivious model is more realistic in this respect, as applications typically have not explicit control over the cache.
 
@@
 
-Erik D.
 Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms \cite{demaine2002coa}.
@@
 
-\subsection{Categories}
-The development of the different memory models means data structures and algorithms can be classified depending on the assumptions about their environment used in their definitions.
+\subsection{Classification}
+To summarize the above: data structures and algorithms can be classified by which assumptions they make about the environment.
+For the purpose of this research project we distinguish between three classes:
 \begin{list}{}{}
-\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
+\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
 \item \textbf{Cache-unaware} data structures and algorithms are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
@@
 Many data structures and algorithms have been analyzed in the cache-oblivious model.
-In many cases, new data structures and algorithms have been introduced that perform optimally in this model.
+Several new data structures and algorithms have been developed that perform optimally in this mo\-del as well.
 Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.
@@
 Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding finger searches.
-The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout used to find elements.
+The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout (named after a similar structure proposed by van Emde Boas et al.
+\cite{vanemdeboas1976dai}).
 
 Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list.
-This means buckets have to resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).
+This means buckets have to be resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).
 
@@
 
-Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-oblivious priority deque.
-Their main result is that although the cache-oblivious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
+Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-obli\-vious priority deque.
+Their main result is that although the cache-obli\-vious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
 
@@
 A scenario that suits these needs, is that of a state space search algorithm (or, in short: state search).
-The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state.
+The purpose of such an algorithm is to find a path from a (set of) given initial state(s) to a goal state.
 State search algorithms are often used to solve reachability problems, which occur for example in formal model checking.
-In a simple implementation, such as given in program \ref{prog-search}, the main data structures used are a queue and a dynamic set.
+In a simple implementation, such as given in Program \ref{prog-search}, the main data structures used are a queue and a dynamic set.
 
@@
 
-Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient enough not to dominate the performance of the algorithm.
+Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#isGoal# and \verb#successors# in the sample code) are efficient enough not to dominate the performance of the algorithm.
 The main focus of the research project will therefore be on alternative implementations of the set data structure.
@@
 In practice, it is useful to be able to test if a set contains an element and add this element if it does not.
-This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation.
+This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in Program \ref{prog-search}) but many data structures allow to combine this into a single operation.
 To take advantage of this possibility, we will require that the \verb#insert# operation also returns whether or not the element was newly inserted.
@@
 \item Profiling tools such as OProfile \cite{oprofile} can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
-\item A mock set implementation will be written that does not implement set functionality itself, but reads the answers to queries from a file.
+\item A mock set implementation will be written that does not implement set functionality itself, but gives precomputed answers to queries.
 This can be used to give an upper bound on the overhead caused by the testing framework.
@@
 We should either ensure our ratio between queries and insertions is realistic, or run experiments with varying ratios to see how this affects the performance of different data structures.
-In the case of state search algorithms, a sparse state space (few transitions between states) results in a high number of insertions.
 \item Data access patterns: some data structures perform better when accessing elements that have recently been accessed before.
@@
 \section{Planning}
-See table \ref{tab-planning}.
-
-\begin{table*}
+\begin{table}
 \begin{center}
@@
 \hline
-Finish research proposal
+Finish proposal
     & Final research proposal
@@
 Finish paper contents
-    & Complete paper for peer review
+    & Complete paper
     & May 6th \\
@@
 \end{center}
-\end{table*}
+\end{table}
+See table \ref{tab-planning}.
 

### 2008-03-26 proposal/proposal.tex
Some final changes.

@@
 One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (fast\-er) main memory when the latter is in short supply.
-Another is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
+Another reason is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
 Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem, but this increases the differences in access time between different types of memory.
@@
 The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
-The cache-oblivious model is more realistic in this respect, as applications typically have not explicit control over the cache.
+The cache-oblivious model is more realistic in this regard, as applications typically have no explicit control over the cache.
 
@@
 Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
-Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding finger searches.
-The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout (named after a similar structure proposed by van Emde Boas et al.
+Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for finger searches.
+The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout (named for similarity to a different data structure proposed by van Emde Boas et al.
 \cite{vanemdeboas1976dai}).
@@
 Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions.
-As will be described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
+As will be described below, we will limit ourselves to one particular kind of data structure, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
 
@@
 Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures.
-That means that the main question will be answered by first addressing the following questions:
+The main question will be answered by first addressing the following questions:
 \newcounter{cnt}
@@
 Many different data structures have been developed that can be used to efficiently implement the operations described above.
-For each of the categories identified earlier, at least one data structure will be implemented:
+For each of the three classes identified earlier, at least one data structure will be implemented:
 \begin{itemize}

### 2008-04-14 sets/BDB_Set.c
Added support for user-supplied data comparison and hash functions. The default comparison is a lexicographical comparison and the default hash function is FNV-1a.

@@ -59,6 +59,8 @@ static Set *BDB_Set_create(const char *filepath, int type)
     set->base.destroy  = (void*)set_destroy;
     set->base.insert   = (void*)set_insert;
     set->base.contains = (void*)set_contains;
+    set->base.compare  = type == DB_BTREE ? default_compare : NULL;
+    set->base.hash     = type == DB_HASH  ? default_hash    : NULL;
     set->db = db;
 
     return &set->base;

### 2008-04-14 sets/Btree_Set.c

@@ -10,10 +10,6 @@
 #include <sys/stat.h>
 #include <sys/mman.h>
 
-/* Defined in comparison.c */
-int default_compare(const void *d1, size_t s1, const void *d2, size_t s2);
-
-
 /* Page lay-out:
 
     |---- 'pagesize' bytes ---|
@@ -275,8 +271,9 @@ static PageEntry *find_or_insert_page( Btree_Set *set, int page,
         int d, mid;
 
         mid = (n + m)/2;
-        d = default_compare( DATA(page) + BEGIN(page, mid), SIZE(page, mid),
-                             key_data, key_size );
+        d = set->base.compare( set->base.context,
+                               DATA(page) + BEGIN(page, mid), SIZE(page, mid),
+                               key_data, key_size );
 
         if (d < 0)
         {
@@ -400,6 +397,8 @@ Set *Btree_Set_create(const char *filepath, size_t pagesize)
     set->base.destroy  = (void*)set_destroy;
     set->base.insert   = (void*)set_insert;
     set->base.contains = (void*)set_contains;
+    set->base.compare  = default_compare;
+
     set->fd       = fd;
     set->pagesize = pagesize;
     set->pages    = 0;

### 2008-04-14 sets/Hash_Set.c

@@ -10,12 +10,6 @@
 #include <sys/stat.h>
 #include <sys/mman.h>
 
-/* Defined in hashing.c */
-unsigned default_hash(const void *ignored, const void *data, size_t size);
-
-/* Defined in comparison.c */
-int default_compare(const void *d1, size_t s1, const void *d2, size_t s2);
-
 
 /* Table lay-out (assuming sizeof(size_t) == 8):
 
@@ -146,7 +140,7 @@ static bool find_or_insert( Hash_Set *set, const void *key_data, size_t key_size
     size_t *next;
 
     /* Find initial entry */
-    hash = default_hash(NULL, key_data, key_size);
+    hash = set->base.hash(set->base.context, key_data, key_size);
     next = (size_t*)set->data + hash%set->capacity;
     while (*next != 0)
     {
@@ -158,8 +152,12 @@ static bool find_or_insert( Hash_Set *set, const void *key_data, size_t key_size
 
         size = *(size_t*)(set->data + *next + sizeof(size_t));
         data = set->data + *next + 2*sizeof(size_t);
-        if (default_compare(key_data, key_size, data, size) == 0)
+        if (set->base.compare( set->base.context,
+                               key_data, key_size, data, size ) == 0)
+        {
+            /* item found */
             break;
+        }
 
         next = (size_t*)(set->data + *next);
     }
@@ -239,9 +237,13 @@ Set *Hash_Set_create(const char *filepath, size_t capacity)
     if (set == NULL)
         return NULL;
 
+    set->base.context  = NULL;
     set->base.destroy  = (void*)set_destroy;
     set->base.insert   = (void*)set_insert;
     set->base.contains = (void*)set_contains;
+    set->base.compare  = default_compare;
+    set->base.hash     = default_hash;
+
     set->fd        = fd;
     set->capacity  = capacity;
     set->data      = NULL;

### 2008-04-14 sets/Set.h

@@ -6,6 +6,9 @@
 
 /* Defines the basic operations on a set data structure.
 
+The "context" field may be set to a specific value by the application.
+It is not directly used by the set implementations.
+
 void destroy(Set *set)
     Frees the data structure and associated resources.
 
@@ -16,11 +19,27 @@ bool insert(Set *set, const void *key_data, size_t key_size)
 bool contains(Set *set, const void *key_data, size_t key_size)
     Returns true if the key element is present in the set; false otherwise.
 
+int compare(const void *context, const void *data1, size_t size1,
+                                 const void *data2, size_t size2)
+    Performs a three-way comparison between two keys, returning an integer
+    less than, equal to, or greater than zero if the first key is respectively
+    less than, equal to, or greater than the second.
+
+unsigned hash(const void *context, const void *key_data, size_t key_size)
+    Computes a hash value for the given key.
+
 */
 typedef struct Set {
+    void *context;
+
+    /* These functions are set by the constructor */
     void (*destroy)(struct Set *);
     bool (*insert)(struct Set *, const void *, size_t);
     bool (*contains)(struct Set *, const void *, size_t);
+
+    /* These functions may be overridden by the caller */
+    int (*compare)(const void *, const void *, size_t, const void *, size_t);
+    unsigned (*hash)(const void *, const void *, size_t);
 } Set;
 
 /* Creates a set data structure backed by a BerkeleyDB B-tree. */
@@ -32,9 +51,13 @@ Set *BDB_Hash_Set_create(const char *filepath);
 /* Creates a set data structure backed by a custom B-tree implementation. */
 Set *Btree_Set_create(const char *filepath, size_t pagesize);
 
-/* Creates a set data structure backed by a custom hash table implementation.
-   The capacity parameter is the capacity of the hash table index (i.e. the
-   number of buckets) which is fixed for the set (no resizing possible). */
-Set *Hash_Set_create(const char *filepath, size_t capacity);
+/* Default comparison function */
+int default_compare( const void *ignored,
+                     const void *d1, size_t s1,
+                     const void *d2, size_t s2 );
+
+/* Default hash function */
+unsigned default_hash( const void *ignored,
+                       const void *data, size_t size );
 
 #endif /* ndef SET_H_INCLUDED */

### 2008-04-14 sets/comparison.c

@@ -2,7 +2,9 @@
 #include <string.h>
 
 /* Lexicographical comparison. */
-int default_compare(const void *d1, size_t s1, const void *d2, size_t s2)
+int default_compare( const void *ignored,
+                     const void *d1, size_t s1,
+                     const void *d2, size_t s2 )
 {
     int dif;
 

### 2008-04-14 paper/paper.tex
Added directory for paper (to be written).

--- 2008-03-24 proposal/proposal.tex
@@
 \title{Evaluation of Cache-Oblivious Data Structures}
-\subtitle{Research Proposal (draft)}
 
@@
 \begin{abstract}
-In modern computer hardware architecture, memory is organized in a hierarchy consisting of several types of memory with different sizes and access times.
-The cache-oblivious model has been proposed reflect this reality more accurately than traditional models.
-A number of algorithms and data structures have been proposed that perfom optimally in this model.
-The goal of the proposed research project is to implement a selection of cache-oblivious set-like data structures, evaluate their performance (in terms of execution time and memory use) in a realistic environment, and compare their performance to traditional data structures.
+%TODO
 \end{abstract}
@@
 \section{Introduction}
-A fundamental part of theoretical computer science is the study of algorithms (formal descriptions of how computations may be performed) and data structures (descriptions of how information is organized and stored in computers).
-Traditionally, algorithms have been evaluated in a simplified model of computation.
-In this model it is assumed that a computer executes an algorithm in discrete steps.
-At each step it performs one elementary operation (e.g.
-comparing two numbers, adding one to another, storing a value in memory, et cetera).
-Each elementary operation is performed within a constant time.
-In this model, both storing and retrieving data values in memory is considered to be an elementary operation.
 
-This model is close enough to the way computers work to be extremely useful in the development and analysis of data structures and algorithms that work well in practice.
-However, like every model, it is a simplification of reality.
-One of the simplifications is the assumption that storing and retrieving data in memory is a constant-time operation (i.e.
-the uniform memory model).
-Advancements in software and hardware design over the last two decades have caused this assumption to be increasingly detached from reality.
-
-One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (faster) main memory when the latter is in short supply.
-Another is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
-Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem, but this increases the differences in access time between different types of memory.
-
-It should be clear by now that a modern computer system lacks a central memory storage with uniform performance characteristics.
-Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage.
-At each of these levels, the next storage type is both significantly slower and significantly larger than the previous type.
-Typically, the processor can directly manipulate data contained in its registers only.
-To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy.
-Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
-
-% TODO: add picture of memory hierarchy
-
-Apparently, the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed.
-Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account.
-To be able to analyze these algorithms and data structures formally, a different, more realistic model of the cost of accessing data in memory is required.
-
-\subsection{Alternative Memory Models}
-In the external memory model, as described by Aggarwal and Vitter \cite{aggarwal1988ioc}, a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited).
-The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories.
-Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy.
-In that case, the possibility of transferring several data blocks at once (bulk transfer) may have to be dropped.
-
-The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice there is a larger memory hierarchy consisting of multiple levels.
-Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
-Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} introduced a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
-This more closely resembles a real memory hierarchy.
-
-We will call algorithms that are designed to minimize the number of transfers in a two-level memory hierarchy ``cache-aware'' (as opposed to traditional ``cache-unaware'' algorithms) or ``cache-concious''.
-These algorithms typically rely on knowledge of the block size to achieve optimal performance.
-
-\subsection{Cache-Oblivious Memory Model}
-The external memory model was generalized and simplified into the cache-oblivious model \cite{prokop1999coa} which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
-Data is transferred between the two in aligned data blocks of size B.
-In this model (and unlike the external memory model) the values of parameters like M and B are not known to algorithms and data structures, so they cannot be referenced explicitly in their definition.
-Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
-
-It should be noted that the cache-oblivious model can also be considered similar to the hierarchical model discussed above.
-The main difference between the two is that in the hierarchical model the application has explicit control over which data are placed at the fastest locations, since these have the lowest addresses, while in the cache-oblivious model the internal memory acts as a transparent cache of the external memory, and the application has no explicit control over which values are kept in the cache.
-The cache-oblivious model is more realistic in this respect.
-
-Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration.
-The key advantage of this class of algorithms is that they are implicitly tuned to all levels in the memory hierarchy at once.
-It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.
-
-Erik D.
-Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms \cite{demaine2002coa}.
-He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.
-
-\subsection{Categories}
-The development of the different memory models means data structures and algorithms can be classified depending on the assumptions about their environment used in their definitions.
-\begin{list}{}{}
-\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
-\item \textbf{Cache-unaware} data structures and algorithms are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
-\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotic sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
-\end{list}
-
-\section{Related Work}
-Many data structures and algorithms have been analyzed in the cache-oblivious model.
-In many cases, new data structures and algorithms have been introduced that perform optimally in this model.
-Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.
-
-Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
-Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding finger searches.
-The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout used to find elements.
-
-Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list.
-This means buckets have to resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).
-
-Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model \cite{vitter2001ema}, which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.
-Unfortunately, his survey lacks empirical results.
-
-Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-oblivious priority deque.
-Their main result is that although the cache-oblivious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
-
-\section{Problem Statement}
-As described above, several cache-oblivious data structures and algorithms have been proposed.
-Complexity analysis shows that the proposed solutions are asymptotically optimal.
-However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms.
-Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).
-%The converse is true as well: some algorithms that are better than widely used alternatives in terms of computational complexity have bad performance on practical data sets, use too much memory, or are too hard to implement correctly (for example, linear time construction of suffix arrays).
-
-This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice.
-These data structures are designed to perform optimally in a model that was designed to more accurately reflect reality, which suggests they should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set.
-However, such empirical data is scarce, as existing research has focused mainly on theoretical analysis.
-The goal of the proposed research project is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).
-
-From the application programmer's point of view, data structures and algorithms can be grouped together based on the functionality they provide.
-For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different.
-Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions.
-As will be described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.
-
-\subsection{Research Questions}
-Following from the above, the main question to be answered is:
-\begin{list}{}{}
-\item Do cache-oblivious data structures have practical merit?
-\end{list}
-
-Because we are primarily interested in practical performance, actual execution time is the most important metric.
-Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures.
-That means that the main question will be answered by first addressing the following questions:
-\newcounter{cnt}
-\begin{list}{\arabic{cnt}.}{\usecounter{cnt}}
-\item How fast are cache-oblivious data structures when compared to cache-aware data structures?
-\item How fast are cache-oblivious data structures when compared to cache-unaware data structures?
-\item How much memory do cache-oblivious data structures use compared to other data structures?
-\end{list}
-These questions can be answered by implementing the relevant data structures and measuring their practical performance on a realistic data set.
-
-\section{Research Approach}
-The goal of the project is to evaluate a variety of data structures in a practical setting.
-For this, we need a realistic scenario that makes heavy use of data structures of which cache-oblivious implementations are available.
-A scenario that suits these needs, is that of a state space search algorithm (or, in short: state search).
-The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state.
-State search algorithms are often used to solve reachability problems, which occur for example in formal model checking.
-In a simple implementation, such as given in program \ref{prog-search}, the main data structures used are a queue and a dynamic set.
-
-\begin{program}
-\begin{verbatim}
-Queue queue = new Queue;
-Set visited = new Set;
-queue.insert(initial_state);
-while (!queue.empty()) {
-    State state = queue.extract();
-    for (State s : successors(state)) {
-        if (isGoal(s)) break;
-        if (visited.contains(s) == false) {
-            visited.insert(s);
-            queue.insert(s);
-        }
-    }
-}
-\end{verbatim}
-\caption{Pseudo-code for a simple state search algorithm.}
-\label{prog-search}
-\end{program}
-
-In this program, a queue is a data structure where elements can be inserted and extracted on a first-in-first-out basis.
-A dynamic set is a data structure that models a set to which elements may be added (and optionally, from which they may be removed) and which may be queried to see if the set contains a particular element.
-
-Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient enough not to dominate the performance of the algorithm.
-The main focus of the research project will therefore be on alternative implementations of the set data structure.
-
-%The situation changes, of course, when instead of a first-in-first-out queue a priority queue is used.
-This is often the case when we are searching for a shortest path (instead of any path that reaches the goal) or when heuristics are used in the search.
-In this case, the complexity of the queue implementation increases.
-If time allows it, the project could be extended to evaluating different implementations of priority queues as well.
-
-\subsection{Set Operations}
-At a minimum, a set data structure should be able to determine whether or not a given element is a member of the set.
-A dynamic set differs from a static set in that it supports operations to add and/or remove elements from the set dynamically.
-In our experiments, we will start out with an empty set and only allow for insertion of new elements.
-This means the our dynamic set must support two basic operations:
-\begin{itemize}
-\item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
-\item \verb#contains(S,X)# -- Returns whether set $S$ contains $X$.
-\end{itemize}
-In practice, it is useful to be able to test if a set contains an element and add this element if it does not.
-This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation.
-To take advantage of this possibility, we will require that the \verb#insert# operation also returns whether or not the element was newly inserted.
-
-\subsection{Set Implementations}
-Many different data structures have been developed that can be used to efficiently implement the operations described above.
-For each of the categories identified earlier, at least one data structure will be implemented:
-\begin{itemize}
-\item \textbf{Cache-unaware}: hash tables.
-Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
-Other possibilities are Patricia trees \cite{morrison1968ppa}, Splay trees \cite{sleator1985sab} or one of the many other variations on search trees.
-\item \textbf{Cache-aware}: B-trees.
-The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
-\item \textbf{Cache-oblivious}: the B-tree-like data structure proposed by Bender, Duan, Iacono and Wu.
-Since it provides functionality comparable to that of a normal B-tree, this seems like a fair candidate for a comparison.
-An alternative is the cache-oblivious exponential search tree proposed by Rahman, Cole and Raman.
-\cite{rahman2001opd}.
-% The second, if I can find it! (If so, add description to related work)
-\end{itemize}
-
-The cache-oblivious data structures have been developed recently and are not commonly used in practice.
-These will have to be implemented from scratch, using existing publications as a guideline.
-The traditional data structures are commonly used and therefore have high-quality implementations already available as software libraries.
-For this research, however, it is undesirable to use existing libraries, for two reasons.
-First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations.
-Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not.
-This may give the existing libraries an unfair advantage.
-
-In an attempt to reduce the bias caused by differences in functionality and quality between existing and newly developed libraries, all data structures used in the performance evaluation will be implemented from scratch.
-
-\subsection{Measurement}
-As described in the problem statement, the main metric of interest is the total execution time of our algorithm.
-To be able to quantify the obtained results, we also need to be able to estimate the overhead incurred by the use of our testing framework and other outside influences.
-Two methods will be employed:
-\begin{itemize}
-\item Profiling tools such as OProfile \cite{oprofile} can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
-\item A mock set implementation will be written that does not implement set functionality itself, but reads the answers to queries from a file.
-This can be used to give an upper bound on the overhead caused by the testing framework.
-\end{itemize}
-
-As a secondary metric, we are interested in the memory use of the different data structures. This is harder to quantify, but the following methods may be employed:
-\begin{itemize}
-\item The maximum resident set size of a process is a lower bound on the amount of memory a process used.
-\item The number of times a memory page has to be read from or written to disk is an indication of how often the application exceeds the available main memory.
-\end{itemize}
-Note that these metrics only given an indication of memory use on the boundary between main memory and disk storage.
-
-\subsection{Test Data}
-The test data will be generated with state space generation software that generates synthetic test data based on parameters supplied by the user.
-This allows us to assess the effect of varying parameters on the performance of the data structures being evaluated.
-Since the same software is already used in state search related research, the generated test cases are expected to be representative of realistic data sets.
-
-The results may depend on the properties of the generated test data. At least three variables can be identified:
-\begin{itemize}
-\item The size of the data set: Cache-oblivious data structures usually rely on asymptotic optimality, which means that they may not perform well on small data sets.
-It would be interesting to see what the effect of various data sizes is on their relative performance.
-\item The ratio between queries and insertions: for a particular data structure, it may be much more efficient to query the existence of a value in the set than to add a new value to it.
-We should either ensure our ratio between queries and insertions is realistic, or run experiments with varying ratios to see how this affects the performance of different data structures.
-In the case of state search algorithms, a sparse state space (few transitions between states) results in a high number of insertions.
-\item Data access patterns: some data structures perform better when accessing elements that have recently been accessed before.
-For example, splay trees explicitly support this principle.
-Again, we should either ensure our access patterns are realistic, or test with varying access patterns and see how data structures are affected.
-\end{itemize}
-
-\section{Planning}
-See table \ref{tab-planning}.
-
-\begin{table*}
-\begin{center}
-\begin{tabular}{ | l | l | l | }
-\hline
-\textbf{Task} & \textbf{Result} & \textbf{Deadline} \\
-\hline
-Finish research proposal
-    & Final research proposal
-    & March 27th \\
-Study literature
-    & "Related Work" section
-    & April 4th \\
-Implement algorithms
-    & Working code
-    & April 28th \\
-Finish paper contents
-    & Complete paper for peer review
-    & May 6th \\
-Revise paper
-    & Revised paper
-    & May 23rd \\
-Final paper formatting
-    & Finished paper
-    & June 11th \\
-Prepare presentation
-    & Presentation
-    & June 20th \\
-\hline
-\end{tabular}
-\caption{Planning}
-\label{tab-planning}
-\end{center}
-\end{table*}
+%TODO: set up paper structure
+%TODO: borrow as much as possible from the proposal
 
 \bibliographystyle{plain}
-\bibliography{proposal}
+\bibliography{paper}
 \end{document}

### 2008-04-14 sets/Set.c
Moved option parsing code to a separate function so it is easier to reuse.

@@ -1,0 +1,169 @@
+#include "Set.h"
+#include <assert.h>
+#include <string.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <unistd.h>
+
+#define SET_BTREE        1
+#define SET_HASH         2
+#define SET_BDB          3
+#define SET_BDB_HASH     4
+#define SET_BDB_BTREE    5
+
+static const char *make_tempfile()
+{
+    int res;
+    static char path[] = "db-XXXXXX";
+
+    res = mkstemp(path);
+    assert(res >= 0);
+    close(res);
+
+    return path;
+}
+
+/*  Creates a set data structure from a string description.
+    Arguments between square brackets are optional.
+
+    If invalid arguments have been passed or a set with the requested
+    parameters cannot be constructed, NULL is returned. (This is not very
+    user-friendly.)
+
+    "btree [pagesize=P] .."
+    Creates a B-tree based set with a pagesize of P bytes (default: 4096).
+
+    "hash [capacity=C] .."
+    Creates a hash table based with capacity C items (default: 1,000,000).
+
+    "BerkeleyDB btree .."
+    Creates a BerkeleyDB B-tree based set.
+
+    "BerkeleyDB hash .."
+    Creates a BerkeleyDB hash table based set.
+
+    Common arguments:
+    ".. [file=path]"
+    Specifies a path to use to create the backing file. Note that if this
+    file exists, it will be overwritten! If no file is specified, a file is
+    created in the working directory.
+
+    ".. [keep]"
+    Keep the temporary file (i.e. do not unlink it after creation).
+*/
+Set *Set_create_from_args(int argc, const char * const *argv)
+{
+    int type, pagesize, capacity, keep;
+    char *path;
+    Set *result;
+
+    if (argc < 1)
+        return NULL;
+
+    path = NULL;
+    keep = 0;
+
+    if (strcmp(*argv, "btree") == 0)
+    {
+        pagesize = 4096;
+        type = SET_BTREE;
+    }
+    else
+    if (strcmp(*argv, "hash") == 0)
+    {
+        capacity = 1000000;
+        type = SET_HASH;
+    }
+    else
+    if (strcmp(*argv, "BerkeleyDB") == 0)
+    {
+        type = SET_BDB;
+    }
+    else
+    {
+        /* Invalid type! */
+        return NULL;
+    }
+
+    --argc, ++argv;
+
[…]

### 2008-04-14 sets/Set.h

@@ -4,6 +4,13 @@
 #include <stdbool.h>
 #include <stdlib.h>
 
+typedef struct Set Set;
+
+/* Creates a set data structure from a string description */
+/* Set *Set_create_from_string(const char *descr); */
+Set *Set_create_from_args(int argc, const char * const *argv);
+
+
 /* Defines the basic operations on a set data structure.
 
 The "context" field may be set to a specific value by the application.
@@ -29,7 +36,7 @@ unsigned hash(const void *context, const void *key_data, size_t key_size)
     Computes a hash value for the given key.
 
 */
-typedef struct Set {
+struct Set {
     void *context;
 
     /* These functions are set by the constructor */
@@ -40,7 +47,7 @@ typedef struct Set {
     /* These functions may be overridden by the caller */
     int (*compare)(const void *, const void *, size_t, const void *, size_t);
     unsigned (*hash)(const void *, const void *, size_t);
-} Set;
+};
 
 /* Creates a set data structure backed by a BerkeleyDB B-tree. */
 Set *BDB_Btree_Set_create(const char *filepath);
@@ -50,6 +57,10 @@ Set *BDB_Hash_Set_create(const char *filepath);
 
 /* Creates a set data structure backed by a custom B-tree implementation. */
 Set *Btree_Set_create(const char *filepath, size_t pagesize);
+
+/* Creates a set data structure backed by a custom hash table implementation. */
+Set *Hash_Set_create(const char *filepath, size_t capacity);
+
 
 /* Default comparison function */
 int default_compare( const void *ignored,

### 2008-04-14 sets/test.c

@@ -5,104 +5,18 @@
 #include <string.h>
 #include <unistd.h>
 
-static Set *set = NULL;
-
-static const char *make_tempfile()
-{
-    int res;
-    static char path[] = "db-XXXXXX";
-
-    res = mkstemp(path);
-    assert(res >= 0);
-    close(res);
-
-    return path;
-}
-
-static void parse_args(int argc, char *argv[])
-{
-    int opt;
-    const char *path = NULL;
-    bool keepdb = false;
-    int type = -1, parameter;
-
-    while ((opt = getopt(argc, argv, "bhc:d:k")) != -1)
-    {
-        switch (opt)
-        {
-        case 'b':
-            assert(type == -1);
-            type = 0;
-            break;
-
-        case 'h':
-            assert(type == -1);
-            type = 1;
-            break;
-
-        case 'c':
-            assert(type == -1);
-            type = 2;
-            parameter = atoi(optarg);
-            break;
-
-        case 'd':
-            assert(type == -1);
-            type = 3;
-            parameter = atoi(optarg);
-            break;
-
-        case 'k':
-            keepdb = true;
-            break;
-
-        default:
-            fprintf(stderr, "Unknown option: %c\n", opt);
-            exit(1);
-        }
-    }
-
-    if (type == -1)
-    {
-        printf("Usage:\n"
-               "\ttest -b [opts]            (BerkeleyDB B-tree backed)\n"
-               "\ttest -h [opts]            (BerkeleyDB hash table backed)\n"
-               "\ttest -c pagesize [opts]   (Custom B-tree backed)\n"
-               "\ttest -d capacity [opts]   (Custom hash table backed)\n"
-               "Options:\n"
-               "\t-k      Keep database file.\n");
-        exit(0);
-    }
-
-    /* Create set */
-    path = make_tempfile();
-    switch (type)
-    {
-    case 0:
-        set = BDB_Btree_Set_create(path);
-        break;
-
-    case 1:
-        set = BDB_Hash_Set_create(path);
-        break;
-
-    case 2:
-        set = Btree_Set_create(path, parameter);
-        break;
-
-    case 3:
-        set = Hash_Set_create(path, parameter);
-        break;
-    }
-    if (!keepdb)
-        unlink(path);
-}
-
 int main(int argc, char *argv[])
 {
+    Set *set;
     char line[4096];
 
-    parse_args(argc, argv);
[…]

### 2008-04-14 sets/Set.c
Changed a few typedefs.

@@ -5,11 +5,9 @@
 #include <stdlib.h>
 #include <unistd.h>
 
-#define SET_BTREE        1
-#define SET_HASH         2
-#define SET_BDB          3
-#define SET_BDB_HASH     4
-#define SET_BDB_BTREE    5
+typedef enum SetType {
+    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree
+} SetType;
 
 static const char *make_tempfile()
 {
@@ -53,7 +51,8 @@ static const char *make_tempfile()
 */
 Set *Set_create_from_args(int argc, const char * const *argv)
 {
-    int type, pagesize, capacity, keep;
+    SetType type;
+    int pagesize, capacity, keep;
     char *path;
     Set *result;
 
@@ -66,18 +65,18 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     if (strcmp(*argv, "btree") == 0)
     {
         pagesize = 4096;
-        type = SET_BTREE;
+        type = Btree;
     }
     else
     if (strcmp(*argv, "hash") == 0)
     {
         capacity = 1000000;
-        type = SET_HASH;
+        type = Hash;
     }
     else
     if (strcmp(*argv, "BerkeleyDB") == 0)
     {
-        type = SET_BDB;
+        type = BDB_Unspecified;
     }
     else
     {
@@ -92,27 +91,27 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     {
         if (strcmp(*argv, "btree") == 0)
         {
-            if (type != SET_BDB)
+            if (type != BDB_Unspecified)
                 return NULL;
-            type = SET_BDB_BTREE;
+            type = BDB_Btree;
         }
         else
         if (strcmp(*argv, "hash") == 0)
         {
-            if (type != SET_BDB)
+            if (type != BDB_Unspecified)
                 return NULL;
-            type = SET_BDB_HASH;
+            type = BDB_Hash;
         }
         else
         if (sscanf(*argv, "pagesize=%d", &pagesize) == 1)
         {
-            if (type != SET_BTREE)
+            if (type != Btree)
                 return NULL;
         }
         else
         if (sscanf(*argv, "capacity=%d", &capacity) == 1)
         {
-            if (type != SET_HASH)
+            if (type != Hash)
                 return NULL;
         }
         else
@@ -144,16 +143,16 @@ Set *Set_create_from_args(int argc, const char * const *argv)
 
     switch (type)
     {
[…]

### 2008-04-14 sets/Set.h

@@ -40,9 +40,9 @@ struct Set {
     void *context;
 
     /* These functions are set by the constructor */
-    void (*destroy)(struct Set *);
-    bool (*insert)(struct Set *, const void *, size_t);
-    bool (*contains)(struct Set *, const void *, size_t);
+    void (*destroy)(Set *);
+    bool (*insert)(Set *, const void *, size_t);
+    bool (*contains)(Set *, const void *, size_t);
 
     /* These functions may be overridden by the caller */
     int (*compare)(const void *, const void *, size_t, const void *, size_t);

### 2008-04-14 search/Deque.c
Added partial NON-FUNCTIONAL implementation of state search algorithm.

@@ -1,0 +1,116 @@
+#include "Deque.h"
+#include <assert.h>
+
+/* Deque implementation
+
+   Current implementation stores queue contents as a sequence of elements,
+   with the size stored at both ends (in order to be able to seek both ways).
+
+   ~------+--------+--------+--------+--~ ~-+--------+--------+------~
+   unused | size_1 |  data  | size_1 |      | data_N | size_N | unused
+   ~------+--------+--------+--------+--~ ~-+--------+--------+------~
+          |-- 8 ---|-size_1-|-- 8 ---|      |-- 8 ---|      |
+        begin                                                end
+
+   (Assuming sizeof(size_t) == 8)
+   Data stored is padded to multiples of sizeof(size_t).
+*/
+
+typedef struct DequeImpl DequeImpl;
+
+struct DequeImpl
+{
+    Deque   base;
+    size_t  count;          /* Number of elements */
+    char    *data;          /* Contents of backing file */
+    size_t  data_size;      /* Size of data used in the backing file */
+    size_t  data_left;      /* Unallocated space left in backing file */
+    size_t  data_begin;     /* Begin of valid data */
+    size_t  data_end;       /* End of valid data */
+};
+
+static void deque_destroy(DequeImpl *deque)
+{
+    free(deque);
+}
+
+static size_t size(DequeImpl *deque)
+{
+    return deque->count;
+}
+
+static bool empty(DequeImpl *deque)
+{
+    return deque->empty();
+}
+
+static void push_back(DequeImpl *deque, const void *data, size_t size)
+{
+    /* TODO */
+}
+
+static void push_front(DequeImpl *deque, const void *data, size_t size)
+{
+    /* NOT IMPLEMENTED */
+    assert(0);
+}
+
+static void get_back(DequeImpl *deque, const void **data, size_t *size)
+{
+    /* TODO */
+}
+
+static void get_front(DequeImpl *deque, const void **data, size_t *size)
+{
+    /* TODO */
+}
+
+static void pop_back(DequeImpl *deque)
+{
+    /* TODO */
+}
+
+static void pop_front(DequeImpl *deque)
+{
+    /* TODO */
+}
+
+Deque *Deque_create(const char *filepath)
+{
+    DequeImpl *deque;
+    int fd;
+
+    /* Open file */
+    fd = open(filepath, O_CREAT | O_RDWR, 0666);
+    if (fd < 0)
+        return NULL;
+
[…]

### 2008-04-14 search/Deque.h

@@ -1,0 +1,58 @@
+#ifndef DEQUE_H_INCLUDED
+#define DEQUE_H_INCLUDED
+
+#include <stdlib.h>
+
+typedef struct Deque Deque;
+
+/* Creates a new deque backed by the given file. */
+Deque *Deque_create(const char *filepath);
+
+/* The Deque structure implements a file-backed deque data structure (which is
+   a queue that supports adding and removing elements both at the front and at
+   the back).
+
+   Contains the following methods:
+
+    void destroy()
+        Destroys the queue and frees all allocated resources.
+
+    bool empty()
+        Returns wether the queue is empty.
+
+    size_t size()
+        Returns the number of elements in the queue.
+
+    void push_back(const void *data, size_t length)
+        Adds an element at the back of the queue.
+
+    void push_front(const void *data, size_t length)
+        Adds an element at the front of the queue.
+
+    void get_back(void **data, size_t *length)
+        Retrieves the element at the back of the queue.
+
+    void get_front(void **data, size_t *length)
+        Retrieves the element at the front of the queue.
+
+    void pop_back(const void *data, size_t length)
+        Removes an element at the back of the queue.
+
+    void pop_front(const void *data, size_t length)
+        Removes an element at the front of the queue.
+
+*/
+struct Deque
+{
+    void (*destroy)(struct Deque *);
+    bool (*empty)(struct Deque *);
+    size_t (*size)(struct Deque *, const void *, size_t);
+    void (*push_back)(struct Deque *, const void *, size_t);
+    void (*push_front)(struct Deque *, const void *, size_t);
+    bool (*get_back)(struct Deque *, void **, size_t *);
+    bool (*get_front)(struct Deque *, void **, size_t *);
+    void (*pop_back)(struct Deque *, const void *, size_t);
+    void (*pop_front)(struct Deque *, const void *, size_t);
+};
+
+#endif /* ndef DEQUE_H_INCLUDED */

### 2008-05-02 datastructures/Btree_Set.c
Removed some superfluous header includes.

@@ -2,14 +2,9 @@
 #include "FileStorage.h"
 #include "Set.h"
 #include <assert.h>
-#include <fcntl.h>
 #include <stdio.h>
 #include <stdbool.h>
 #include <string.h>
-#include <unistd.h>
-#include <sys/types.h>
-#include <sys/stat.h>
-#include <sys/mman.h>
 
 /* Page lay-out:
 
@@ -52,7 +47,7 @@ typedef struct Btree_Set
     int     pages;          /* Number of pages */
     int     root;           /* Index of root page */
 
-    FileStorage fs;         /* File storage (first member is data pointer) */
+    FileStorage fs;         /* File storage */
 
     /* Temporary memory pool */
     char    *mem;           /* Allocated memory pool */

### 2008-05-02 datastructures/Hash_Set.c

@@ -2,13 +2,9 @@
 #include "Set.h"
 #include "FileStorage.h"
 #include <assert.h>
-#include <fcntl.h>
 #include <stdio.h>
 #include <stdbool.h>
 #include <string.h>
-#include <unistd.h>
-#include <sys/types.h>
-#include <sys/stat.h>
 #include <sys/mman.h>
 
 
@@ -33,7 +29,7 @@ struct Hash_Set
 {
     Set     base;
     size_t  capacity;       /* Index capacity (number of buckets) */
-    FileStorage fs;         /* File storage (first member is data pointer) */
+    FileStorage fs;         /* File storage */
 };
 
 

### 2008-05-02 datastructures/Bender_Impl.c
Started implementation of cache-oblivious data structure.

@@ -1,0 +1,40 @@
+#include "config.h"
+#include "Bender_Impl.h"
+#include "FileStorage.h"
+#include <assert.h>
+
+/* Implementation of Bender's cache-oblivious set data structure. */
+
+void Bender_Impl_create( Bender_Impl *bi,
+                         const char *filepath, size_t value_size )
+{
+    bool ok;
+
+    /* Check for valid size (positive integer multiple of sizeof(int)) */
+    assert(value_size > 0 && value_size%sizeof(int) == 0);
+    assert((int)value_size == value_size);  /* check for overflow */
+
+    ok = FS_create(&bi->fs, filepath);
+    assert(ok);
+
+    bi->value_size = value_size;
+}
+
+void Bender_Impl_destroy(Bender_Impl *bi)
+{
+    FS_destroy(&bi->fs);
+}
+
+bool Bender_Impl_insert( Bender_Impl *bi,
+                         const void *key_data, size_t key_size )
+{
+    /* TODO */
+    assert(0);
+}
+
+bool Bender_Impl_contains( Bender_Impl *set,
+                           const void *key_data, size_t key_size )
+{
+    /* TODO */
+    assert(0);
+}

### 2008-05-02 datastructures/Bender_Impl.h

@@ -1,0 +1,38 @@
+#ifndef BENDER_IMPL_H_INCLUDED
+#define BENDER_IMPL_H_INCLUDED
+
+#include "FileStorage.h"
+
+/* Implementation of Bender's cache-oblivious set data structure.
+
+   Stores values upto a fixed size (which must be a multiple of sizeof(int))
+   and tags data with the size as well. As a result, if value_size == X,
+   then the data structure can store elements of size upto (X - sizeof(int)).
+*/
+
+typedef struct Bender_Impl Bender_Impl;
+
+struct Bender_Impl
+{
+    size_t value_size;
+    FileStorage fs;
+};
+
+/* Create a Bender set implementation. */
+void Bender_Impl_create( Bender_Impl *bi,
+                         const char *filepath, size_t value_size );
+
+/* Destroy a Bender set implementation and free all associated resources. */
+void Bender_Impl_destroy(Bender_Impl *bi);
+
+/* Insert a value into a set implementation.
+   Callers must ensure that (key_size + sizeof(int) <= bi->value_size) */
+bool Bender_Impl_insert( Bender_Impl *bi,
+                         const void *key_data, size_t key_size );
+
+/* Insert a value into a set implementation.
+   Callers must ensure that (key_size + sizeof(int) <= bi->value_size) */
+bool Bender_Impl_contains( Bender_Impl *set,
+                           const void *key_data, size_t key_size );
+
+#endif /* ndef BENDER_H_INCLUDED */

### 2008-05-02 datastructures/Bender_Set.c

@@ -1,0 +1,107 @@
+#include "config.h"
+#include "Set.h"
+#include "Bender_Impl.h"
+#include <assert.h>
+#include <stdio.h>
+#include <stdbool.h>
+#include <string.h>
+
+/* Set interface to Bender's cache-oblivious set data structure.
+
+   Since Bender's data structure works with fixed-size elements, this wrapper
+   implementation is required that keeps different sets storing elements of
+   different maximum sizes, ranging from 2**4 to 2**15. (Here, ** denotes
+   exponentiation.)
+
+   Values are stored in the implementation with an integer that denotes their
+   size, so the maximum size that can be stored is actually
+   (2**15) - sizeof(int) or slightly less than 32KB.
+*/
+
+typedef struct Bender_Set Bender_Set;
+
+struct Bender_Set
+{
+    Set         base;
+    Bender_Impl impl[12];
+};
+
+/* Computes the index of the set to use for storage.
+
+   Note that the implementation at index i stores values of size
+   upto and including 2**(4+i) but this must include an integer storing
+   the size of the data.
+
+   This function therefore computes and returns the smallest non-negative
+   integer i such that 2**(4+i) >= size + sizeof(int)
+*/
+static unsigned get_index(size_t size)
+{
+    int storage_size, index;
+
+    storage_size = (int)size + sizeof(int);
+    assert(storage_size > size); /* check for overflow */
+    index = (8*sizeof(int) - __builtin_clz(storage_size - 1)) - 4;
+
+    return index < 0 ? 0 : (unsigned)index;
+}
+
+static bool set_insert(Bender_Set *set, const void *key_data, size_t key_size)
+{
+    unsigned index = get_index(key_size);
+    assert(index < 12);
+    return Bender_Impl_contains(&set->impl[index], key_data, key_size);
+}
+
+static bool set_contains(Bender_Set *set, const void *key_data, size_t key_size)
+{
+    unsigned index = get_index(key_size);
+    assert(index < 12);
+    return Bender_Impl_contains(&set->impl[index], key_data, key_size);
+}
+
+/* Destroys a set data structure, by closing the backing file
+   and freeing all associated resources. */
+static void set_destroy(Bender_Set *set)
+{
+    int n;
+
+    for (n = 0; n < 12; ++n)
+        Bender_Impl_destroy(&set->impl[n]);
+    free(set);
+}
+
+/* Creates a set data structure. */
[…]

### 2008-05-02 datastructures/Set.c

@@ -6,7 +6,7 @@
 #include <unistd.h>
 
 typedef enum SetType {
-    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree
+    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree, Bender
 } SetType;
 
 static const char *make_tempfile()
@@ -39,6 +39,9 @@ static const char *make_tempfile()
 
     "BerkeleyDB hash .."
     Creates a BerkeleyDB hash table based set.
+
+    "bender"
+    Creates a cache-oblivious set (as proposed by Bender et al.)
 
     Common arguments:
     ".. [file=path]"
@@ -77,6 +80,11 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     if (strcmp(*argv, "BerkeleyDB") == 0)
     {
         type = BDB_Unspecified;
+    }
+    else
+    if (strcmp(*argv, "Bender") == 0)
+    {
+        type = Bender;
     }
     else
     {
@@ -146,18 +154,27 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     case Btree:
         result = Btree_Set_create(path, pagesize);
         break;
+
     case Hash:
         result = Hash_Set_create(path, capacity);
         break;
+
     case BDB_Btree:
         result = BDB_Btree_Set_create(path);
         break;
+
     case BDB_Hash:
         result = BDB_Hash_Set_create(path);
         break;
+
+    case Bender:
+        result = Bender_Set_create(path);
+        break;
+
     default:
         /* No valid set selected */
         result = NULL;
+        break;
     }
 
     if (!keep || !result)

### 2008-05-02 datastructures/Set.h

@@ -61,6 +61,10 @@ Set *Btree_Set_create(const char *filepath, size_t pagesize);
 /* Creates a set data structure backed by a custom hash table implementation. */
 Set *Hash_Set_create(const char *filepath, size_t capacity);
 
+/* Creates a cache-oblivious set data structure backed as proposed by
+   Bender at al. in "A locality-preserving cache-oblivious data structure". */
+Set *Bender_Set_create(const char *filepath);
+
 
 /* Default comparison function */
 int default_compare( const void *ignored,

### 2008-05-07 datastructures/Bender_Impl.c
Made allocation method more flexible. It is now easy to switch between malloc() and mmap() based allocations.

@@ -78,7 +78,7 @@ const bool opt_fast_update = true;
 
 
 /* Returns a pointer to the i-th element in the data array. */
-#define ARRAY_AT(i) ((ArrayNode*)(bi->fs.data+(i)*(sizeof(ArrayNode)+bi->V)))
+#define ARRAY_AT(i) ((ArrayNode*)(bi->data+(i)*(sizeof(ArrayNode)+bi->V)))
 
 /* Copy value of ArrayNode *q to ArrayNode *p while keeping the pointer data
    unmodified. */
@@ -92,7 +92,7 @@ const bool opt_fast_update = true;
     } while(0);
 
 /* Convert ArrayNode pointer to index */
-#define ARRAY_IDX(an) (((char*)(an) - bi->fs.data)/(sizeof(ArrayNode) + bi->V))
+#define ARRAY_IDX(an) (((char*)(an) - bi->data)/(sizeof(ArrayNode) + bi->V))
 
 /* Convert TreeNode pointer to index */
 #define TREE_IDX(tn) (((char*)(tn) - (char*)bi->tree)/(sizeof(TreeNode) + bi->V))
@@ -513,15 +513,14 @@ static void recompute_populations(Bender_Impl *bi, size_t i, size_t j)
 
 static void resize(Bender_Impl *bi, int new_order)
 {
-    bool ok;
     int l;
     size_t n;
 
     /* Resize file; we need C array elements and 2*C-1 tree nodes. */
-    ok = FS_resize(&bi->fs,
+    bi->data = (*bi->allocator)(&bi->alloc, bi->data, 
         ((sizeof(ArrayNode) + bi->V)<<new_order) +
         2*((sizeof(TreeNode) + bi->V)<<(new_order)) - 1);
-    assert(ok);
+    assert(bi->data != NULL);
 
     /* Write blank values to the new part of the array */
     for ( n = (bi->O > 0) ? ((size_t)1 << (bi->O)) : 0;
@@ -548,8 +547,8 @@ static void resize(Bender_Impl *bi, int new_order)
     recompute_populations(bi, 0, NUM_WINDOWS(bi->L - 1));
 
     /* Recreate tree index */
-    bi->tree = (TreeNode*)( bi->fs.data +
-                            ((sizeof(ArrayNode) + bi->V)<<new_order) );
+    bi->tree = (TreeNode*)(bi->data +
+                          ((sizeof(ArrayNode) + bi->V)<<new_order));
     create_tree(bi);
 
     /* Update entire tree */
[…]

### 2008-05-07 datastructures/Bender_Impl.h

@@ -1,7 +1,7 @@
 #ifndef BENDER_IMPL_H_INCLUDED
 #define BENDER_IMPL_H_INCLUDED
 
-#include "FileStorage.h"
+#include "Alloc.h"
 
 /* Implementation of Bender's cache-oblivious set data structure.
 
@@ -41,15 +41,15 @@ struct TreeNode
 
 struct Bender_Impl
 {
-    size_t      V;          /* Size of values */
-    int         O;          /* Order (capacity == pow(2,order)) */
-
+    size_t      V;              /* Size of values */
+    int         O;              /* Order (capacity == pow(2,order)) */
     int         L;              /* Number of levels */
     size_t      *upper_bound;   /* Population upper bound per level */
     size_t      *population;    /* Population per window */
-
-    TreeNode    *tree;
-    FileStorage fs;
+    TreeNode    *tree;          /* Pointer to index tree root */
+    char        *data;          /* Allocated data */
+    Allocator   *allocator;     /* Allocator function */
+    Alloc       alloc;          /* Allocator context */
 
     /* Custom comparison function */
     int (*compare)(const void *, const void *, size_t, const void *, size_t);
@@ -58,7 +58,7 @@ struct Bender_Impl
 
 /* Create a Bender set implementation. */
 void Bender_Impl_create( Bender_Impl *bi,
-                         const char *filepath, size_t value_size );
+                         Allocator *allocator, size_t value_size );
 
 /* Destroy a Bender set implementation and free all associated resources. */
 void Bender_Impl_destroy(Bender_Impl *bi);

### 2008-05-07 datastructures/Bender_Set.c

@@ -77,7 +77,7 @@ static void set_destroy(Bender_Set *set)
 }
 
 /* Creates a set data structure. */
-Set *Bender_Set_create(const char *filepath)
+Set *Bender_Set_create(Allocator *allocator)
 {
     Bender_Set *set;
     int index;
@@ -95,17 +95,7 @@ Set *Bender_Set_create(const char *filepath)
 
     /* Create statically sized sets */
     for (index = 0; index < 12; ++index)
-    {
-        char buf[1024];
-        int len;
-
-        len = snprintf(buf, sizeof(buf), "%s-%02d", filepath, 4 + index);
-        assert((size_t)len < sizeof(buf));
-
-        Bender_Impl_create(&set->impl[index], buf, 16 << index);
-
-        unlink(buf);
-    }
+        Bender_Impl_create(&set->impl[index], allocator, 16 << index);
 
     return &set->base;
 }

### 2008-05-07 datastructures/Btree_Set.c

@@ -9,7 +9,7 @@
 
 #include "config.h"
 #include "comparison.h"
-#include "FileStorage.h"
+#include "Alloc.h"
 #include "Set.h"
 #include <assert.h>
 #include <stdio.h>
@@ -40,7 +40,7 @@
 */
 
 /* Note: these macros assume 'set' is in scope */
-#define DATA(index)     (set->fs.data + (index)*set->pagesize)
+#define DATA(index)     (set->data + (index)*set->pagesize)
 #define IDX(p)          ((int*)(DATA(p) + set->pagesize))
 #define COUNT(p)        (IDX(p)[-1])          /* number of values in page */
 #define BEGIN(p, i)     (IDX(p)[-2*(i)-2])    /* offset to start of i-th value */
@@ -51,18 +51,20 @@
 
 typedef struct Btree_Set
 {
-    Set     base;
+    Set         base;
 
-    size_t  pagesize;       /* Size of data pages */
-    int     pages;          /* Number of pages */
-    int     root;           /* Index of root page */
+    size_t      pagesize;       /* Size of data pages */
+    int         pages;          /* Number of pages */
+    int         root;           /* Index of root page */
 
-    FileStorage fs;         /* File storage */
+    char        *data;          /* Allocated data */
+    Allocator   *allocator;     /* Allocator function */
+    Alloc       alloc;          /* Allocator context */
 
     /* Temporary memory pool */
-    char    *mem;           /* Allocated memory pool */
-    size_t  mem_size;       /* Total amount of memory allocated */
-    size_t  mem_used;       /* Memory used (set to zero before each operation) */
+    char        *mem;           /* Allocated memory pool */
+    size_t      mem_size;       /* Total amount of memory allocated */
+    size_t      mem_used;       /* Memory used (set to zero before each operation) */
 } Btree_Set;
 
 typedef struct PageEntry
@@ -127,7 +129,7 @@ static void *alloc_mem(Btree_Set *set, size_t size)
    and freeing all associated resources. */
 static void set_destroy(Btree_Set *set)
 {
-    FS_destroy(&set->fs);
+    set->allocator(&set->alloc, set->data, 0);
     free(set->mem);
     free(set);
 }
@@ -135,12 +137,11 @@ static void set_destroy(Btree_Set *set)
 /* Allocates a new page and returns its index. */
 static int create_page(Btree_Set *set)
 {
[…]

### 2008-05-07 datastructures/FileStorage.c

@@ -14,12 +14,8 @@
 bool FS_create(FileStorage *fs, const char *path)
 {
     int fd;
-    long chunk_size;
 
-    /* Get page size (used as the allocation chunk size) */
-    chunk_size = sysconf(_SC_PAGESIZE);
-    if (chunk_size <= 0)
-        return false;
+    assert(ALLOC_CHUNK_SIZE > 0);
 
     /* Open the file */
     fd = open(path, O_CREAT | O_RDWR, 0666);
@@ -27,26 +23,21 @@ bool FS_create(FileStorage *fs, const char *path)
         return false;
 
     /* Initialize FileStorage structure */
-    fs->data       = NULL;
     fs->size       = 0;
     fs->capacity   = 0;
     fs->fd         = fd;
-    fs->chunk_size = (size_t)chunk_size;
 
     return true;
 }
 
-void FS_destroy(FileStorage *fs)
+void FS_destroy(FileStorage *fs, void *data)
 {
-    if (fs->data != NULL)
-    {
-        munmap(fs->data, fs->size);
-        fs->data = NULL;
-    }
+    if (data != NULL)
+        munmap(data, fs->size);
     close(fs->fd);
 }
 
-bool FS_resize(FileStorage *fs, size_t size)
+void *FS_resize(FileStorage *fs, void *data, size_t size)
 {
     int res;
     size_t old_size, new_size;
@@ -55,51 +46,47 @@ bool FS_resize(FileStorage *fs, size_t size)
     if (size <= fs->capacity)
     {
         fs->size = size;
-        return true;
+        return data;
     }
 
     /* Round desired size up to the next page boundary. */
     old_size = fs->capacity;
     new_size = size;
-    if (new_size%fs->chunk_size != 0)
+    if (new_size%ALLOC_CHUNK_SIZE != 0)
     {
-        new_size += fs->chunk_size - new_size%fs->chunk_size;
+        new_size += ALLOC_CHUNK_SIZE - new_size%ALLOC_CHUNK_SIZE;
         if (new_size < size)
-            return false;   /* overflow */
+            return NULL;   /* overflow */
     }
 
     /* Change file size */
     assert(sizeof(off_t) == sizeof(size_t));
     res = ftruncate(fs->fd, (off_t)new_size);
     if (res != 0)
-        return false;
+        return NULL;
 
     /* Remap memory */
-    if (HAVE_MREMAP && fs->data != NULL)
+    if (HAVE_MREMAP && data != NULL)
     {
[…]

### 2008-05-07 datastructures/FileStorage.h

@@ -9,12 +9,9 @@ typedef struct FileStorage FileStorage;
 /* Models a memory-mapped storage area backed by a file. */
 struct FileStorage
 {
-    char    *data;          /* Memory mapped data */
     size_t  size;           /* Size of memory used */
     size_t  capacity;       /* Size of memory allocated */
-
     int     fd;             /* Open file descriptor */
-    size_t  chunk_size;     /* Allocation chunk size */
 };
 
 /* Creates an empty storage area backed by a file with the specified path.
@@ -23,7 +20,7 @@ struct FileStorage
 bool FS_create(FileStorage *fs, const char *path);
 
 /* Releases all associated resources */
-void FS_destroy(FileStorage *fs);
+void FS_destroy(FileStorage *fs, void *data);
 
 /* Resizes the used memory size.
    This may cause the underlying file to be extended, in which case ``data''
@@ -31,6 +28,6 @@ void FS_destroy(FileStorage *fs);
 
    Returns true if the request was completed succesfully, or returns false and
    sets errno. */
-bool FS_resize(FileStorage *fs, size_t new_size);
+void *FS_resize(FileStorage *fs, void *data, size_t new_size);
 
 #endif /* ndef FILE_STORAGE_H_INCLUDED */

### 2008-05-07 datastructures/File_Deque.c

@@ -37,7 +37,8 @@ struct FileDeque
     size_t      count;                  /* Number of elements */
     size_t      begin;                  /* Offset to start of data */
     size_t      end;                    /* Offset to end of data */
-    FileStorage fs;
+    char        *data;                  /* Allocated data */
+    FileStorage fs;                     /* FileStorage for allocated data */
 };
 
 /* Rounds argument up to a multiple of sizeof(size_t) */
@@ -50,7 +51,7 @@ static size_t align(size_t size)
 
 static void destroy(FileDeque *deque)
 {
-    FS_destroy(&deque->fs);
+    FS_destroy(&deque->fs, deque->data);
     free(deque);
 }
 
@@ -67,13 +68,19 @@ static bool empty(FileDeque *deque)
 static bool push_back(FileDeque *deque, const void *data, size_t size)
 {
     size_t aligned_size = align(size);
+    void *new_data;
 
-    if (!FS_resize(&deque->fs, deque->end + 2*sizeof(size_t) + aligned_size))
+    /* Reallocate deque */
+    new_data = FS_resize( &deque->fs, deque->data,
+                          deque->end + 2*sizeof(size_t) + aligned_size );
+    if (new_data == NULL)
         return false;
+    deque->data = new_data;
 
-    *(size_t*)(deque->fs.data + deque->end) = size;
-    memcpy(deque->fs.data + deque->end + sizeof(size_t), data, size);
-    *(size_t*)(deque->fs.data + deque->end +
+    /* Append item */
+    *(size_t*)(deque->data + deque->end) = size;
+    memcpy(deque->data + deque->end + sizeof(size_t), data, size);
+    *(size_t*)(deque->data + deque->end +
                sizeof(size_t) + aligned_size) = size;
     deque->end += 2*sizeof(size_t) + aligned_size;
     ++deque->count;
@@ -92,8 +99,8 @@ static bool get_back(FileDeque *deque, const void **data, size_t *size)
     if (deque->count == 0)
         return false;
 
-    *size = *(size_t*)(deque->fs.data + deque->end - sizeof(size_t));
-    *data = deque->fs.data + deque->end - sizeof(size_t) - align(*size);
+    *size = *(size_t*)(deque->data + deque->end - sizeof(size_t));
+    *data = deque->data + deque->end - sizeof(size_t) - align(*size);
     return true;
 }
 
[…]

### 2008-05-07 datastructures/Hash_Set.c

@@ -28,9 +28,12 @@ typedef struct Hash_Set Hash_Set;
 
 struct Hash_Set
 {
-    Set     base;
-    size_t  capacity;       /* Index capacity (number of buckets) */
-    FileStorage fs;         /* File storage */
+    Set         base;
+    size_t      capacity;       /* Index capacity (number of buckets) */
+    char        *data;          /* Allocated data */
+    size_t      size;           /* Allocated size */
+    Allocator   *allocator;     /* Allocator */
+    Alloc       alloc;          /* Allocator context */
 };
 
 
@@ -43,9 +46,9 @@ static void debug_print_data(Hash_Set *set, FILE *fp)
     fprintf(fp, "Capacity: %d buckets\n", (int)set->capacity);
     for (n = 0; n < set->capacity; ++n)
         fprintf( fp, "Bucket %d: offset %d\n",
-                 (int)n, (int)((size_t*)set->fs.data)[n] );
+                 (int)n, (int)((size_t*)set->data)[n] );
 
-    for (n = 0; n < set->fs.size; ++n)
+    for (n = 0; n < set->size; ++n)
     {
         if (n > 0)
         {
@@ -60,44 +63,18 @@ static void debug_print_data(Hash_Set *set, FILE *fp)
                     putc(' ', fp);
             }
         }
-        putc("0123456789ABCDEF"[(set->fs.data[n]>>4)&15], fp);
-        putc("0123456789ABCDEF"[(set->fs.data[n]>>0)&15], fp);
+        putc("0123456789ABCDEF"[(set->data[n]>>4)&15], fp);
+        putc("0123456789ABCDEF"[(set->data[n]>>0)&15], fp);
     }
     putc('\n', fp);
     fprintf(fp, "----\n");
 }
 
-/* Resize the data file to the given size.
-   NB. This invalidates pointers to set->fs.data! */
-static void resize(Hash_Set *set, size_t size)
+static void resize(Hash_Set *set, size_t new_size)
 {
-    int res;
-    bool resize_ok;
-
-    if (size <= set->fs.capacity)
-    {
-        set->fs.size = size;
-        return;
-    }
-
-    /* Unlock memory */
-    if (HAVE_MLOCK && USE_MLOCK && set->fs.data != NULL)
-    {
-        res = munlock(set->fs.data, set->capacity*sizeof(size_t));
-        assert(res == 0);
-    }
-
-    /* Resize */
[…]

### 2008-05-07 datastructures/Set.c

@@ -9,17 +9,6 @@ typedef enum SetType {
     Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree, Bender
 } SetType;
 
-static const char *make_tempfile()
-{
-    int res;
-    static char path[] = "db-XXXXXX";
-
-    res = mkstemp(path);
-    assert(res >= 0);
-    close(res);
-
-    return path;
-}
 
 /*  Creates a set data structure from a string description.
     Arguments between square brackets are optional.
@@ -34,36 +23,35 @@ static const char *make_tempfile()
     "hash [capacity=C] .."
     Creates a hash table based with capacity C items (default: 1,000,000).
 
-    "BerkeleyDB btree .."
+    "BerkeleyDB path=FP btree .."
     Creates a BerkeleyDB B-tree based set.
 
-    "BerkeleyDB hash .."
+    "BerkeleyDB path=FP hash .."
     Creates a BerkeleyDB hash table based set.
 
     "Bender"
     Creates a cache-oblivious set (as proposed by Bender et al.)
 
-    Common arguments:
-    ".. [file=path]"
-    Specifies a path to use to create the backing file. Note that if this
-    file exists, it will be overwritten! If no file is specified, a file is
-    created in the working directory.
+    Optional arguments:
 
-    ".. [keep]"
-    Keep the temporary file (i.e. do not unlink it after creation).
+    [malloc]
+    Use the malloc allocator (default)
+
+    [mmap]
+    Use the mmap (file backed) allocator
 */
 Set *Set_create_from_args(int argc, const char * const *argv)
 {
     SetType type;
-    int pagesize, capacity, keep;
+    int pagesize, capacity;
     char *path;
     Set *result;
+    Allocator *allocator = NULL;
 
     if (argc < 1)
         return NULL;
 
     path = NULL;
-    keep = 0;
 
     if (strcmp(*argv, "btree") == 0)
     {
@@ -123,15 +111,24 @@ Set *Set_create_from_args(int argc, const char * const *argv)
                 return NULL;
         }
         else
-        if (path == NULL && sscanf(*argv, "file=%as", &path))
+        if (path == NULL && sscanf(*argv, "file=%as", &path) == 1)
         {
             if (path == NULL)
                 return NULL;
         }
         else
-        if (strcmp(*argv, "keep") == 0)
+        if (strcmp(*argv, "malloc") == 0)
         {
-            keep = 1;
+            if (allocator != NULL)
+                return NULL;
+            allocator = Allocator_malloc;
+        }
[…]

### 2008-05-07 datastructures/Set.h

@@ -3,6 +3,7 @@
 
 #include <stdbool.h>
 #include <stdlib.h>
+#include "Alloc.h"
 
 typedef struct Set Set;
 
@@ -56,13 +57,13 @@ Set *BDB_Btree_Set_create(const char *filepath);
 Set *BDB_Hash_Set_create(const char *filepath);
 
 /* Creates a set data structure backed by a custom B-tree implementation. */
-Set *Btree_Set_create(const char *filepath, int pagesize);
+Set *Btree_Set_create(Allocator *alloc, int pagesize);
 
 /* Creates a set data structure backed by a custom hash table implementation. */
-Set *Hash_Set_create(const char *filepath, size_t capacity);
+Set *Hash_Set_create(Allocator *alloc, size_t capacity);
 
 /* Creates a cache-oblivious set data structure backed as proposed by
    Bender at al. in "A locality-preserving cache-oblivious data structure". */
-Set *Bender_Set_create(const char *filepath);
+Set *Bender_Set_create(Allocator *alloc);
 
 #endif /* ndef SET_H_INCLUDED */

### 2008-05-07 datastructures/config.h

@@ -1,4 +1,3 @@
 #define _GNU_SOURCE
 #define HAVE_MREMAP 1
-#define HAVE_MLOCK 1
-#define USE_MLOCK 0
+#define ALLOC_CHUNK_SIZE 4096

### 2008-05-07 datastructures/FileStorage.c
Switched to using private mapping of /dev/zero for the mmap allocator.

@@ -62,8 +62,12 @@ void *FS_resize(FileStorage *fs, void *data, size_t size)
     /* Change file size */
     assert(sizeof(off_t) == sizeof(size_t));
     res = ftruncate(fs->fd, (off_t)new_size);
+    /* truncation may fail for special files like /dev/zero; in that case,
+       this not an error. (If it is, then mmap/mremap will fail later on) */
+    /*
     if (res != 0)
         return NULL;
+    */
 
     /* Remap memory */
     if (HAVE_MREMAP && data != NULL)

### 2008-05-07 datastructures/Set.c

@@ -34,11 +34,12 @@ typedef enum SetType {
 
     Optional arguments:
 
-    [malloc]
-    Use the malloc allocator (default)
-
     [mmap]
-    Use the mmap (file backed) allocator
+    Use the mmap (file backed) allocator (default)
+
+    [malloc]
+    Use the malloc allocator
+
 */
 Set *Set_create_from_args(int argc, const char * const *argv)
 {
@@ -139,7 +140,7 @@ Set *Set_create_from_args(int argc, const char * const *argv)
 
     /* Set default allocator */
     if (allocator == NULL)
-        allocator = Allocator_malloc;
+        allocator = Allocator_mmap;
 
     switch (type)
     {

### 2008-05-07 search/main.c

@@ -107,7 +107,7 @@ int main(int argc, char *argv[])
     }
 
     /* Create deque data structure */
-    deque = Memory_Deque_create();
+    deque = File_Deque_create("/dev/zero");
     if (deque == NULL)
     {
         perror("Could not create deque");

### 2008-05-07 datastructures/Alloc.c
Added missing files!

@@ -1,0 +1,63 @@
+#include "Alloc.h"
+#include "config.h"
+#include <assert.h>
+#include <unistd.h>
+
+/* Uses malloc to allocate memory. */
+void *Allocator_malloc(Alloc *a, void *old_data, size_t size)
+{
+    void *new_data;
+    size_t new_size;
+
+    if (size == 0)
+    {
+        /* Free allocation */
+        if (old_data != NULL)
+            free(old_data);
+        return NULL;
+    }
+
+    if (old_data != NULL && a->ms.capacity >= size)
+    {
+        return old_data;
+    }
+
+    new_size = size;
+    if (new_size%ALLOC_CHUNK_SIZE != 0)
+    {
+        new_size += ALLOC_CHUNK_SIZE - new_size%ALLOC_CHUNK_SIZE;
+        if (new_size < size)
+            return NULL;   /* overflow */
+    }
+    new_data = realloc(old_data, new_size);
+    if (new_data == NULL)
+        return NULL;
+    a->ms.capacity = new_size;
+    return new_data;
+}
+
+/* Uee the FileStorage to mmap() data backed by a temporary file. */
+void *Allocator_mmap(Alloc *a, void *old_data, size_t size)
+{
+    if (size == 0)
+    {
+        /* Free allocation */
+        if (old_data != NULL)
+            FS_destroy(&a->fs, old_data);
+        return NULL;
+    }
+    else
+    if (old_data == NULL)
+    {
+        assert(size > 0);
+        if (!FS_create(&a->fs, "/dev/zero"))
+            return NULL;
+
+        return FS_resize(&a->fs, NULL, size);
+    }
+    else
+    {
+        /* Resize allocation */
+        return FS_resize(&a->fs, old_data, size);
+    }
+}

### 2008-05-07 datastructures/Alloc.h

@@ -1,0 +1,34 @@
+#ifndef ALLOC_H_INCLUDED
+#define ALLOC_H_INCLUDED
+
+#include "FileStorage.h"
+
+typedef union Alloc Alloc;
+typedef struct MemStorage MemStorage;
+
+/* Describes an allocator function.
+
+   Pass NULL as old to create a new allocation with initial size ``size'',
+   pass a valid allocation to resize, or pass (size_t)-1 as size to deallocate.
+
+   The return value is either a non-NULL pointer (if allocation succeeds) or
+   NULL if allocation fails (in that case, the old pointer should still be
+   valid). On deallocation, NULL is returned.
+*/
+typedef void *(Allocator)(Alloc *a, void *old, size_t size);
+
+void *Allocator_malloc(Alloc *a, void *old, size_t size);
+void *Allocator_mmap(Alloc *a, void *old, size_t size);
+
+struct MemStorage
+{
+    size_t capacity;        /* Size of memory allocated */
+};
+
+union Alloc
+{
+    FileStorage fs;
+    MemStorage  ms;
+};
+
+#endif /* ndef ALLOC_H_INCLUDED */

### 2008-05-07 datastructures/Bender_Impl.c
Removed confusing comments about height division in Bender set implementation.

@@ -292,25 +292,9 @@ static TreeNode *create_subtree(
         /* Create a subtree by dividing the height into two halves, first
            creating the top subtree, and then creating the bottom subtrees,
            which are connected to the leaf nodes of the top subtree. */
-
-        /* NOTE: if this is slow, we can easily pre-compute the required
-                 splits since height will be small (less than 40 or so). */
-        int top_height, bottom_height;
         TreeNode *root, *leaf;
 
-        /* Bender specifies the size of the bottom tree should be a power
-           of two (without giving a rationale): */
-        /*
-        bottom_height = 1;
-        while (2*bottom_height < height)
-            bottom_height *= 2;
-        top_height = height - bottom_height;
-        */
-        /* Instead, we use height/2 which performs slightly better. */
-        top_height = height/2;
-        bottom_height = height - top_height;
-
-        root = create_subtree(bi, tree_pos, array_pos, top_height, level);
+        root = create_subtree(bi, tree_pos, array_pos, height/2, level);
 
         /* Find first leaf node in subtree */
         leaf = root;
@@ -320,10 +304,10 @@ static TreeNode *create_subtree(
         /* Create all subtrees */
         do {
             leaf->left = create_subtree( bi, tree_pos, array_pos,
-                                         bottom_height, level + top_height );
+                                         height - height/2, level + height/2 );
             leaf->left->parent = leaf;
             leaf->right = create_subtree( bi, tree_pos, array_pos,
-                                          bottom_height, level + top_height );
+                                          height - height/2, level + height/2 );
             leaf->right->parent = leaf;
         } while ((leaf = next_leaf(leaf)) != NULL);
 

### 2008-05-07 datastructures/Bender_Impl.c
Added mock set implementation.

@@ -25,28 +25,37 @@
    Therefore, there are L=O-ceil(log2(O)) levels (numbered from 0 to L-1);
    level x consists of 2^x windows of size C/(2^x).
 
-   For each window (at each level) a population count is kept (the number of
-   non-blank values stored in the window).
+   For the windows on the lowest level (level L-1) a population count is kept
+   (the number of non-blank values stored in the window). These counts are
+   used to compute the population counts of higher-level windows on the fly.
 
    Each level has an associated lower and upper bound on the number of values
    stored in it. If a window stores more (/less) values than its upper
    (/lower) bound permits, it is said to be overflowing (/underflowing).
+   Note that this is not an error condition; it is just a state that will be
+   rectified when a higher-level window is rebalanced.
 
-   The index is a binary search tree; the leaf nodes store the same elements
-   as the array (i.e. the i-th leaf stores the i-th element in the array,
+   Since we will only insert new elements in the array, undeflow cannot occur
+   and we will ignore it entirely.
+
+   The upper bounds for different levels are chosen in such a way that windows
+   on the lowest level have a density of 1 (the upper bound is then equal to
+   the window size) while higher levels have a lower bound.
+
+   New values are inserted in the lowest level (with the highest number)
+   that is not overflowing and then the entire window is rebalanced (meaning
+   that blank and non-blank values are evenly redistributed) and the index is
+   updated accordingly.
+
+   For fast finding of elements (both when inserting and looking up element)
+   a binary search tree index is kept. Its leaf nodes store the same elements
+   as the array (i.e. the i-th leaf node stores the i-th element in the array,
    which may be blank if there is no value at index i in the sparse array)
    while each interior node stores the maximum non-blank value of its children
    (or a blank value if there are no non-blank children).
 
-   New values are inserted in the lowest level (with the highest number)
-   that is not overflowing and not underflowing, and then the entire window
-   is rebalanced (meaning that blank and non-blank values are evenly
-   redistributed) and the index is updated accordingly.
-
-   Since we will only insert new elements in the array, we will ignore
-   underflow entirely, only rebalancing when overflow occurs. When the single
-   window at level 0 overflows, the size of the array (and its index) is
-   doubled (and the window then rebalanced).
+   The tree index then is stored in van Emde Boas lay-out to make it
+   cache-friendly without requiring parametrization.
 */
 
 

### 2008-05-07 datastructures/Mock_Set.c

@@ -1,0 +1,124 @@
+#include "Set.h"
+#include "comparison.h"
+#include <assert.h>
+#include <stdio.h>
+
+typedef struct Mock_Set Mock_Set;
+
+struct Mock_Set
+{
+    Set     base;
+    Set     *impl;      /* Actual implementation (when recording) */
+    FILE    *fp;        /* File being written to/read from */
+};
+
+static bool set_insert(Mock_Set *set, const void *key_data, size_t key_size)
+{
+    int ch;
+    bool res;
+
+    if (set->impl != NULL)
+    {
+        /* Use real set implementation to obtain result */
+        set->impl->hash    = set->base.hash;
+        set->impl->compare = set->base.compare;
+        res = set->impl->insert(set->impl, key_data, key_size);
+
+        /* Write result to file */
+        ch = fputc(res, set->fp);
+        assert(ch == 0 || ch == 1);
+        return ch;
+    }
+    else
+    {
+        /* Read result from file */
+        ch = fgetc(set->fp);
+        assert(ch == 0 || ch == 1);
+        res = ch;
+    }
+
+    return res;
+}
+
+static bool set_contains(Mock_Set *set, const void *key_data, size_t key_size)
+{
+    int ch;
+    bool res;
+
+    if (set->impl != NULL)
+    {
+        /* Use real set implementation to obtain result */
+        res = set->impl->insert(set->impl, key_data, key_size);
+
+        /* Write result to file */
+        ch = fputc(res, set->fp);
+        assert(ch == 0 || ch == 1);
+        return ch;
+    }
+    else
+    {
+        /* Read result from file */
+        ch = fgetc(set->fp);
+        assert(ch == 0 || ch == 1);
+        res = ch;
+    }
+
+    return res;
+}
+
+static void set_destroy(Mock_Set *set)
+{
+    if (set->impl != NULL)
+    {
+        set->impl->destroy(set->impl);
+    }
+    fclose(set->fp);
+    free(set);
+}
+
+Set *Mock_Set_create(const char *path, bool recording)
+{
+    Mock_Set    *set;
+    Set         *impl;
+    FILE        *fp;
+
+    /* Allocate memory */
+    set = malloc(sizeof(Mock_Set));
+    if (set == NULL)
+        return NULL;
+
+    /* Create real set instance, if recording */
+    if (recording)
+    {
+        impl = Btree_Set_create(Allocator_mmap, 4096);
+        if (impl == NULL)
+        {
[…]

### 2008-05-07 datastructures/Set.c

@@ -6,7 +6,7 @@
 #include <unistd.h>
 
 typedef enum SetType {
-    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree, Bender
+    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree, Bender, Mock
 } SetType;
 
 
@@ -23,23 +23,25 @@ typedef enum SetType {
     "hash [capacity=C] .."
     Creates a hash table based with capacity C items (default: 1,000,000).
 
-    "BerkeleyDB path=FP btree .."
+    "BerkeleyDB btree path=FP .."
     Creates a BerkeleyDB B-tree based set.
 
-    "BerkeleyDB path=FP hash .."
+    "BerkeleyDB hash path=FP .."
     Creates a BerkeleyDB hash table based set.
 
     "Bender"
     Creates a cache-oblivious set (as proposed by Bender et al.)
 
-    Optional arguments:
+    "Mock path=FP [record|replay]"
+    Creates a mock implementation recording/replaying to/from a file.
+
+    Common arguments:
 
     [mmap]
-    Use the mmap (file backed) allocator (default)
+    Use the mmap allocator (default)
 
     [malloc]
     Use the malloc allocator
-
 */
 Set *Set_create_from_args(int argc, const char * const *argv)
 {
@@ -47,12 +49,16 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     int pagesize, capacity;
     char *path;
     Set *result;
-    Allocator *allocator = NULL;
+    Allocator *allocator;
+    bool record, replay;
 
     if (argc < 1)
         return NULL;
 
     path = NULL;
+    allocator = NULL;
+    record = false;
+    replay = false;
 
     if (strcmp(*argv, "btree") == 0)
     {
@@ -74,6 +80,11 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     if (strcmp(*argv, "Bender") == 0)
     {
         type = Bender;
+    }
+    else
+    if (strcmp(*argv, "mock") == 0)
+    {
+        type = Mock;
     }
     else
     {
@@ -116,6 +127,8 @@ Set *Set_create_from_args(int argc, const char * const *argv)
         {
             if (path == NULL)
                 return NULL;
+            if (!(type == BDB_Btree || type == BDB_Hash || type == Mock))
+                return NULL;
         }
         else
         if (strcmp(*argv, "malloc") == 0)
[…]

### 2008-05-07 datastructures/Set.h

@@ -66,4 +66,8 @@ Set *Hash_Set_create(Allocator *alloc, size_t capacity);
    Bender at al. in "A locality-preserving cache-oblivious data structure". */
 Set *Bender_Set_create(Allocator *alloc);
 
+/* Creates a mock set data structure that records/replays answers to/from the
+   given file path. This is useful for benchmarking purposes. */
+Set *Mock_Set_create(const char *filepath, bool record);
+
 #endif /* ndef SET_H_INCLUDED */

### 2008-05-07 datastructures/Set.c
Added dummy set implementation.

@@ -6,7 +6,7 @@
 #include <unistd.h>
 
 typedef enum SetType {
-    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree, Bender, Mock
+    Btree, Hash, BDB_Unspecified, BDB_Hash, BDB_Btree, Bender, Mock, Dummy
 } SetType;
 
 
@@ -85,6 +85,11 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     if (strcmp(*argv, "mock") == 0)
     {
         type = Mock;
+    }
+    else
+    if (strcmp(*argv, "dummy") == 0)
+    {
+        type = Dummy;
     }
     else
     {
@@ -199,6 +204,10 @@ Set *Set_create_from_args(int argc, const char * const *argv)
         if (!(record || replay) || path == NULL)
             return NULL;
         result = Mock_Set_create(path, record);
+        break;
+
+    case Dummy:
+        result = Dummy_Set_create();
         break;
 
     default:

### 2008-05-07 datastructures/Set.h

@@ -70,4 +70,8 @@ Set *Bender_Set_create(Allocator *alloc);
    given file path. This is useful for benchmarking purposes. */
 Set *Mock_Set_create(const char *filepath, bool record);
 
+/* Creates a dummy set data structure that always returns false.
+   This is useful for benchmarking purposes. */
+Set *Dummy_Set_create();
+
 #endif /* ndef SET_H_INCLUDED */

### 2008-05-07 datastructures/Mock_Set.c
Changed mock set implementation so it uses a mmap()ed file while recording. This improves performance; replaying is now nearly as fast as using the dummy set, so this is good enough as a base level for comparison.

@@ -2,23 +2,35 @@
 #include "comparison.h"
 #include <assert.h>
 #include <stdio.h>
+#include <fcntl.h>
+#include <sys/mman.h>
+#include <sys/types.h>
+#include <sys/stat.h>
+#include <unistd.h>
 
 typedef struct Mock_Set Mock_Set;
 
 struct Mock_Set
 {
     Set     base;
-    Set     *impl;      /* Actual implementation (when recording) */
-    FILE    *fp;        /* File being written to/read from */
+
+    /* When recording: */
+    Set     *impl;
+    FILE    *fp;
+
+    /* When replaying */
+    char    *begin, *end, *pos;
+    int     fd;
 };
 
 static bool set_insert(Mock_Set *set, const void *key_data, size_t key_size)
 {
-    int ch;
     bool res;
 
     if (set->impl != NULL)
     {
+        char ch;
+
         /* Use real set implementation to obtain result */
         set->impl->hash    = set->base.hash;
         set->impl->compare = set->base.compare;
@@ -27,14 +39,12 @@ static bool set_insert(Mock_Set *set, const void *key_data, size_t key_size)
         /* Write result to file */
         ch = fputc(res, set->fp);
         assert(ch == 0 || ch == 1);
-        return ch;
     }
     else
     {
         /* Read result from file */
-        ch = fgetc(set->fp);
-        assert(ch == 0 || ch == 1);
-        res = ch;
+        assert(set->pos < set->end);
+        res = *set->pos++;
     }
 
     return res;
@@ -42,25 +52,24 @@ static bool set_insert(Mock_Set *set, const void *key_data, size_t key_size)
 
 static bool set_contains(Mock_Set *set, const void *key_data, size_t key_size)
 {
-    int ch;
     bool res;
 
     if (set->impl != NULL)
     {
+        char ch;
+
         /* Use real set implementation to obtain result */
-        res = set->impl->insert(set->impl, key_data, key_size);
+        res = set->impl->contains(set->impl, key_data, key_size);
 
         /* Write result to file */
         ch = fputc(res, set->fp);
         assert(ch == 0 || ch == 1);
-        return ch;
     }
     else
     {
         /* Read result from file */
-        ch = fgetc(set->fp);
-        assert(ch == 0 || ch == 1);
-        res = ch;
+        assert(set->pos < set->end);
+        res = *set->pos++;
     }
 
     return res;
[…]

### 2008-05-07 datastructures/Dummy_Set.c
Added missing file: Dummy_Set.c (implementing the dummy set)

@@ -1,0 +1,35 @@
+#include "Set.h"
+
+static bool insert(Set *set, const void *key_data, size_t key_size)
+{
+    return false;
+}
+
+static bool contains(Set *set, const void *key_data, size_t key_size)
+{
+    return false;
+}
+
+static void destroy(Set *set)
+{
+    free(set);
+}
+
+Set *Dummy_Set_create()
+{
+    Set *set;
+
+    /* Allocate memory */
+    set = malloc(sizeof(Set));
+    if (set == NULL)
+        return NULL;
+
+    set->context  = NULL;
+    set->destroy  = destroy;
+    set->insert   = insert;
+    set->contains = contains;
+    set->hash     = NULL;
+    set->compare  = NULL;
+
+    return set;
+}

### 2008-05-07 datastructures/Bender_Impl.c
Made density parameter of Bender set configurable.

@@ -59,20 +59,7 @@
 */
 
 
-/* Maximum density on level 0.
-
-   Since the capacity is doubled when this is exceeded, the minimum density at
-   any time is actually max_density/2.
-
-   It isn't clear what the optimal value should be; low values (around 0.2)
-   paradoxically seem to work pretty well (despite the fact that they require
-   a lot of storage space, worst case).
-
-   FIXME: Make this configurable per set instance?
-*/
-const double max_density = 0.25;
-
-/* Enable an optimization that makes updates somewhat (~50%) faster.
+/* Enable an optimization that makes updates somewhat (~33%) faster.
 
    Normally, updates are performed by finding the smallest non-overflowing
    window where a new element can be inserted before its successor, and then
@@ -92,8 +79,7 @@ const bool opt_fast_update = true;
 /* Copy value of ArrayNode *q to ArrayNode *p while keeping the pointer data
    unmodified. */
 #define ARRAY_COPY(p, q)                                                    \
-    do { assert(q->size != (size_t)-1);                                     \
-         if (p == q) break; /* Necessary because memset doesn't allow       \
+    do { if (p == q) break; /* Necessary because memset doesn't allow       \
                                overlapping memory regions. */               \
          memcpy( (char*)(p) + offsetof(ArrayNode, size),                    \
                  (char*)(q) + offsetof(ArrayNode, size),                    \
@@ -412,7 +398,6 @@ static void overwrite_blank( Bender_Impl *bi, size_t i,
     size_t win;
 
     /* Set value */
-    assert(ARRAY_AT(i)->size == (size_t)-1);
     ARRAY_AT(i)->size = size;
     memcpy(ARRAY_AT(i)->data, data, size);
 
@@ -533,7 +518,7 @@ static void resize(Bender_Impl *bi, int new_order)
     for (l = 0; l < bi->L; ++l)
     {
         bi->upper_bound[l] = (size_t)WINDOW_SIZE(l)*
-            (max_density + (1 - max_density)*l/(bi->L - 1));
+            (bi->density + (1 - bi->density)*l/(bi->L - 1));
     }
 
     /* Recompute population counts */
@@ -605,7 +590,6 @@ static void insert_and_redistribute (Bender_Impl *bi,
             if (ARRAY_AT(p)->size == (size_t)-1)
                 break;
         }
[…]

### 2008-05-07 datastructures/Bender_Impl.h

@@ -44,6 +44,7 @@ struct Bender_Impl
     size_t      V;              /* Size of values */
     int         O;              /* Order (capacity == pow(2,order)) */
     int         L;              /* Number of levels */
+    double      density;        /* Upper bound on density on level 0 */
     size_t      *upper_bound;   /* Population upper bound per level */
     size_t      *population;    /* Population per window */
     TreeNode    *tree;          /* Pointer to index tree root */
@@ -57,8 +58,8 @@ struct Bender_Impl
 };
 
 /* Create a Bender set implementation. */
-void Bender_Impl_create( Bender_Impl *bi,
-                         Allocator *allocator, size_t value_size );
+void Bender_Impl_create( Bender_Impl *bi, Allocator *allocator,
+                         size_t value_size, double density );
 
 /* Destroy a Bender set implementation and free all associated resources. */
 void Bender_Impl_destroy(Bender_Impl *bi);

### 2008-05-07 datastructures/Bender_Set.c

@@ -77,7 +77,7 @@ static void set_destroy(Bender_Set *set)
 }
 
 /* Creates a set data structure. */
-Set *Bender_Set_create(Allocator *allocator)
+Set *Bender_Set_create(Allocator *allocator, double density)
 {
     Bender_Set *set;
     int index;
@@ -95,7 +95,7 @@ Set *Bender_Set_create(Allocator *allocator)
 
     /* Create statically sized sets */
     for (index = 0; index < 12; ++index)
-        Bender_Impl_create(&set->impl[index], allocator, 16 << index);
+        Bender_Impl_create(&set->impl[index], allocator, 16 << index, density);
 
     return &set->base;
 }

### 2008-05-07 datastructures/Set.c

@@ -29,8 +29,9 @@ typedef enum SetType {
     "BerkeleyDB hash path=FP .."
     Creates a BerkeleyDB hash table based set.
 
-    "Bender"
+    "Bender density=0.x"
     Creates a cache-oblivious set (as proposed by Bender et al.)
+    The density parameter must be in range [0-1] (default: 0.5)
 
     "Mock path=FP [record|replay]"
     Creates a mock implementation recording/replaying to/from a file.
@@ -51,6 +52,7 @@ Set *Set_create_from_args(int argc, const char * const *argv)
     Set *result;
     Allocator *allocator;
     bool record, replay;
+    double density = -1;
 
     if (argc < 1)
         return NULL;
@@ -163,6 +165,13 @@ Set *Set_create_from_args(int argc, const char * const *argv)
                 return NULL;
             replay = true;
         }
+        if (density == -1 && sscanf(*argv, "density=%lf", &density) == 1)
+        {
+            if (type != Bender)
+                return NULL;
+            if (density < 0 || density > 1)
+                return NULL;
+        }
         else
         {
             /* no option matched! */
@@ -197,7 +206,10 @@ Set *Set_create_from_args(int argc, const char * const *argv)
         break;
 
     case Bender:
-        result = Bender_Set_create(allocator);
+        /* Set default density (if none specified) */
+        if (density == -1)
+            density = 0.5;
+        result = Bender_Set_create(allocator, density);
         break;
 
     case Mock:

### 2008-05-07 datastructures/Set.h

@@ -63,8 +63,12 @@ Set *Btree_Set_create(Allocator *alloc, int pagesize);
 Set *Hash_Set_create(Allocator *alloc, size_t capacity);
 
 /* Creates a cache-oblivious set data structure backed as proposed by
-   Bender at al. in "A locality-preserving cache-oblivious data structure". */
-Set *Bender_Set_create(Allocator *alloc);
+   Bender at al. in "A locality-preserving cache-oblivious data structure".
+
+   Density is the density of the top-level window before it is overflowing
+   (between 0 and 1). Low values make updates cheaper, but result in larger
+   files. */
+Set *Bender_Set_create(Allocator *alloc, double density);
 
 /* Creates a mock set data structure that records/replays answers to/from the
    given file path. This is useful for benchmarking purposes. */

### 2008-05-08 search/main.c
Added some notes to the paper (for my own use).

@@ -16,9 +16,9 @@ static void usage()
         "Usage:\n"
         "    search [<options>] (set description)\n"
         "Options:\n"
-        "    -b          -- use breadth-first search\n"
-        "    -d          -- use depth-first search (default)\n"
-        "    -f bytecode -- path to bytecode field\n");
+        "    -B          -- use breadth-first search (default)\n"
+        "    -D          -- use depth-first search\n"
+        "    -m model    -- path to model bytecode file\n");
     exit(1);
 }
 
@@ -29,22 +29,22 @@ static void parse_args(int argc, char *argv[])
 
     if (argc < 2) usage();
 
-    while ((ch = getopt(argc, argv, "bdf:")) >= 0)
+    while ((ch = getopt(argc, argv, "BDm:")) >= 0)
     {
         switch (ch)
         {
-        case 'b':
+        case 'B':
             bfs = true;
             break;
 
-        case 'd':
+        case 'D':
             dfs = true;
             break;
 
-        case 'f':
+        case 'm':
             if (opt_bytecode_path != NULL)
             {
-                printf("At most one bytecode file can be specified!\n\n");
+                printf("At most one model can be specified!\n\n");
                 usage();
             }
             opt_bytecode_path = optarg;
@@ -57,7 +57,7 @@ static void parse_args(int argc, char *argv[])
 
     if (bfs && dfs)
     {
-        printf("At most one of -b or -d may be given!\n\n");
+        printf("At most one of -B or -D may be given!\n\n");
         usage();
     }
     else
@@ -67,7 +67,7 @@ static void parse_args(int argc, char *argv[])
 
     if (opt_bytecode_path == NULL)
     {
-        printf("A bytecode file must be specified!\n\n");
+        printf("A model must be specified!\n\n");
         usage();
     }
 

### 2008-05-08 search/search.c

@@ -22,14 +22,56 @@ typedef struct SearchContext
 void hex_print(FILE *fp, void *data, size_t size)
 {
     unsigned char *p = data;
-    printf("size=%d %d\n", (int)nipsvm_state_size(data), (int)size);
-    return;
+    size_t i;
+
+    for (i = 0; i < size; ++i)
+    {
+        fputc("0123456789abcdef"[*p++&15], fp);
+        fputc("0123456789abcdef"[*p++>>4], fp);
+        if (i%16 == 0)
+        {
+            fputc('\n', fp);
+        }
+        else
+        {
+            fputc(' ', fp);
+            if (i%8 == 0)
+                fputc(' ', fp);
+        }
+    }
+    fputc('\n', fp);
+}
+
+/* For debugging */
+void base64_print(FILE *fp, void *data, size_t size)
+{
+    unsigned char *p = data;
+    unsigned int v;
+    static const char *digits =
+        "ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz0123456789+/";
+
     while (size > 0)
     {
-        fputc("0123456789ABCDEF"[*p&15], fp);
-        fputc("0123456789ABCDEF"[*p>>4], fp);
-        --size, ++p;
+        v = (size > 0 ? (p[0] << 16) : 0) |
+            (size > 1 ? (p[1] << 8)  : 0) |
+            (size > 2 ? (p[2] << 0)  : 0);
+        fputc(           digits[(v >> 18)&63]      , fp);
+        fputc(           digits[(v >> 12)&63]      , fp);
+        fputc(size > 1 ? digits[(v >>  6)&63] : '=', fp);
+        fputc(size > 2 ? digits[(v >>  0)&63] : '=', fp);
+        if (size < 3)
+        {
+            p += size;
+            size = 0;
+        }
+        else
+        {
+            p    += 3;
+            size -= 3;
+        }
     }
+
+    fputc('\n', fp);
 }
 
 /* Makes a copy of the given state with specified size in dynamic memory
@@ -153,6 +195,9 @@ static int breadth_first_search(SearchContext *sc)
     {
         if (!queue->get_front(queue, (void**)&state, &state_size))
             return -1;
+
+        base64_print(stdout, state, state_size);
+        fflush(stdout);
 
         if (!expand_state(sc, state))
             return -1;

### 2008-05-08 search/models/build.sh
Added PROMELA example models and scripts to compile them to NIPS VM bytecode.

@@ -1,0 +1,16 @@
+#!/bin/sh
+
+NIPSHOME=../../nips_vm/
+NIPSASM="${NIPSHOME}"/nips_asm.pl
+CLASSPATH=../../nips_c/lib/nips-promela-compiler.jar
+
+mkdir -p build
+for file in $@
+do
+    base=`basename "$file"`
+    cpp "$file" | grep -v '^#' > build/"$base"
+    java xml.XmlSerializer build/"$base" && \
+    java CodeGen.CodeGen build/"$base".ast.xml && \
+    $NIPSASM build/"$base".s && \
+    mv build/"$base".b "$base".b
+done

### 2008-05-08 datastructures/Btree_Set.c
Fixed an obscure bug in the Btree implementation.

@@ -218,8 +218,6 @@ static PageEntry *insert_entry( Btree_Set *set,
         COUNT(new_page) = N - (k + 1);
         memcpy(DATA(new_page), DATA(page) + BEGIN(page, k + 1),
                END(page, N - 1) - BEGIN(page, k + 1));
-        memcpy(DATA(new_page) + set->pagesize - ISIZE(N - (k + 1)),
-               DATA(page) - ISIZE(N), 2*sizeof(int)*(N - (k + 1)));
         BEGIN(new_page, 0) = 0;
         for (n = k + 1; n < N; ++n)
             END(new_page, n - (k + 1)) = END(page, n) - BEGIN(page, k + 1);

### 2008-05-08 search/main.c

@@ -7,7 +7,7 @@
 #include "nips_vm/bytecode.h"
 
 static const char *opt_bytecode_path;
-static bool opt_bfs = false;
+static bool opt_dfs;
 static Set *set = NULL;
 
 static void usage()
@@ -61,8 +61,14 @@ static void parse_args(int argc, char *argv[])
         usage();
     }
     else
+    if (bfs || dfs)
     {
-        opt_bfs = bfs;
+        opt_dfs = dfs;
+    }
+    else
+    {
+        /* Default */
+        opt_dfs = false;
     }
 
     if (opt_bytecode_path == NULL)
@@ -116,7 +122,7 @@ int main(int argc, char *argv[])
     }
 
     /* Do search */
-    if (search(bytecode, deque, set, opt_bfs, &expanded, &transitions) < 0)
+    if (search(bytecode, deque, set, opt_dfs, &expanded, &transitions) < 0)
     {
         fprintf(stderr, "State space search failed!\n");
         status = 1;

### 2008-05-08 search/search.c

@@ -158,6 +158,7 @@ static int depth_first_search(SearchContext *sc)
         {
             return -1;
         }
+
         state = duplicate_state(state, state_size);
         if (state == NULL)
         {
@@ -195,9 +196,6 @@ static int breadth_first_search(SearchContext *sc)
     {
         if (!queue->get_front(queue, (void**)&state, &state_size))
             return -1;
-
-        base64_print(stdout, state, state_size);
-        fflush(stdout);
 
         if (!expand_state(sc, state))
             return -1;
@@ -267,6 +265,7 @@ int search( st_bytecode *bytecode,
     }
 
     /* Do bfs/dfs search */
+    printf("%d\n", dfs);
     if (dfs)
         status = depth_first_search(&sc);
     else

### 2008-05-08 search/main.c
Changed state search interface so it is easier to change the benchmark configuration.

@@ -6,9 +6,11 @@
 #include "search.h"
 #include "nips_vm/bytecode.h"
 
-static const char *opt_bytecode_path;
-static bool opt_dfs;
-static Set *set = NULL;
+static const char   *opt_bytecode_path      = NULL;
+static long         opt_max_iterations      = 0;
+static long         opt_report_interval     = 0;
+static bool         opt_dfs                 = false;
+static Set          *set                    = NULL;
 
 static void usage()
 {
@@ -18,7 +20,10 @@ static void usage()
         "Options:\n"
         "    -B          -- use breadth-first search (default)\n"
         "    -D          -- use depth-first search\n"
-        "    -m model    -- path to model bytecode file\n");
+        "    -m model    -- path to model bytecode file\n"
+        "    -l cnt      -- iteration limit\n"
+        "    -i cnt      -- reporting interval\n"
+        );
     exit(1);
 }
 
@@ -29,7 +34,7 @@ static void parse_args(int argc, char *argv[])
 
     if (argc < 2) usage();
 
-    while ((ch = getopt(argc, argv, "BDm:")) >= 0)
+    while ((ch = getopt(argc, argv, "BDm:l:i:")) >= 0)
     {
         switch (ch)
         {
@@ -50,6 +55,24 @@ static void parse_args(int argc, char *argv[])
             opt_bytecode_path = optarg;
             break;
 
+        case 'l':
+            opt_max_iterations = atol(optarg);
+            if (opt_max_iterations <= 0)
+            {
+                printf("Iteration limit must be a positive integer!\n\n");
+                usage();
+            }
+            break;
+
+        case 'i':
+            opt_report_interval = atol(optarg);
+            if (opt_report_interval <= 0)
+            {
+                printf("Reporting interval must be a positive integer!\n\n");
+                usage();
+            }
+            break;
+
         case '?':
             usage();
         }
@@ -64,11 +87,6 @@ static void parse_args(int argc, char *argv[])
     if (bfs || dfs)
     {
         opt_dfs = dfs;
-    }
-    else
-    {
-        /* Default */
-        opt_dfs = false;
     }
 
     if (opt_bytecode_path == NULL)
@@ -95,26 +113,31 @@ static void parse_args(int argc, char *argv[])
 
 int main(int argc, char *argv[])
 {
-    Deque *deque = NULL;
-    st_bytecode *bytecode = NULL;
     int status = 0;
[…]

### 2008-05-08 search/search.c

@@ -10,6 +10,10 @@ typedef struct SearchContext
     Set             *visited;
     long            expanded;
     long            transitions;
+    long            iterations_left;
+    long            report_iterations_left;
+    long            report_interval;
+    FILE            *report_fp;
 
     /* To capture VM errors: */
     int             err_code;
@@ -73,6 +77,24 @@ void base64_print(FILE *fp, void *data, size_t size)
 
     fputc('\n', fp);
 }
+
+void report_header(FILE *fp)
+{
+    fprintf( fp,
+        " expanded   queued   transit. \n");
+    fprintf( fp,
+        "--------- --------- --------- \n");
+}
+
+void report(FILE *fp, SearchContext *sc)
+{
+    fprintf( fp, "%9ld %9ld %9ld\n",
+             sc->expanded,
+            (long)sc->queue->size(sc->queue),
+            sc->transitions );
+
+}
+
 
 /* Makes a copy of the given state with specified size in dynamic memory
    which must be freed by the caller using free(). */
@@ -139,6 +161,21 @@ static bool expand_state(SearchContext *sc, nipsvm_state_t *state)
     if (sc->err_code != -1)
         return false;
 
+    if (sc->iterations_left > 0)
+    {
+        if (--sc->iterations_left == 0)
+            return true;
+    }
+
+    if (sc->report_iterations_left > 0)
+    {
+        if (--sc->report_iterations_left == 0)
+        {
+            report(sc->report_fp, sc);
+            sc->report_iterations_left = sc->report_interval;
+        }
+    }
+
     return true;
 }
 
@@ -151,7 +188,7 @@ static int depth_first_search(SearchContext *sc)
     nipsvm_state_t *state;
     size_t state_size;
 
-    while (!queue->empty(queue))
+    while (!queue->empty(queue) && sc->iterations_left != 0)
     {
         /* Remove state from the queue */
         if (!queue->get_back(queue, (void**)&state, &state_size))
@@ -192,7 +229,7 @@ static int breadth_first_search(SearchContext *sc)
     nipsvm_state_t *state;
     size_t state_size;
 
-    while (!queue->empty(queue))
+    while (!queue->empty(queue) && sc->iterations_left != 0)
     {
         if (!queue->get_front(queue, (void**)&state, &state_size))
             return -1;
@@ -207,9 +244,7 @@ static int breadth_first_search(SearchContext *sc)
     return 0;
 }
 
-int search( st_bytecode *bytecode,
[…]

### 2008-05-08 search/search.h

@@ -2,18 +2,36 @@
 #define SEARCH_H_INCLUDED
 
 #include <stdbool.h>
+#include <stdio.h>
 #include <datastructures/Set.h>
 #include <datastructures/Deque.h>
 #include <nips_vm/bytecode.h>
 
-/* Does a state space search and returns the total number of states expanded,
-   and returns 0, or -1 if an error occurs while executing.
-   Initially,  ``queue'' and ``visited'' should be empty.
+/* A structure describing parameters used for searching.
 
-   If ``dfs'' is set to true, a depth-first search is performed; otherwise,
-   a breadth-first search is performed.
+    bytecode            NIPS VM bytecode of the model to use.
+    queue               Deque instance to use for the BFS queue/DFS stack.
+    visited             Set instance to use to record visited states.
+    dfs                 Use depth-first search instead of breadth-first search.
+    max_iterations      Maximum number of iterations to perform (0: no limit).
+    report_fp           File to write status reports to.
+    report_interval     Number of iterations between reporting.
+
+    In the above, an iteration is a single state expansion.
 */
-int search( st_bytecode *bytecode, Deque *queue, Set *visited, bool dfs,
-            long *expanded, long *transitions );
+struct SearchParams
+{
+    st_bytecode *model;
+    Deque       *queue;
+    Set         *visited;
+    bool        dfs;
+    long        max_iterations;
+    FILE        *report_fp;
+    long        report_interval;
+};
+
+/* Does a state space search and returns 0, or -1 if an error occurs while
+   executing. */
+int search(const struct SearchParams *params);
 
 #endif /* ndef SEARCH_H_INCLUDED */

### 2008-05-08 datastructures/Set.c
Fixed a bug in the set specifiation parsing and in the search framework.

@@ -130,7 +130,7 @@ Set *Set_create_from_args(int argc, const char * const *argv)
                 return NULL;
         }
         else
-        if (path == NULL && sscanf(*argv, "file=%as", &path) == 1)
+        if (path == NULL && sscanf(*argv, "path=%as", &path) == 1)
         {
             if (path == NULL)
                 return NULL;
@@ -165,6 +165,7 @@ Set *Set_create_from_args(int argc, const char * const *argv)
                 return NULL;
             replay = true;
         }
+        else
         if (density == -1 && sscanf(*argv, "density=%lf", &density) == 1)
         {
             if (type != Bender)

### 2008-05-08 search/main.c

@@ -106,7 +106,7 @@ static void parse_args(int argc, char *argv[])
     set = Set_create_from_args(argc, (const char**)argv);
     if (set == NULL)
     {
-        perror("Could not create set (invalid description?)\n\n");
+        printf("Could not create set (invalid description?)\n\n");
         exit(1);
     }
 }

### 2008-05-08 search/search.c

@@ -2,6 +2,7 @@
 #include <assert.h>
 #include <stdio.h>
 #include <nips_vm/nipsvm.h>
+#include <sys/time.h>
 
 typedef struct SearchContext
 {
@@ -20,10 +21,14 @@ typedef struct SearchContext
     nipsvm_pid_t    err_pid;
     nipsvm_pc_t     err_pc;
 
+    /* For reporting stats: */
+    double          time_start;
+
 } SearchContext;
 
+
 /* For debugging */
-void hex_print(FILE *fp, void *data, size_t size)
+static void hex_print(FILE *fp, void *data, size_t size)
 {
     unsigned char *p = data;
     size_t i;
@@ -47,7 +52,7 @@ void hex_print(FILE *fp, void *data, size_t size)
 }
 
 /* For debugging */
-void base64_print(FILE *fp, void *data, size_t size)
+static void base64_print(FILE *fp, void *data, size_t size)
 {
     unsigned char *p = data;
     unsigned int v;
@@ -78,6 +83,16 @@ void base64_print(FILE *fp, void *data, size_t size)
     fputc('\n', fp);
 }
 
+static double now()
+{
+    struct timeval tv;
+    int res;
+
+    res = gettimeofday(&tv, NULL);
+    assert(res == 0);
+    return (double)tv.tv_sec + 1e-6*tv.tv_usec;
+}
+
 void report_header(FILE *fp)
 {
     fprintf( fp,
@@ -94,7 +109,6 @@ void report(FILE *fp, SearchContext *sc)
             sc->transitions );
 
 }
-
 
 /* Makes a copy of the given state with specified size in dynamic memory
    which must be freed by the caller using free(). */
@@ -227,21 +241,42 @@ static int breadth_first_search(SearchContext *sc)
 {
     Deque *queue = sc->queue;
     nipsvm_state_t *state;
-    size_t state_size;
+    size_t state_size, max_state_size = 0;
+    nipsvm_state_t *state_copy = NULL;
+    int status = 0;
 
     while (!queue->empty(queue) && sc->iterations_left != 0)
     {
         if (!queue->get_front(queue, (void**)&state, &state_size))
-            return -1;
+        {
+            status = -1;
+            break;
+        }
 
-        if (!expand_state(sc, state))
-            return -1;
+        if (state_size > max_state_size)
+        {
+            max_state_size = state_size;
+            state_copy = realloc(state_copy, max_state_size);
[…]

### 2008-05-08 datastructures/Deque.h
Attempt to improve BFS performance by avoiding copying states. Unfortunately, the effect on total run time seems negligible.

@@ -51,6 +51,10 @@ Deque *Memory_Deque_create();
         Removes an element at the front of the queue or returns false if the
         deque is empty or the element could not be removed.
 
+    bool reserve(size_t count, size_t size)
+        Tries to reserve space for ``count'' more elements of size ``size''
+        or returns false if the reservation could not be made. This prevents
+        reallocations when growing.
 */
 struct Deque
 {
@@ -63,6 +67,7 @@ struct Deque
     bool (*get_front)(struct Deque *, void **, size_t *);
     bool (*pop_back)(struct Deque *);
     bool (*pop_front)(struct Deque *);
+    bool (*reserve)(struct Deque *, size_t count, size_t size);
 };
 
 #endif /* ndef DEQUE_H_INCLUDED */

### 2008-05-08 datastructures/FileStorage.c

@@ -39,15 +39,20 @@ void FS_destroy(FileStorage *fs, void *data)
 
 void *FS_resize(FileStorage *fs, void *data, size_t size)
 {
+    data = FS_reserve(fs, data, size);
+    if (data != NULL)
+        fs->size = size;
+    return data;
+}
+
+void *FS_reserve(FileStorage *fs, void *data, size_t size)
+{
     int res;
     size_t old_size, new_size;
 
-    /* First, check to see if we have free space left */
+    /* First, check to see if any reallocation is required */
     if (size <= fs->capacity)
-    {
-        fs->size = size;
         return data;
-    }
 
     /* Round desired size up to the next page boundary. */
     old_size = fs->capacity;
@@ -88,8 +93,7 @@ void *FS_resize(FileStorage *fs, void *data, size_t size)
     if (data == NULL || data == MAP_FAILED)
         return NULL;
 
-    /* Update size/capacity */
-    fs->size     = size;
+    /* Update capacity */
     fs->capacity = new_size;
 
     return data;

### 2008-05-08 datastructures/FileStorage.h

@@ -24,10 +24,15 @@ void FS_destroy(FileStorage *fs, void *data);
 
 /* Resizes the used memory size.
    This may cause the underlying file to be extended, in which case ``data''
-   may have tobe moved!
+   may have to be moved!
 
-   Returns true if the request was completed succesfully, or returns false and
-   sets errno. */
+   Returns a new data pointer or returns NULL and sets errno. */
 void *FS_resize(FileStorage *fs, void *data, size_t new_size);
+
+/* Similar to FS_resize, but only changes the available capacity
+   (not the actual size).
+
+   Returns a new data pointer or returns NULL and sets errno. */
+void *FS_reserve(FileStorage *fs, void *data, size_t new_size);
 
 #endif /* ndef FILE_STORAGE_H_INCLUDED */

### 2008-05-08 datastructures/File_Deque.c

@@ -142,6 +142,18 @@ static bool pop_front(FileDeque *deque)
     return true;
 }
 
+static bool reserve(FileDeque *deque, size_t count, size_t size)
+{
+    void *new_data;
+
+    new_data = FS_reserve( &deque->fs, deque->data,
+                           deque->end + count*(2*sizeof(size_t) + size) );
+    if (new_data == NULL)
+        return false;
+    deque->data = new_data;
+    return true;
+}
+
 Deque *File_Deque_create(const char *filepath)
 {
     FileDeque *deque;
@@ -159,6 +171,7 @@ Deque *File_Deque_create(const char *filepath)
     deque->base.get_front  = (void*)get_front;
     deque->base.pop_back   = (void*)pop_back;
     deque->base.pop_front  = (void*)pop_front;
+    deque->base.reserve    = (void*)reserve;
 
     deque->count = 0;
     deque->begin = 0;

### 2008-05-08 datastructures/Memory_Deque.c

@@ -154,6 +154,11 @@ static void destroy(MemDeque *deque)
     free(deque);
 }
 
+static bool reserve(MemDeque *deque, size_t count, size_t size)
+{
+    return false;
+}
+
 Deque *Memory_Deque_create()
 {
     MemDeque *deque;
@@ -171,6 +176,7 @@ Deque *Memory_Deque_create()
     deque->base.get_front  = (void*)get_front;
     deque->base.pop_back   = (void*)pop_back;
     deque->base.pop_front  = (void*)pop_front;
+    deque->base.reserve    = (void*)reserve;
 
     deque->count = 0;
     deque->last  = NULL;

### 2008-05-08 search/search.c

@@ -241,27 +241,25 @@ static int breadth_first_search(SearchContext *sc)
 {
     Deque *queue = sc->queue;
     nipsvm_state_t *state;
-    size_t state_size, max_state_size = 0;
-    nipsvm_state_t *state_copy = NULL;
-    int status = 0;
+    size_t state_size;
+    int status;
 
+    status = 0;
+    state_size = 1;
     while (!queue->empty(queue) && sc->iterations_left != 0)
     {
+        /* HACK: reserve some space in advance, otherwise the pointer to the
+                 current state may be invalidated when new states are added to
+                 the queue when expand_state() is called. */
+        queue->reserve(queue, 100, state_size);
+
         if (!queue->get_front(queue, (void**)&state, &state_size))
         {
             status = -1;
             break;
         }
 
-        if (state_size > max_state_size)
-        {
-            max_state_size = state_size;
-            state_copy = realloc(state_copy, max_state_size);
-            assert(state_copy != NULL);
-        }
-        memcpy(state_copy, state, state_size);
-
-        if (!expand_state(sc, state_copy))
+        if (!expand_state(sc, state))
         {
             status = -1;
             break;
@@ -273,8 +271,6 @@ static int breadth_first_search(SearchContext *sc)
             break;
         }
     }
-
-    free(state_copy);
 
     return status;
 }

### 2008-05-08 search/search.c
Added some more statistics to the search report.

@@ -4,6 +4,8 @@
 #include <nips_vm/nipsvm.h>
 #include <sys/time.h>
 
+const int PAGESIZE = 4096;
+
 typedef struct SearchContext
 {
     nipsvm_t        *vm;
@@ -83,6 +85,11 @@ static void base64_print(FILE *fp, void *data, size_t size)
     fputc('\n', fp);
 }
 
+static double tv_to_sec(struct timeval *tv)
+{
+    return (double)tv->tv_sec + 1e-6*tv->tv_usec;
+}
+
 static double now()
 {
     struct timeval tv;
@@ -90,23 +97,43 @@ static double now()
 
     res = gettimeofday(&tv, NULL);
     assert(res == 0);
-    return (double)tv.tv_sec + 1e-6*tv.tv_usec;
+    return tv_to_sec(&tv);
 }
 
 void report_header(FILE *fp)
 {
     fprintf( fp,
-        " expanded   queued   transit. \n");
+        " expanded   queued   transit. wc.time  u.time  s.time  res.size   virt.size\n");
     fprintf( fp,
-        "--------- --------- --------- \n");
+        "--------- --------- --------- ------- -------  ------ ----------- -----------\n");
 }
 
 void report(FILE *fp, SearchContext *sc)
 {
-    fprintf( fp, "%9ld %9ld %9ld\n",
+    FILE *stat;
+    int res;
+    unsigned long utime, stime, vsize;
+    long rss;
+
+    stat = fopen("/proc/self/stat", "rt");
+    assert(stat != NULL);
+    res = fscanf( stat, "%*d %*s %*c %*d %*d %*d %*d %*d %*u "
+                        "%*lu %*lu %*lu %*lu "
+                        "%lu %lu %*ld %*ld %*ld %*ld %*ld %*ld %*llu "
+                        "%lu %ld",
+                  &utime, &stime, &vsize, &rss );
+    assert(res == 4);
+    fclose(stat);
+
+    fprintf( fp, "%9ld %9ld %9ld %7.3f %7.3f %7.3f %11ld %11lu\n",
              sc->expanded,
-            (long)sc->queue->size(sc->queue),
-            sc->transitions );
+             (long)sc->queue->size(sc->queue),
+             sc->transitions,
+             now() - sc->time_start,
+             (double)utime/100,
+             (double)stime/100,
+             PAGESIZE*rss,
+             vsize );
 
 }
 

### 2008-05-08 paper/paper.tex
Added paper outline.
Made density parameter of Bender set configurable.
Added some notes to the paper (for my own use).
Started on the paper contents.

@@
+% TODO: ensure all references in paper.bib are cited in the paper!
 \documentclass{acm_proc_article-sp}
@@
 \title{Evaluation of Cache-Oblivious Data Structures}
+\subtitle{\textit{Work In Progress!}}
 
@@
 \begin{abstract}
-%TODO
+% TODO: rewrite (NB: max. 10 lines)
+%In modern computer hardware architecture, memory is organized in a hierarchy
+consisting of several types of memory with different memory sizes, block
+transfer sizes and access times. The cache-oblivious model has been proposed
+to reflect this reality more accurately than traditional models. A number of
+algorithms and data structures have been proposed that perform optimally in
+this model. The goal of the proposed research project is to implement a
+selection of cache-oblivious set-like data structures, evaluate their
+performance (in terms of execution time and memory use) in a realistic
+environment, and compare their performance to traditional data structures.
 \end{abstract}
 
+% TODO: update keywords?
 \keywords{cache efficiency, locality of reference, algorithms}
 
+% TODO: more meaningful section titles?
+
 \section{Introduction}
+% Copy and shorten introduction from proposal
+% -> Motivate research into cache-efficient algorithms
+% -> Move "classification" subsection describing three models to front
+% -> Describe each of the three models
+% -> Introduce three data structures:
+%     - note that hash tables have poor locality of reference
+%     - note that btrees have good locality of reference but require pagesize
+% -> Finally, present goal of the research (incorporate "Research Questions" in
+proposal)
 
-%TODO: set up paper structure
-%TODO: borrow as much as possible from the proposal
+% Marielle suggested:
+%  - give the "standard" memory model an explicit name
+%  - move example of cache oblivious data structures to front
+%  - mention examples of cache unaware and cache aware algorithms
+
+% Note explicitly that cache-oblivious algorithms are relatively complex, so
+% that the main question is wether the improvedlocality of reference 
+% compensates for the overhead incurred.
+
+Text here will be mostly copied from the proposal.
+
+\section{Related Work}
+% Present related work
+
+Text here will be mostly copied from the proposal.
+
+\section{Test Case}
+As a test case, the scenario of a state space search has been chosen. This is a
+suitable scenario for two reasons. First, it is commonly used as a practical
+component of formal methods for software verification, and therefore good
+algorithms are of great practical significance. Second, as we will explain
+below, the performance of state space search algorithms depends for a large
+part on the performance of the data structures that are used to implement
+them; therefore, research into efficient data structures is of particular
+interest to this application.
+
+State space search is used to verify the correctness of software programs. For
+this purpose, programs are first modeled using a formal language, that is also
+used to specify properties of the program that should hold during its
+execution. A program can be in a (possibly infinite) amount of states, one of
+which is usually designated the \emph{initial state}. If a transition from one
+state to another is possible (according to the rules of the formal language
+used) the latter state is said to be a \emph{successor state} of the former.
+Generating the successors of a particular state is also called
+\emph{expanding} the state. Of course, the model must be non-deterministic to
+allow more than one successor to exist for a single state.
+
+The set of all states reachable by transitions from the initial state is called
+the \emph{state space} of a model, and it is the goal of a state space search
+algorithm to generate all of these states, in order to check that desired
+properties hold in all of them. For this approach it is often required that
+the state space be finite (for obvious reasons).
+
+The outline of a state space search algorithm is given in Program
+\ref{prog-search}. Note that in addition to the initial state and a function
+to generate successors, a queue and a set data structure are used. The queue
+holds states that have been generated but not yet expanded and is used to
+determine what state to expand next. The set holds all states that have been
+generated so far and is used to prevent a single state from being expanded
+more than once.
+
+% TODO: note that the algorithm uses the queue, successors and set, and that
+the set is the bottleneck.
+
+\begin{program}
+\begin{verbatim}
+Queue queue = new Queue;
+Set generated = new Set;
+queue.insert(initial_state);
+while (!queue.empty()) {
+    State state = queue.extract();
+    for (State s : successors(state)) {
+        if (generated.contains(s) == false) {
+            generated.insert(s);
+            queue.insert(s);
+        }
+    }
+}
+\end{verbatim}
+\caption{Pseudo-code for a simple state search algorithm.}
+\label{prog-search}
+\end{program}
+
+% TODO: different name?
+
+% Describe case: state search, set data structure.
+% Describe (globally) what to measure: execution time
+
+\section{Methodology}
+
+% - NIPS VM
+% - SPIN model checker
+% - Promela-to-NIPS compiler
+
+For our experimental framework, we need a collection of formal models that are
+representative of those typically used for formal verification, and a way to
+execute them. For the first part, we have looked at Spin \cite{spin}, a widely
+used model checking tool. Spin uses a custom modeling language, called
+PROMELA, to specify models and comes with a collection of example models that
+are suitable for our experiment.
+
+For the execution of these models we will use NIPS VM \cite{weber2007evm}, a
+high-performance virtual machine for state space generation that is easily
+embeddable in our framework. Although the NIPS VM only executes bytecode in a
+custom format, a PROMELA compiler is also available to generate the required
+bytecode from the Spin models.\cite{nips} The NIPS VM is prefered to over the
+Spin tools because it was designed to be embedded in other programs and as
+such is more easily integrated in our framework.
+
+\section{Data Structures}
+% TODO: copy motivation for reimplementation from proposal
+% TODO: specify required operations (insert&contains)
+% TODO: note that all data structures require contiguous memory
+%       (except the Bender set)
+% TODO: note use of mmap for automatic storage management
+
+\subsection{Hash Table Implementation}
+Hash tables are widely used as a fast (O($1$)) data structure especially when
+used for in-memory storage. In its simplest form, a hash table consists of an
+index array (the \emph{index}) with a fixed number of slots. A \emph{hash
+function} is used to map key values onto slot numbers. If the slot for a value
+is empty, it may be inserted there. Queries for the existance of an element in
+the hash table similarly see if the queried value is stored at its designated
+slot.
+
+Note that if we insert several different values, some values may map to the
+same slot, which is problematic if each slot can only store one value. There
+are a lot of different ways to resolve this collision problem; we use
+\emph{separate chaining} which means that slots do not store values directly,
+but instead each slot stores a pointer to a singly-linked list of values that
+map to that slot. This particular implementation technique is well-suited to
+the scenario where values may have different sizes (as the slots only need to
+store a fixed-size pointer and not a variable-size value) and maintains
+relatively good performance when the number of values stored exceeds the size
+of the index array \cite{sedgewick1998ac}.
+
+% TODO: picture?
+
+Our hash table implementation uses a fixed size index which must be specified
+when creating the hash table, as this simplifies the implementation
+considerably. The index is stored at the front of the file, after which values
+are simply appended in the order in which they are added. Note that we do not
+support removing elements from the hash table, which means we do not have to
+deal with holes that would otherwise occur in the stored file. For our
+experiments the FNV-1a hash function \cite{noll2004fnv} is used (modulo the
+size of the index) to map values to slots.
+
+\subsection{B-tree Implementation}
+The B-tree data structure was first proposed by Bayer and McCreight
+\cite{bayer1970oam} and is widely implemented and often described in
+textbooks. Our implementation is based on the description by Kifer, Bernstein
+and Lewis \cite{kifer2006dsa}.
+
+B-trees are similar to other search tree data structures. A defining property
+is that they organize data in pages of a fixed size, each containing several
+values, which makes them especially useful for purposes of on-disk storage
+where reading and writing data in a few large blocks is relatively cheap
+compared to accessing several smaller blocks.
+
+% TODO: picture and description
+
+Our does not support storing values larger than the size of a page (which
+requires special handling, usually by assigning special overflow pages to hold
+the portion of the value that does not fit in the page). Deletion of items is
+not supported either.
+
+\section{Cache-Oblivious Dictionary Implementation}
+% TODO: reference (if not included earlier)
+
+% sparse array
+% configurable density parameter
+% binary search tree index
+% van Emde Boas lay-out
+% special optimization on insertion (note that it's not described by Bender)
+
+\subsection{Experimental Environment}
+
+\section{Metrics}
+% Describe how performance was measured
+% Describe data set used
+
+\section{Results}
+% Present results!
+% Tables, graphs, etc.
+% What to include:
+%   per program:     transitions/states graph
+%   per program/set: 
+%
+%
+
+\section{Discussions}
+% Explain results, draw conclusions.
+
+\section{Conclusion}
+% Repeat findings.
 

### 2008-05-09 datastructures/File_Deque.c
Added compaction to File_Deque.

@@ -21,9 +21,10 @@
     Some notes:
     - pushing data in front of the deque is not supported in this
       implementation!)
-    - in order to save space, the allocated data can be moved to the front of
-      the file periodically (e.g. whenever end - begin > begin) to save disk
-      space while maintaining amortized (but not worst-case) complexity bounds.
+    - in order to save space, the allocated data is moved to the front of the
+      file periodically (e.g. whenever end - begin > begin) to save disk
+      space while maintaining amortized (but not per-operation) complexity
+      bounds.
     - 16 bytes per entry is a lot of overhead; using 32-bit sizes would
       reduce that to 8 bytes per entry and is probably sufficient as well (but
       in that case data should probably still be stored on 8-byte boundaries)
@@ -138,6 +139,15 @@ static bool pop_front(FileDeque *deque)
     size = *(size_t*)(deque->data + deque->begin);
     deque->begin += 2*sizeof(size_t) + align(size);
     --deque->count;
+
+    if (deque->begin > deque->end - deque->begin)
+    {
+        /* Compact space */
+        memmove( deque->data,
+                 deque->data + deque->begin, deque->end - deque->begin );
+        deque->end   -= deque->begin;
+        deque->begin -= deque->begin;
+    }
 
     return true;
 }

### 2008-05-09 datastructures/Bender_Impl.c
Improved Bender set implementation of array contents redistribution.
The old implementation would not distribute gaps uniformly over all windows,
which is problematic when insertions are repeatedly made in the same window
(under some additional circumstances), because rebalancing of higher-level
windows would not generate gaps in some of its child windows (which would then
still be overflowing).

Unfortunately, the new implementation appears to be slightly (around 5%)
slower when elements are inserted on random locations. This can probably be
resolved by a faster computation of where gaps are

@@ -541,11 +541,13 @@ static void resize(Bender_Impl *bi, int new_order)
 static void insert_and_redistribute (Bender_Impl *bi,
     int lev, size_t win, size_t idx, size_t N, const void *data, size_t size )
 {
-    size_t begin, end, n, p, q, W, gap_space, gap_extra;
+    size_t begin, end, n, p, q, W;
 
     W = WINDOW_SIZE(lev);
     begin = win*W;
     end   = begin + W;
+    /* printf("Redistributing window %d on level %d/%d (%d/%d elements); inserting at %d\n",
+              (int)win, (int)lev, bi->L, (int)N + 1, (int)W, (int)idx); */
 
     /* We start by packing all the elements to the front of the array.
        If there is a gap at or before ``idx'' this is easy, otherwise
@@ -622,40 +624,27 @@ static void insert_and_redistribute (Bender_Impl *bi,
        "Evenly" means that if the window has size "W" and contains "N"
        elements, then there are (W-N) spaces to be divided over N+1 gaps,
        which means that each gap is at least (W-N)/(N+1) (rounded down)
-       in size, while the last (W-N)%(N+1) gaps have an extra space.
-
-       (This definition if evenly is not specifically required, as long
-        as elements are divided roughly evenly among windows of the lower
-        levels.)
+       in size; there are (W-N)%(N+1) gaps with an extra space, which must
+       be divided uniformly over all windows.
     */
     N += 1;
-    gap_space = (W - N)/(N + 1);
-    gap_extra = (W - N)%(N + 1);
 
     p = end;
-    while (q > begin)
-    {
-        /* Insert gap */
-        for (n = 0; n < gap_space; ++n)
-        {
-            p -= 1;
-            ARRAY_AT(p)->size = (size_t)-1;
-        }
-        if (gap_extra > 0)
-        {
-            p -= 1;
-            ARRAY_AT(p)->size = (size_t)-1;
-            gap_extra -= 1;
-        }
-
-        /* Copy element */
-        p -= 1, q -= 1;
-        ARRAY_COPY(ARRAY_AT(p), ARRAY_AT(q));
-    }
     while (p > begin)
     {
-        p -= 1;
-        ARRAY_AT(p)->size = (size_t)-1;
+        /* FIXME: potential overflow here (though unlikely when using 64-bit
+                  aritmetic); try to implement in a safer way (and if possible,
[…]

### 2008-05-09 paper/paper.tex
Added Makefile for building model bytecode from PROMELA source.

@@
 %In modern computer hardware architecture, memory is organized in a hierarchy
-consisting of several types of memory with different memory sizes, block
-transfer sizes and access times. The cache-oblivious model has been proposed
-to reflect this reality more accurately than traditional models. A number of
-algorithms and data structures have been proposed that perform optimally in
-this model. The goal of the proposed research project is to implement a
-selection of cache-oblivious set-like data structures, evaluate their
-performance (in terms of execution time and memory use) in a realistic
-environment, and compare their performance to traditional data structures.
+%consisting of several types of memory with different memory sizes, block
+%transfer sizes and access times. The cache-oblivious model has been proposed
+%to reflect this reality more accurately than traditional models. A number of
+%algorithms and data structures have been proposed that perform optimally in
+%this model. The goal of the proposed research project is to implement a
+%selection of cache-oblivious set-like data structures, evaluate their
+%performance (in terms of execution time and memory use) in a realistic
+%environment, and compare their performance to traditional data structures.
 \end{abstract}
@@
 % -> Finally, present goal of the research (incorporate "Research Questions" in
-proposal)
+% proposal)
 
@@
 % configurable density parameter
+% note that equal division of gaps over all windows is important
+%   (and note the problems with the "old" behaviour)
 % binary search tree index

### 2008-05-09 search/models/build.sh

@@ -1,8 +1,10 @@
 #!/bin/sh
 
-NIPSHOME=../../nips_vm/
-NIPSASM="${NIPSHOME}"/nips_asm.pl
+NIPS_HOME=../../nips_vm/
+NIPSASM="${NIPS_HOME}"/nips_asm.pl
 CLASSPATH=../../nips_c/lib/nips-promela-compiler.jar
+
+export NIPS_HOME CLASSPATH
 
 mkdir -p build
 for file in $@

### 2008-05-10 datastructures/Btree_Set.c
Made Btree allow a smaller pagesize by reducing the maximum key size from pagesize/4 to pagesize/2.

@@ -16,6 +16,8 @@
 #include <stdbool.h>
 #include <string.h>
 
+#define MIN_VALUES_PER_PAGE   2
+
 /* Page lay-out:
 
     |---- 'pagesize' bytes ---|
@@ -203,7 +205,7 @@ static PageEntry *insert_entry( Btree_Set *set,
         int n, k, new_page;
 
         /* For now, just split in the middle.
-           This works because we require all keys to be less than 1/4th of a
+           This works because we require all keys to be less than 1/2 of a
            page (minus the required index size) so no matter how we split, we
            can always insert the new element in either of the two new pages. */
         k = N/2;
@@ -335,7 +337,8 @@ static bool find_or_insert( Btree_Set *set,
 
 static bool set_insert(Btree_Set *set, const void *key_data, size_t key_size)
 {
-    assert(key_size <= (set->pagesize - ISIZE(4))/4);
+    assert( key_size <= (set->pagesize - ISIZE(MIN_VALUES_PER_PAGE))
+                        / MIN_VALUES_PER_PAGE );
     set->mem_used = 0;
     return find_or_insert(set, key_data, key_size, true);
 }
@@ -353,7 +356,7 @@ Set *Btree_Set_create(Allocator *allocator, int pagesize)
     size_t mem_size;
 
     /* Ensure page size is valid */
-    assert(pagesize > ISIZE(4));
+    assert(pagesize > ISIZE(MIN_VALUES_PER_PAGE));
     assert(pagesize%sizeof(int) == 0);
     assert(pagesize%sizeof(int) == 0);
 
@@ -364,7 +367,7 @@ Set *Btree_Set_create(Allocator *allocator, int pagesize)
 
     /* Temporary memory is needed to copy entries, so an upper bound is:
        H*pagesize/4, where H is the maximum height of the B-tree. */
-    mem_size = 32*pagesize/4;
+    mem_size = 64*pagesize/4;
     mem = malloc(mem_size);
     if (mem == NULL)
     {

### 2008-05-10 search/search.c

@@ -103,9 +103,9 @@ static double now()
 void report_header(FILE *fp)
 {
     fprintf( fp,
-        " expanded   queued   transit. wc.time  u.time  s.time  res.size   virt.size\n");
+        "#expanded   queued   transit. wc.time  u.time  s.time  res.size   virt.size\n");
     fprintf( fp,
-        "--------- --------- --------- ------- -------  ------ ----------- -----------\n");
+        "#-------- --------- --------- ------- -------  ------ ----------- -----------\n");
 }
 
 void report(FILE *fp, SearchContext *sc)

### 2008-05-10 paper/paper.tex
Added some initial test configurations.

@@
 %  - give the "standard" memory model an explicit name
-%  - move example of cache oblivious data structures to front
+%    (isn't "cache-unaware" enough?)
+%  - move examples of cache oblivious data structures to front
 %  - mention examples of cache unaware and cache aware algorithms
+%    (at least Btree, tries and hash tables are mentioned)
 
-% Note explicitly that cache-oblivious algorithms are relatively complex, so
-% that the main question is wether the improvedlocality of reference 
-% compensates for the overhead incurred.
+% Note explicitly that cache-oblivious algorithms are relatively complex
+% resulting in higher constant consts, so that the main question is wether
+% the improved locality of reference compensates for the overhead incurred.
 
@@
 used to specify properties of the program that should hold during its
-execution. A program can be in a (possibly infinite) amount of states, one of
-which is usually designated the \emph{initial state}. If a transition from one
-state to another is possible (according to the rules of the formal language
-used) the latter state is said to be a \emph{successor state} of the former.
+execution. An executing program can be in a (possibly infinite) amount of states,
+one of which is usually designated the \emph{initial state}. If a transition
+from one state to another is possible (according to the rules of the formal
+language used) the latter state is said to be a \emph{successor state} of the
+former.
 Generating the successors of a particular state is also called
-\emph{expanding} the state. Of course, the model must be non-deterministic to
-allow more than one successor to exist for a single state.
+\emph{expanding} the state. Of course, the model must include some form of
+non-determinism to allow more than one successor to exist for a single state.
+In practice, this non-determinism usually comes from processes that execute in
+parallel, where the precise interleaving of the (observed) execution of
+instructions in these processes is non-deterministic, unless synchronization
+primitives (such as channels, semaphores, atomic execution blocks, et cetera)
+are used to enforce a particular ordering.
 
@@
 the state space be finite (for obvious reasons).
-
-The outline of a state space search algorithm is given in Program
-\ref{prog-search}. Note that in addition to the initial state and a function
-to generate successors, a queue and a set data structure are used. The queue
-holds states that have been generated but not yet expanded and is used to
-determine what state to expand next. The set holds all states that have been
-generated so far and is used to prevent a single state from being expanded
-more than once.
-
-% TODO: note that the algorithm uses the queue, successors and set, and that
-the set is the bottleneck.
 
@@
 
-% TODO: different name?
+The outline of a state space search algorithm is given in Program
+\ref{prog-search}. Note that in addition to the initial state and a function
+to generate successors, a queue and a set data structure are used. The queue
+holds states that have been generated but not yet expanded and is used to
+determine what state to expand next. The set holds all states that have been
+generated so far and is used to prevent a single state from being expanded
+more than once.
 
-% Describe case: state search, set data structure.
-% Describe (globally) what to measure: execution time
+Depending on how states are inserted and extracted from the queue, the search
+may be breadth-first (meaning that all states that are reachable in $N$ steps
+from the initial state are expanded before any states that require more than
+$N$ steps) or depth-first (meaning that states that are generated most recently
+by the successor function are expanded first). In the first case, the queue
+operates by a first-in first-out principle, while in the second case the queue
+works in a last-in first-out (stack) principle. In general, breadth-first
+search requires a much larger queue capacity than depth-first search.
+
+From the pseudo-code it is clear that the state space search algorithm does not
+accomplish much by itself; instead, the real work is done by the successor
+function and the queue and the set datastructures. Queues can easily be
+implemented efficiently (adding and removing states takes O(1) time).
+The efficiency of the successor function depends on the execution model used,
+but is typically linear in the number of states produced. In practice,
+therefore, the bottleneck of the algorithm is the set data structure.
 
@@
 
+% Describe (globally) what to measure: execution time
 % - NIPS VM
@@
 such is more easily integrated in our framework.
+
+The NIPS VM represents program state as a binary string; the size of the state
+depends on the number of active processes in the program, and is typically in
+the order of a few hundred bytes.
 
@@
 % Describe how performance was measured
-% Describe data set used
+%  (mock set implementation; times/sizes subtracted)
+
+% Describe data set used; per model:
+%   average state size
+%   total size
+%   ratio of insertions/queries
+
+% Note that state size depends on number of processes
+
+The following models are used for benchmarking:
+\begin{itemize}
+\item\verb#eratosthenes# is a parallel implementation of the sieve
+of Eratosthenes which is used to find prime numbers upto a given maximum.
+It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
+tested. A new process is created for every prime number found, and as a result
+the state space can only by exhaustively searched for relatively small values
+of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
+also varies during execution.
+
+\item\verb#leader2# is a leader election algorithm with a configurable number
+of processes competing for leadership ($N$). Since a constant number of
+processes will be created and these processes cannot make progress until all
+of them are created, almost all states have the same size.
+
+\item\verb#peterson_N# is a Peterson's solution to the multi-process mutual
+exclusion problem (using global variables instead of channels for
+communication) with a configurable number of processes ($N$). The processes
+are created atomically, so the state size is constant after initialization.
+
+% mobile1
+% pftp
+% snoopy
+% sort
+
+% TODO: table of models with parameters,
+
+\end{itemize}
 

### 2008-05-10 search/search.c

@@ -103,7 +103,7 @@ static double now()
 void report_header(FILE *fp)
 {
     fprintf( fp,
-        "#expanded   queued   transit. wc.time  u.time  s.time  res.size   virt.size\n");
+        "#expanded   queued   transit. wc.time  u.time  s.time  res.size    virt.size\n");
     fprintf( fp,
         "#-------- --------- --------- ------- -------  ------ ----------- -----------\n");
 }

### 2008-05-10 search/tests/run-test.sh

@@ -1,0 +1,27 @@
+#!/bin/sh
+
+MODELS=../models/
+
+[ -z "$1" ] && echo 'Missing argument: config' && exit 1
+
+# External file declares:
+MODEL=                                  # base name of model to use
+SET=                                    # set configuration 
+MAXVSZ=unlimited                        # maximum virtual set size
+INTERVAL=1000                           # reporting interval
+MAXIT=10000000                          # maximum number of iterations
+OUTPUT="results/`basename "$1"`"        # output file
+
+# Read from config
+source $1
+[ -z "$MODEL" ] && echo 'No model given' && exit 1
+[ -z "$SET" ] && echo 'No set configuration given' && exit 1
+[ -z "$OUTPUT" ] && echo 'No ouput file given' && exit 1
+[ ! -e "${MODELS}"/"$MODEL".b ] && echo 'Model bytecode file does not exist' \
+                                && exit 1
+
+mkdir -p "`basename "$OUTPUT"`"
+( ulimit -v $MAXVSZ &&
+  ../search -l "$MAXIT" -i "$INTERVAL" -m "${MODELS}"/"$MODEL".b $SET \
+  > "$OUTPUT" )
+

### 2008-05-23 paper/paper.tex
Some updates to the paper.

@@
 % TODO: ensure all references in paper.bib are cited in the paper!
+% TODO: format checklist (see e-mail)
 \documentclass{acm_proc_article-sp}
@@
 \section{Introduction}
+A fundamental part of theoretical computer science is the study of algorithms (formal descriptions of how computations may be performed) and data structures (descriptions of how information is organized and stored in computers).
+Traditionally, algorithms have been evaluated in a simplified model of computation.
+In this model it is assumed that a computer executes an algorithm in discrete steps.
+At each step it performs one elementary operation (e.g.
+comparing two numbers, adding one to another, storing a value in memory, et cetera).
+Each elementary operation is performed within a constant time.
+In this model, both storing and retrieving data values in memory is considered to be an elementary operation.
+
+This model is close enough to the way computers work to be extremely useful in the development and analysis of data structures and algorithms that work well in practice.
+However, like every model, it is a simplification of reality.
+One of the simplifications is the assumption that storing and retrieving data in memory is a constant-time operation (i.e.
+the uniform memory model).
+Advancements in software and hardware design over the last two de\-cades have caused this assumption to be increasingly detached from reality.
+
+One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (fast\-er) main memory when the latter is in short supply.
+Another reason is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
+Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem, but this increases the differences in access time between different types of memory.
+
+As a result, a modern computer system lacks a central memory storage with uniform performance characteristics.
+Instead, it employs a hierarchy of memory storage types.
+Figure \ref{fig-memhier} gives a typical example of such a hierarchy.
+The processor can directly manipulate data contained in its registers only.
+To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy.
+Memory is typically transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
+
+\begin{figure}
+\centering
+\includegraphics{memhier}
+\caption{Schematic depiction of the memory hierarchy}\label{fig-memhier}
+\end{figure}
+
+In the memory hierarchy, every next level of storage is both significantly slower and significantly larger than the one above it and with increasing memory sizes, the block size increases as well.
+Table \ref{tab-memhier} gives an overview of typical memory sizes, block sizes, and access times.
+The values for the (general purpose) registers, L1 cache and L2 cache are those for the Pentium M processor as given in \cite{intel-opt}; other values are approximate.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{ l l l l }
+\hline
+\textbf{Type} & \textbf{Total Size} & \textbf{Block Size} & \textbf{Access Time} \\
+\hline
+Registers  &   32 bytes & 4 bytes & 1 cycle \\
+L1 Cache   &   32 KB    & 64 bytes & 3 cyles \\
+L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
+RAM        &   ~2 GB    & 1 KB     & 50-100 cycles \\
+Disk       & ~300 GB    & 4 KB     & 5,000-10,000 cycles \\
+\hline
+\end{tabular}
+\caption{Sizes and access times of various types of storage}
+\label{tab-memhier}
+\end{center}
+\end{table}
+
+To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed.
+Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account.
+This has prompted research into new memory models that are more realistic.
+We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
+\begin{list}{}{}
+\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existance of a memory hierarchy, and may therefore not perform optimally in practice.
+\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (,described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
+\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (which will also be described below) which does not allow parametrization with properties of the cache layout.
+\end{list}
+
+Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice.
+Altough some research has been done for cache-oblivious data structures, they are seldom used in practice.
+The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.
+
+\subsection{External Memory Model}
+One of the earliest models to take the differences in memory access cost into account is the external memory model, described by Aggarwal and Vitter \cite{aggarwal1988ioc}.
+They make a distinction between internal memory (which is limited in size) and external memory (which is virtually unlimited).
+The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories; additionally, consecutive blocks can be transfered at reduced cost (so-called bulk transfer).
+Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy.
+In that case, the possibility of bulk transfer may have to be dropped.
+
+We will call algorithms that are designed to minimize the number of transfers in a two-level memory hierarchy ``cache-aware'' (as opposed to traditional ``cache-unaware'' algorithms) or ``cache-conscious''.
+These algorithms typically rely on knowledge of the block size to achieve optimal performance.
+
+\subsection{Hierarchical Memory Model}
+The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice the memory hierarchy contains more than just two levels.
+Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
+Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
+This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where, which is usually not the case in practice, making their model of little use in the design and evaluation of practical algorithms.
+
+\subsection{Cache-Oblivious Memory Model}
+A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model, which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
+Data is transferred between the two in aligned data blocks of size B.
+In contrast with the hierarchical memory model, the application does not have explicit control over the transfering of blocks between the two memories.
+Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program.
+Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and datastructures.
+Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
+
+Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration.
+The key advantage of this class of algorithms is that they are implicitly tuned to all levels in the memory hierarchy at once.
+It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.
+
+Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms \cite{demaine2002coa}.
+He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.
+
 % Copy and shorten introduction from proposal
@@
 
-Text here will be mostly copied from the proposal.
+\section{Contributions}
+Several cache-oblivious data structures and algorithms have been proposed.
+Complexity analysis shows that the proposed solutions are asymptotically optimal.
+However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms.
+Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often prefered that are easier to implement and require less memory).
+
+This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice.
+To determine if cache-oblivious data structures have practical merit, empirical performance data is required, which is scarce as existing research has focused mainly on theoretical analysis.
+This paper addresses the question by reporting on the performance of newly implemented (but previously described) cache-oblivious data structures and their traditional counterparts (both cache-aware and cache-unaware data structures).
 
 \section{Related Work}
-% Present related work
+Many data structures and algorithms have been analyzed in the cache-oblivious model.
+Several new data structures and algorithms have been developed that perform optimally in this mo\-del as well.
+Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.
 
-Text here will be mostly copied from the proposal.
+Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
+Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for finger searches.
+The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout (named for similarity to a different data structure proposed by van Emde Boas et al.
+\cite{vanemdeboas1976dai}).
 
-\section{Test Case}
-As a test case, the scenario of a state space search has been chosen. This is a
+Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list.
+This means buckets have to be resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).
+
+Vitter presents a theoretical survey of algorithms evaluated in a parallel disk model \cite{vitter2001ema}, which is a refinement of the external memory model described by Aggarwal and Vitter, but only allows parallel transfer of multiple blocks from different disks, which is more realistic.
+Unfortunately, his survey lacks empirical results.
+
+Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-obli\-vious priority deque.
+Their main result is that although the cache-obli\-vious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
+
+\section{Methods}
+In order to gather emperical data, the research approach must be made more concrete.
+We will need to limit ourselves to a specific class of data structures, since algorithms and data structures offering different functionality cannot be compared in a meaningful way.
+Furthermore, we will need to select a proper scenario in which the data structures are evaluated, as conclusions on the practical merit of the data structures depend on the degree to which the test scenario is realistic.
+
+Furthermore, we will need to define what practical performance is.
+We will measure two properties: primarily, the total running time of our test application, and secondarily the memory used.
+The rationale for selecting these metrics is that if the data structures perform identical functions and plenty of memory is available, the only observable difference in running a program using different data structures will be the run time.
+Memory use is of secondary interest because in practice memory may be limited, which would preclude the use of certain data structures that require a lot of memory to function.
+
+\subsection{State Space Search}
+As the test application, that of a state space search has been chosen. This is a
 suitable scenario for two reasons. First, it is commonly used as a practical
@@
 
-\section{Methodology}
+\subsection{Experimental Framework}
 
@@
 
-\section{Data Structures}
+\subsection{Data Structures}
 % TODO: copy motivation for reimplementation from proposal
@@
 
-\subsection{Hash Table Implementation}
+\subsubsection{Hash Table Implementation}
 Hash tables are widely used as a fast (O($1$)) data structure especially when
@@
 
-\subsection{B-tree Implementation}
+\subsubsection{B-tree Implementation}
 The B-tree data structure was first proposed by Bayer and McCreight
@@
 
+B-trees work by keeping as much values as possible, in order, in a single
+page. Pages are ordered in a tree structure, where interior nodes store
+pointers to child nodes. Pointers originate at a pair of adjacent values,
+and point to a subtree containing only values that fall in between those
+values.
+
+When a new value does not fit in the page where it is to be inserted,
+the page is split into two pages; the median value from the old page is
+moved to the parent page, and two new pages (containing the values less
+than respectively greater than the median) are created. When the top-most
+page needs to be split, a new (empty) top-level page is created and the height
+of the B-tree is increased by one.
+
+B-trees are well suited to storing variable length values, as pages can
+store many values as long as the combined size of the values and pointers
+does not exceed the pagesize.
+
 % TODO: picture and description
 
-Our does not support storing values larger than the size of a page (which
-requires special handling, usually by assigning special overflow pages to hold
-the portion of the value that does not fit in the page). Deletion of items is
-not supported either.
+Note that there is no obvious way to store values that are larger than the
+size of a single page. Production-quality implementations handle this case
+separately, but our implementation does not support this scenario and
+requires stored values to be smaller than the pagesize. Additionally,
+deletion of items has not been implemented.
 
-\section{Cache-Oblivious Dictionary Implementation}
+\subsubsection{Cache-Oblivious Dictionary Implementation}
 % TODO: reference (if not included earlier)
@@
 
-\subsection{Experimental Environment}
-
-\section{Metrics}
+\subsection{Measurements}
 % Describe how performance was measured
@@
 
-\section{Discussions}
+\section{Discussion}
 % Explain results, draw conclusions.

### 2008-05-25 paper/paper.tex
Added abstract and keywords to paper.

@@
 \begin{abstract}
-% TODO: rewrite (NB: max. 10 lines)
-%In modern computer hardware architecture, memory is organized in a hierarchy
-%consisting of several types of memory with different memory sizes, block
-%transfer sizes and access times. The cache-oblivious model has been proposed
-%to reflect this reality more accurately than traditional models. A number of
-%algorithms and data structures have been proposed that perform optimally in
-%this model. The goal of the proposed research project is to implement a
-%selection of cache-oblivious set-like data structures, evaluate their
-%performance (in terms of execution time and memory use) in a realistic
-%environment, and compare their performance to traditional data structures.
+In modern computer hardware architecture, memory is organized in a hierarchy
+consisting of several types of memory with different memory sizes, block
+transfer sizes and access times. To reflect this situation more accurately
+than tradition models, the cache-oblivious model has been proposed, and several
+algorithms and data structures have been proposed that perform optimally in
+this model.
+This paper presents the results of an emperical evaluation of various data
+structures in a realistic scenario, and gives insight into the applicability
+of cache-oblivious data structure in practice.
 \end{abstract}

### 2008-05-25 paper/paper.tex
Updates to paper (and added an image).

@@
 this model.
-This paper presents the results of an emperical evaluation of various data
+This paper presents the results of an empirical evaluation of various data
 structures in a realistic scenario, and gives insight into the applicability
@@
 % TODO: update keywords?
-\keywords{cache efficiency, locality of reference, algorithms}
+\keywords{cache efficiency, locality of reference, algorithms, empirical evaluation}
 
@@
 Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
+This data structure is mostly of theoretic interest, as it is fairly complex to implement and incurs a relatively large overhead in practice.
+
 Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for finger searches.
-The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static search tree in van Emde Boas layout (named for similarity to a different data structure proposed by van Emde Boas et al.
+The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static tree in van Emde Boas layout (named for similarity to a different data structure proposed by van Emde Boas et al.
 \cite{vanemdeboas1976dai}).
+
+Several data structures were proposed by Rahman, Cole and Raman \cite{rahman2001opd} amongst them a cache-oblivious exponential search tree with a similar structure as the static search tree proposed by Prokop.
+In their experimental results the cache-oblivious tree performs worse than the alternatives.
+Nevertheless, they conclude that \textit{"cache-obli\-vious data structures may have significant practical importance"}.
 
@@
 \section{Methods}
-In order to gather emperical data, the research approach must be made more concrete.
+In order to gather empirical data, the research approach must be made more concrete.
 We will need to limit ourselves to a specific class of data structures, since algorithms and data structures offering different functionality cannot be compared in a meaningful way.
@@
 custom format, a PROMELA compiler is also available to generate the required
-bytecode from the Spin models.\cite{nips} The NIPS VM is prefered to over the
+bytecode from the Spin models.\cite{nipsvm} The NIPS VM is prefered to over the
 Spin tools because it was designed to be embedded in other programs and as
@@
 \subsection{Data Structures}
-% TODO: copy motivation for reimplementation from proposal
-% TODO: specify required operations (insert&contains)
-% TODO: note that all data structures require contiguous memory
-%       (except the Bender set)
-% TODO: note use of mmap for automatic storage management
+In the introduction we identified three classes of data structures.
+For our evaluation we have implemented a single representative data structure
+for each class:
+\begin{itemize}
+\item \textbf{Cache-unaware}: hash tables.
+Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
+\item \textbf{Cache-aware}: B-trees.
+The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
+\item \textbf{Cache-oblivious}: the B-tree-like data structure proposed by Bender, Duan, Iacono and Wu.
+Since it provides functionality comparable to that of a normal B-tree, this seems like a fair candidate for a comparison.
+For brevity, this data structure will be refered to as a \emph{Bender set}.
+\end{itemize}
+
+Cache-oblivious data structures are not yet commonly used in practice and to
+our knowledge there are no high-quality implementations publicly available.
+The Bender set will therefore have to be implemented from scratch.
+
+In contrast, both hash tables and B-trees are widely used and there are several
+high quality implementations available as software libraries. It is, however,
+undesirable to use existing libraries for a fair comparison, for two reasons.
+
+First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations.
+Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not.
+This may give the existing libraries an unfair advantage.
+
+In an attempt to reduce the bias caused by differences in functionality and
+quality between existing and newly developed libraries, all data structures
+used in the performance evaluation have been implemented from scratch.
+
+\subsubsection{Set Operations}
+Dynamic set data structures may support various operations, such as adding and removing elements, testing if a value is an element of the set, finding elements nearby, succeeding or preceding a given value, counting elements in a range, et cetera.
+However, our state space search algorithm only needs two operations: insertion of new elements (in a set that is initially empty) and testing for the existence of elements.
+In fact, these two operations can be combined into a single operation.
+We call this operation \verb#insert(S,x)#.
+If an element $x$ does not exist in $S$, then $insert(S,x)$ inserts it and returns $0$.
+If $x$ is already an element of $S$, $insert(S,x)$ returns $1$.
+The inner loop of our state space algorithm can now be written as follows:
+\begin{verbatim}
+    for (State s : successors(state)) {
+        if (generated.insert(s) == 0) {
+            queue.insert(s);
+        }
+    }
+\end{verbatim}
+
+We now have a single operation that must be implemented by all set data structures.
+Note that the values that need to be stored in the state space search algorithm are virtual machine state descriptions, which (in the NIPS VM) are opaque variable-length strings.
+All data structures must therefore support storing strings of varying length efficiently.
+
+\subsubsection{Common Implementation Details}
+For memory management, the data structures make use of memory mapped files. The
+B-tree and hash table implementations use a single file for all data storage;
+the Bender tree uses several files (as discussed below) and some main memory.
+
+The use of memory mapped files means that the operating system is in charge of
+when and which pages are transfered between main memory and disk storage; the
+application assumes no control over this policy. This is in accordance with the
+cache-oblivious model, that assumes no control over the cache is possible.
 
@@
 
-\subsubsection{Cache-Oblivious Dictionary Implementation}
-% TODO: reference (if not included earlier)
+\subsubsection{Bender Set Implementation}
+The first implementation challenge for the Bender set is that the description
+given by Bender et al. assumes that all values stored in the set are of a
+fixed size, which is not the case in our experimental framework. To work
+around this, we create several separate instances of the Bender set with
+different value sizes which are powers of two. When a value is to be inserted,
+its size is rounded up to a power of two and it is inserted in the
+corresponding set instance. This ensures that the amount of spaces wasted
+remains below 50\% while still allowing values of various sizes to be inserted.
 
-% sparse array
-% configurable density parameter
-% note that equal division of gaps over all windows is important
-%   (and note the problems with the "old" behaviour)
-% binary search tree index
-% van Emde Boas lay-out
-% special optimization on insertion (note that it's not described by Bender)
+The implementation of the Bender set consists of two parts: a sparse array
+storing the values in-order, and a binary tree stored in van Emde Boas
+layout (similar to the one proposed by Prokop). Both the tree and the array
+may contain special \emph{empty values}. A depiction of a Bender set storing
+four values (with a capacity for eight) is given in Figure \ref{fig-bender-set}.
+
+\begin{figure}
+\centering
+\includegraphics{bender-set}
+\caption{A depiction of a Bender set}\label{fig-bender-set}
+\end{figure}
+
+A Bender set is created with a capacity $C$ that is a power of two.
+Initially, the array stores only empty values and is partitioned into
+\emph{windows} on several \emph{levels}.
+On the highest level, there is a single window of size $C$. Each subsequent level
+has twice as many windows of half that size, and the lowest level has size $log_2 C$
+(rounded up to the next power of 2).
+For each level, a maximum density is given, with the lowest level having density 1,
+the highest level having density $\delta$ (these densities are interpolated linearly
+to obtain the density for the intermediary levels).
+
+The top-level density must be a number between $\frac{1}{2}$ and $1$; the optimal value
+depends on practical circumstances such as available memory and the relative cost of
+rebuilding the data structures. Whenever the number of elements stored in
+a window (the window's \emph{population count}) exceeds the size of the window
+multiplied by its maximum density, the window is said to be \emph{overflowing}.
+Overflows are resolved when a higher-level window is \emph{rebalanced},
+meaning that the values in the window are redistributed evenly over the window,
+which will cause some of the values to be moved out from the lower-level window
+which is too full.
+
+Now, to insert a value, the binary tree is used to find a position in the array
+where the value may be inserted (just before its successor). Then, the windows
+overlapping this position are considered from bottom to top, and the first
+window that can support another element without overflowing is selected, and
+rebalanced (thereby resolving the overflow in all lower level windows).
+
+In the case that all windows overlapping the goal position are overflowing, the
+capacity of the data structure is doubled, a new top level is created and then
+the entire array is rebalanced. This is how the set grows when the number of
+elements stored comes too close to the storage capacity.
+
+According to this description (which follows the paper by Bender et al) a window
+is rebalanced every time an element is inserted. As an optimization, our
+implementation does not always rebalance a window. When there is free space
+between the successor and predecessor of the value to be inserted, the value
+is simply inserted in the middle. In our experiments, this optimization
+increased performance somewhat.
+
+One thing that is not specified in the paper by Bender et al, is how the
+population count for windows is kept. One simply option is to keep no such
+information and simply scan part of the array whenever a population count is
+required. Another extreme is to keep population counts for all windows on all
+levels, and update these counts whenever values are inserted or moved. In
+our experiments a compromise seemed to work best: keep population counts only
+for the lowest-level windows, and recompute the counts for higher-level windows
+when required. This saves a lot of recomputation, while keeping the additional
+costs of updating low.
+
+Finally, the Bender set uses a complete binary tree as an index data structure
+to efficiently find the successor element for a given value. The tree is
+stored in memory in van Emde Boas lay-out (as described by Prokop) to allow
+efficient cache-oblivious traversal.
+
+In the tree, the leaf nodes (when numbered from left to right) store the same values
+as the sparse array (including empty values).
+Interior nodes store the maximum value of their two children (or an empty value,
+if both children store an empty value).
+This tree can be used in a similar way as a binary search tree: by examining the
+two children of a node, we can determine wether a searched-for value lies in the
+left or right subtree.
+
+Note that the \emph{structure} of the tree is static and only changes when the
+capacity of the set is increased (in this case, the tree is recreated from scratch).
+Of course, the values stored in the tree have to be updated when the corresponding
+elements of the array change.
 

### 2008-05-25 search/tests/run-test.sh
-

@@ -7,9 +7,9 @@ MODELS=../models/
 # External file declares:
 MODEL=                                  # base name of model to use
 SET=                                    # set configuration 
-REPEAT=5                                # number of iterations
+REPEAT=3                                # number of iterations
 MAXVSZ=unlimited                        # maximum virtual set size
-INTERVAL=1000                           # reporting interval
+INTERVAL=10000                          # reporting interval
 MAXIT=10000000                          # maximum number of iterations
 OUTPUT="results/`basename "$1"`"        # output file
 

### 2008-05-25 search/tests/aggr-results.sh
Added and updated a bunch of scripts.

@@ -1,0 +1,21 @@
+#!/bin/sh
+
+mkdir -p aggregated
+
+aggregate() {
+    program=$1
+    column=$2
+    base=$3
+    shift 3
+    base_path=aggregated/"$program"-"$base"_"$column"
+    ./aggregate.py "$column" results/"$program"-"$base"-* > "$base_path"
+    for ext in $@
+    do
+        ./aggregate.py "$column" results/"$program"-"$ext"-* > /tmp/aggregated.txt
+        ./subtract.py "$base_path" /tmp/aggregated.txt > aggregated/"$program"-"$ext"_"$column"
+    done
+    rm /tmp/aggregated.txt
+}
+
+aggregate eratosthenes vss    base 01 02 03 04 05 06 07 08 09
+aggregate eratosthenes wctime base 01 02 03 04 05 06 07 08 09

### 2008-05-25 search/tests/aggregate.py

@@ -11,7 +11,7 @@
 # median, minimum and maximum values of the data files are reported.
 #
 
-COLS = [ 'its', 'qsz', 'trans', 'wctime', 'utime', 'stime', 'rss', 'vsz' ]
+COLS = [ 'its', 'qsz', 'trans', 'wctime', 'utime', 'stime', 'rss', 'vss' ]
 
 from sys import argv, exit
 

### 2008-05-25 search/tests/aggregate.py
Added scaling of input to aggregation script (not very nice, but it works). Added/updated some plots.

@@ -10,8 +10,12 @@
 # The first value on each line is interpreted as a key; for each key, the
 # median, minimum and maximum values of the data files are reported.
 #
+# HACK:
+# Now also scales values (not really the proper place to that, but ok).
+#
 
-COLS = [ 'its', 'qsz', 'trans', 'wctime', 'utime', 'stime', 'rss', 'vss' ]
+COLS  = [ 'its', 'qsz', 'trans', 'wctime', 'utime', 'stime',    'rss',    'vss' ]
+divby = [  1000,  1000,    1000,        1,       1,       1,  1048576, 1048576  ]
 
 from sys import argv, exit
 
@@ -53,10 +57,11 @@ for path in argv[2:]:
         val = row[col]
 
         # Interpret value
-        if val.find('.') < 0:
-            val = int(row[col])
-        else:
-            val = float(row[col])
+        val = float(row[col])
+        #if val.find('.') < 0:
+        #    val = int(row[col])
+        #else:
+        #    val = float(row[col])
 
         # Store result
         if key not in values:
@@ -76,4 +81,4 @@ for key, vals in items:
     else:
         medi = vals[len(vals)/2]
 
-    print "%19s %19s %19s %19s" % (str(key), str(medi), str(mini), str(maxi))
+    print "%19s %19s %19s %19s" % (str(key/divby[0]), str(medi/divby[col]), str(mini/divby[col]), str(maxi/divby[col]))

### 2008-05-25 paper/paper.tex
-
Added and updated a bunch of scripts.
Added scaling of input to aggregation script (not very nice, but it works). Added/updated some plots.
Added some more figures to the paper (these will change later).

@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{ l l l l }
+\hline
+\textbf{Case} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
+\hline
+eratosthenes & $N=40$ & 1,019,960 & 4,923,218 \\
+leader2 & $N=5$ & 5,950,945 & 23,856,363 \\
+peterson\_N & \\
+\hline
+\end{tabular}
+\caption{Properties of test cases used}
+\label{tab-cases}
+\end{center}
+\end{table}
+
 \end{itemize}
@@
 \section{Results}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
+\caption{Eratosthenes/Bender set execution time}
+\label{fig-eratosthenes-bender_wctime}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
+\caption{Eratosthenes/Bender set memory usage}
+\label{fig-eratosthenes-bender_vss}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
+\caption{Eratosthenes/hash table execution time}
+\label{fig-eratosthenes-hash_wctime}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
+\caption{Eratosthenes/hash table memory usage}
+\label{fig-eratosthenes-hash_vss}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
+\caption{Eratosthenes/B-tree execution time}
+\label{fig-eratosthenes-btree}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
+\caption{Eratosthenes/B-tree memory usage}
+\label{fig-eratosthenes-btree}
+\end{figure}
+
 % Present results!
@@
 % What to include:
-%   per program:     transitions/states graph
-%   per program/set: 
+%   queue size over iterations
+%       for each program
 %
+%   insertions per query:
+%       for each program
+%
+%   per program:
+%       wctime over iterations
+%          for each set
+%       memory over iterations
+%          for each set
 %
@@
 
+\include{fig-queuesize}
+
 \bibliographystyle{plain}

### 2008-05-25 search/tests/run-test.sh
Added VSS limit to problematic test case.

@@ -8,7 +8,7 @@ MODELS=../models/
 MODEL=                                  # base name of model to use
 SET=                                    # set configuration 
 REPEAT=3                                # number of iterations
-MAXVSZ=unlimited                        # maximum virtual set size
+MAXVSS=unlimited                        # maximum virtual set size
 INTERVAL=10000                          # reporting interval
 MAXIT=10000000                          # maximum number of iterations
 OUTPUT="results/`basename "$1"`"        # output file
@@ -34,7 +34,7 @@ do
   fi
 
   # Run test
-  ( ulimit -v $MAXVSZ &&
+  ( ulimit -v $MAXVSS &&
     ../search -l "$MAXIT" -i "$INTERVAL" -m "${MODELS}"/"$MODEL".b $SET \
     > "$out" )
 

### 2008-05-25 search/tests/aggr-results.sh
Added new plots.

@@ -19,3 +19,7 @@ aggregate() {
 
 aggregate eratosthenes vss    base 01 02 03 04 05 06 07 08 09
 aggregate eratosthenes wctime base 01 02 03 04 05 06 07 08 09
+aggregate leader2      vss    base 01 02 03
+aggregate leader2      wctime base 01 02 03
+aggregate peterson     vss    base 01 02 03
+aggregate peterson     wctime base 01 02 03

### 2008-05-26 paper/paper.tex
Added new plots.
Some more updates to the paper.

@@
 
-\title{Evaluation of Cache-Oblivious Data Structures}
-\subtitle{\textit{Work In Progress!}}
+\title{Evaluation of Cache-Oblivious Data Structures [draft-1]}
 
@@
 
-% TODO: update keywords?
 \keywords{cache efficiency, locality of reference, algorithms, empirical evaluation}
@@
 \subsection{Experimental Framework}
-
-% Describe (globally) what to measure: execution time
-% - NIPS VM
-% - SPIN model checker
-% - Promela-to-NIPS compiler
-
 For our experimental framework, we need a collection of formal models that are
@@
 We now have a single operation that must be implemented by all set data structures.
-Note that the values that need to be stored in the state space search algorithm are virtual machine state descriptions, which (in the NIPS VM) are opaque variable-length strings.
+Recall that the values to be stored are virtual machine state descriptions, which are variable-length binary strings.
 All data structures must therefore support storing strings of varying length efficiently.
@@
 
-Note that if we insert several different values, some values may map to the
-same slot, which is problematic if each slot can only store one value. There
+When several values are inserted, some values may map to the same slot,
+which is problematic if each slot can only store one value. There
 are a lot of different ways to resolve this collision problem; we use
@@
 \subsection{Measurements}
-% Describe how performance was measured
-%  (mock set implementation; times/sizes subtracted)
+Measuring the memory usage of a process can be done by retrieving how much of the
+address space has been allocated by a process; typically called the
+\emph{virtual set size}.
+Since a single page can be mapped into memory more than once (either in multiple
+processes, or in the same process) this is not exactly equal to the amount of
+memory allocated for the process, but since the majority of the memory used by
+the data structures is mapped at exactly one location, it is good enough as
+an indication of memory use.
+
+There are several ways of measuring the time a process takes to execute.
+The simplest is counting the number of seconds elapsed since the start of the
+program (called "wall clock time"). This has the disadvantage that concurrently
+executing processes affect the runtime of the process being measured, so on a
+busy system the reported runtime may vary with different executions.
+
+A different way to measure time is using the statistics the kernel collects
+of how many seconds the CPU spends executing instructions for the process
+(both in user space and in kernel space). These values are independent of
+what other processes are doing, which makes them more consistent across
+multiple executions. However, these values do not account for the time the
+system spends waiting (for example, for a page to be swapped in from disk)
+or the time spend by other processes on behalf of the executing process
+(for example, the kernel swap daemon on Linux). Since these factors may have
+a large effect on the total running time of the algorithm, it is undesirable
+to leave these out.
+
+Therefore, we decided to measure wall clock time. In order to deal with
+variations across multiple executions, each experiment is run three times,
+and the median value is used for further analysis.
+
+\subsubsection{Framework Overhead}
+Finally, there is another factor that must be taken into account: the test
+framework uses several components (such as the NIPS VM and a queue) that are
+not being evaluated, yet which do contribute to the memory use and runtime
+of the process. In order to discount these factors, tests are performed using
+a mock set implementation that gives a near-minimal overhead. This works
+by first running with a real set implementation and logging all results
+(the return values of the \verb#insert(S,x)# calls) to a file, and then
+running a second time while reading the stored values. As a result, the mock
+set implementation has very little memory and runtime overhead.
+
+In the results below, all values are reported relative to the values obtained
+using the mock set implementation, which means the overhead of the test
+framework is removed from the results.
+
+\subsubsection{System Configuration}
+The experiments where performed on a 64-bit Linux (kernel 2.6) system
+with an Intel Xeon E5320 processor (1.86 GHz, 4MB cache) and 8GB of main
+memory and 2GB swap file. Although the system is a multiprocessor system
+(and the E5320 is a dual-core processor) all code is single-threaded so
+only a single core is used when executing tests.
 
 % Describe data set used; per model:
-%   average state size
-%   total size
-%   ratio of insertions/queries
-
-% Note that state size depends on number of processes
+%   average state size?
+%   total size?
+%   ratio of insertions/queries?
 
@@
 of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
-also varies during execution.
+increases during execution.
 
@@
 are created atomically, so the state size is constant after initialization.
-
-% mobile1
-% pftp
-% snoopy
-% sort
-
-% TODO: table of models with parameters,
+\end{itemize}
 
@@
 leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
-peterson\_N & \\
+peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
 \hline
@@
 
-\end{itemize}
+In Table \ref{tab-cases} an overview is given of the parameters of the models used
+and the properties of the execution of the model. Note that the number of transitions
+relative to the number of iterations indicates the percentage of insertion operations
+that results in a new element being added to the set.
+
+% Possible additional cases:
+% - mobile1
+% - pftp
+% - snoopy
+% - sort
 
 \section{Results}
+
+TODO: describe parameters of data structures
+Explain results for B-tree (higher page size packs more data), hash table (larger index, less collissions, cache miss anyway)
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-queuesize}
+\caption{Queue size}
+\label{fig-quesize}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-transitions}
+\caption{Transitions}
+\label{fig-transitions}
+\end{figure}
 
@@
 \includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
-\caption{Eratosthenes/Bender set execution time}
+\caption{Bender set execution time on Eratosthenes}
 \label{fig-eratosthenes-bender_wctime}
@@
 \includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
-\caption{Eratosthenes/Bender set memory usage}
+\caption{Bender set memory usage on Eratosthenes}
 \label{fig-eratosthenes-bender_vss}
@@
 \includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
-\caption{Eratosthenes/hash table execution time}
+\caption{Hash table execution time on Eratosthenes}
 \label{fig-eratosthenes-hash_wctime}
@@
 \includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
-\caption{Eratosthenes/hash table memory usage}
+\caption{Hash table memory usage on Eratosthenes}
 \label{fig-eratosthenes-hash_vss}
@@
 \includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
-\caption{Eratosthenes/B-tree execution time}
-\label{fig-eratosthenes-btree}
+\caption{B-tree execution time on Eratosthenes}
+\label{fig-eratosthenes-btree_wctime}
 \end{figure}
@@
 \includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
-\caption{Eratosthenes/B-tree memory usage}
-\label{fig-eratosthenes-btree}
+\caption{B-tree memory usage on Eratosthenes}
+\label{fig-eratosthenes-btree_vss}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-leader2_wctime}
+\caption{Execution times on Leader2}
+\label{fig-leader2_wctime}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-leader2_vss}
+\caption{Memory usage on Leader2}
+\label{leader2_vss}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-peterson_wctime}
+\caption{Execution time on Peterson\_N}
+\label{fig-peterson_wctime}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{fig-peterson_vss}
+\caption{Memory usage on Peterson\_N}
+\label{fig-peterson_vss}
 \end{figure}
@@
 %          for each set
-%
 
@@
 % Explain results, draw conclusions.
+% - relate to insert/test ratio
+% - relate to hardware (not many levels of hierarchy)
 
 \section{Conclusion}
-% Repeat findings.
+The experimental results clearly show that in terms of both memory usage and execution time,
+the cache-oblivious data structure proposed by Bender, Duan, Iacono and Wu is outperformed
+by traditional data structures in our test scenario.
+The advantage of more cache-friendly behaviour does not appear to be large enough to
+compensate for the increased complexity of the data structure, which results in higher
+memory requirements and increased computational overhead.
+If the data structure does have asymptotic performance benefits, then realistic work loads
+on current hardware systems are not enough to reveal this.
 
-\include{fig-queuesize}
+Of course, this does not prove definitely that cache-oblivious data structures
+are entirely without merit. It should be noted that the experiments performed
+for this research are limited in scope: only a single cache-oblivious data
+structure has been examined, in a single test scenario, on a single hardware
+and software platform. It is possible that in a different setting,
+cache-oblivious data structure do compare favorably to their traditional
+counterparts.
+
+We hope that future research will focus not only on the theoretical performance
+of newly developed algorithms and data structure, but also on the performance
+in practice. Theoretical results are invalueble, but to be of greatest
+significance newly developed data structures and algorithms should have
+demonstrable practical merit as well.
 

### 2008-05-26 search/tests/aggr-results.sh

@@ -14,12 +14,20 @@ aggregate() {
         ./aggregate.py "$column" results/"$program"-"$ext"-* > /tmp/aggregated.txt
         ./subtract.py "$base_path" /tmp/aggregated.txt > aggregated/"$program"-"$ext"_"$column"
     done
-    rm /tmp/aggregated.txt
+    rm -f /tmp/aggregated.txt
 }
 
+aggregate eratosthenes qsz    base
+aggregate eratosthenes trans  base
 aggregate eratosthenes vss    base 01 02 03 04 05 06 07 08 09
 aggregate eratosthenes wctime base 01 02 03 04 05 06 07 08 09
+
+aggregate leader2      qsz    base
+aggregate leader2      trans  base
 aggregate leader2      vss    base 01 02 03
 aggregate leader2      wctime base 01 02 03
+
+aggregate peterson     qsz    base
+aggregate peterson     trans  base
 aggregate peterson     vss    base 01 02 03
 aggregate peterson     wctime base 01 02 03

### 2008-05-26 paper/paper.tex
Revised paper.

@@
 \begin{abstract}
-In modern computer hardware architecture, memory is organized in a hierarchy
+In modern computer hardware architecture memory is organized in a hierarchy
 consisting of several types of memory with different memory sizes, block
 transfer sizes and access times. To reflect this situation more accurately
-than tradition models, the cache-oblivious model has been proposed, and several
-algorithms and data structures have been proposed that perform optimally in
-this model.
-This paper presents the results of an empirical evaluation of various data
+than traditional models, the cache-oblivious model has been proposed, and
+several algorithms and data structures have been developed that perform
+optimally in this model.
+This paper presents the results of an empirical evaluation of several data
 structures in a realistic scenario, and gives insight into the applicability
@@
 
-\keywords{cache efficiency, locality of reference, algorithms, empirical evaluation}
+\keywords{cache efficiency, locality of reference, algorithms}
 
@@
 The processor can directly manipulate data contained in its registers only.
-To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy.
-Memory is typically transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
+To access data in a lower level in the memory hierarchy, the data must be transferred upward through the memory hierarchy.
+Memory is typically transferred in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).
 
@@
 \item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existance of a memory hierarchy, and may therefore not perform optimally in practice.
-\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (,described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
-\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (which will also be described below) which does not allow parametrization with properties of the cache layout.
+\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
+\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (explained below) which does not allow para\-metri\-zation with properties of the cache layout.
 \end{list}
@@
 Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice.
-Altough some research has been done for cache-oblivious data structures, they are seldom used in practice.
+Altough some research has been done on cache-oblivious data structures, they are seldom used in practice.
 The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.
@@
 They make a distinction between internal memory (which is limited in size) and external memory (which is virtually unlimited).
-The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories; additionally, consecutive blocks can be transfered at reduced cost (so-called bulk transfer).
+The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories; additionally, consecutive blocks can be transferred at reduced cost (so-called bulk transfer).
 Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy.
@@
 Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
-Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses.
-This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where, which is usually not the case in practice, making their model of little use in the design and evaluation of practical algorithms.
+Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory addresses is described by a non-decreasing function of these addresses, which means that accessing data at higher addresses can be slower than at lower addresses.
+This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where and that data at all locations can be accessed directly, which is usually not the case in practice.
+As a result, the applicability of their model to the design and evaluation of \emph{practical} algorithms is limited.
 
@@
 Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration.
-The key advantage of this class of algorithms is that they are implicitly tuned to all levels in the memory hierarchy at once.
+The key advantage of this class of algorithms is that, even though they are defined in a two-level memory model, they are implicitly tuned to all levels in the memory hierarchy at once.
 It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.
@@
 However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms.
-Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often prefered that are easier to implement and require less memory).
+Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often preferred that are easier to implement and require less memory).
 
@@
 Several data structures were proposed by Rahman, Cole and Raman \cite{rahman2001opd} amongst them a cache-oblivious exponential search tree with a similar structure as the static search tree proposed by Prokop.
-In their experimental results the cache-oblivious tree performs worse than the alternatives.
-Nevertheless, they conclude that \textit{"cache-obli\-vious data structures may have significant practical importance"}.
+In their experimental results the cache-oblivious tree performs worse than the (non-oblivious) alternatives.
+Nevertheless, they conclude that \textit{``cache-obli\-vious data structures may have significant practical importance''}.
 
-Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list.
-This means buckets have to be resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).
+Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize seperate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory.
+Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.
 
@@
 \subsection{State Space Search}
-As the test application, that of a state space search has been chosen. This is a
-suitable scenario for two reasons. First, it is commonly used as a practical
+The test application that we use to gather performance data is an implementation
+of a state space search algorithm.
+This is a suitable scenario for two reasons.
+First, it is commonly used as a practical
 component of formal methods for software verification, and therefore good
@@
 
-State space search is used to verify the correctness of software programs. For
-this purpose, programs are first modeled using a formal language, that is also
-used to specify properties of the program that should hold during its
+State space search can be used to verify the correctness of software programs.
+For this purpose, programs are first modeled using a formal language, that is also used to specify properties of the program that should hold during its
 execution. An executing program can be in a (possibly infinite) amount of states,
@@
 Generating the successors of a particular state is also called
-\emph{expanding} the state. Of course, the model must include some form of
-non-determinism to allow more than one successor to exist for a single state.
+\emph{expanding} the state. Of course, the the execution model
+must include some form of non-determinism to allow more than one successor to
+exist for a single state.
 In practice, this non-determinism usually comes from processes that execute in
-parallel, where the precise interleaving of the (observed) execution of
+parallel, where the precise interleaving of the execution of
 instructions in these processes is non-deterministic, unless synchronization
@@
 $N$ steps) or depth-first (meaning that states that are generated most recently
-by the successor function are expanded first). In the first case, the queue
-operates by a first-in first-out principle, while in the second case the queue
-works in a last-in first-out (stack) principle. In general, breadth-first
-search requires a much larger queue capacity than depth-first search.
+by the successor function are expanded first). We will use breadth-first
+search, which means the queue operates by a first-in first-out principle.
 
@@
 execute them. For the first part, we have looked at Spin \cite{spin}, a widely
-used model checking tool. Spin uses a custom modeling language, called
-PROMELA, to specify models and comes with a collection of example models that
-are suitable for our experiment.
+used model checking tool. Spin uses a custom modeling language (called
+PROMELA) to specify models and is distributed with a collection of example
+models that are suitable for our experiments.
 
-For the execution of these models we will use NIPS VM \cite{weber2007evm}, a
+For the execution of these models we will use the NIPS VM \cite{weber2007evm}, a
 high-performance virtual machine for state space generation that is easily
@@
 custom format, a PROMELA compiler is also available to generate the required
-bytecode from the Spin models.\cite{nipsvm} The NIPS VM is prefered to over the
+bytecode from the Spin models.\cite{nipsvm} The NIPS VM is preferred to the
 Spin tools because it was designed to be embedded in other programs and as
@@
 The NIPS VM represents program state as a binary string; the size of the state
-depends on the number of active processes in the program, and is typically in
-the order of a few hundred bytes.
+depends amongst other on the number of active processes in the program (which may change over the execution of the program). A typical state size is in the order of a few hundred bytes.
 
@@
 The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
-\item \textbf{Cache-oblivious}: the B-tree-like data structure proposed by Bender, Duan, Iacono and Wu.
-Since it provides functionality comparable to that of a normal B-tree, this seems like a fair candidate for a comparison.
-For brevity, this data structure will be refered to as a \emph{Bender set}.
+\item \textbf{Cache-oblivious}: the data structure proposed by Bender, Duan, Iacono and Wu.
+Since it provides functionality comparable to that of a B-tree, this seems like a fair candidate for a comparison.
+For brevity, this data structure will be referred to as a \emph{Bender set}.
 \end{itemize}
@@
 undesirable to use existing libraries for a fair comparison, for two reasons.
-
 First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations.
@@
 We call this operation \verb#insert(S,x)#.
-If an element $x$ does not exist in $S$, then $insert(S,x)$ inserts it and returns $0$.
-If $x$ is already an element of $S$, $insert(S,x)$ returns $1$.
-The inner loop of our state space algorithm can now be written as follows:
+If an element $x$ does not exist in $S$, then \verb#insert(S,x)# inserts it and returns $0$.
+If $x$ is already an element of $S$, \verb#insert(S,x)# returns $1$ and no modifications are made.
+The inner loop of our state space algorithm can then be rewritten as follows:
 \begin{verbatim}
@@
 \subsubsection{Common Implementation Details}
-For memory management, the data structures make use of memory mapped files. The
-B-tree and hash table implementations use a single file for all data storage;
+For memory management, the data structures make use of memory mapped files.
+The B-tree and hash table implementations use a single file for all data storage;
 the Bender tree uses several files (as discussed below) and some main memory.
@@
 The use of memory mapped files means that the operating system is in charge of
-when and which pages are transfered between main memory and disk storage; the
+when and which pages are transferred between main memory and disk storage; the
 application assumes no control over this policy. This is in accordance with the
 cache-oblivious model, that assumes no control over the cache is possible.
+Additionally, memory mapped files bypass the C library's allocation functions
+and give the implementor complete control over how data is ordered in (virtual)
+memory.
 
 \subsubsection{Hash Table Implementation}
-Hash tables are widely used as a fast (O($1$)) data structure especially when
-used for in-memory storage. In its simplest form, a hash table consists of an
+In its simplest form, a hash table consists of an
 index array (the \emph{index}) with a fixed number of slots. A \emph{hash
@@
 
-B-trees are similar to other search tree data structures. A defining property
-is that they organize data in pages of a fixed size, each containing several
-values, which makes them especially useful for purposes of on-disk storage
-where reading and writing data in a few large blocks is relatively cheap
-compared to accessing several smaller blocks.
+B-trees are similar to other search tree data structures in the sense that
+they store ordered values hierarchically in such a way that each subtree store
+values that are close together.
+A key property of B-trees is that they do not fix the number of children per node, but instead organize data in pages of a fixed size, each containing
+as many values as will fit. This makes them especially suitable for purposes
+of on-disk storage where reading and writing data in a few large blocks is
+relatively cheap compared to accessing several smaller blocks.
 
-B-trees work by keeping as much values as possible, in order, in a single
-page. Pages are ordered in a tree structure, where interior nodes store
-pointers to child nodes. Pointers originate at a pair of adjacent values,
-and point to a subtree containing only values that fall in between those
-values.
+B-tree pages are ordered in a tree structure. Values in a page are stored
+in order, and pointers are added that originate between a pair of adjacent
+values, and point to a subtree containing only values that fall in between those
+values. The resulting tree is typically very shallow, which is benefical,
+as the number of pages that need to be retrieved (worst case) is equal to
+the height of the tree.
 
@@
 page needs to be split, a new (empty) top-level page is created and the height
-of the B-tree is increased by one.
-
-B-trees are well suited to storing variable length values, as pages can
-store many values as long as the combined size of the values and pointers
-does not exceed the pagesize.
+of the B-tree is increased by one. As a result, all leaf nodes in a B-tree are
+at the same depth, and all pages stored are at least half full.
 
@@
 
-Note that there is no obvious way to store values that are larger than the
-size of a single page. Production-quality implementations handle this case
-separately, but our implementation does not support this scenario and
-requires stored values to be smaller than the pagesize. Additionally,
-deletion of items has not been implemented.
+B-trees can easily support the insertion of variable-length values, as long
+as individual values fit in the size of a single page. However, there is no
+obvious way to store values that are larger than the size of a single page.
+Production-quality implementations handle this case separately. Our our
+implementation does not support this requires stored values to
+be smaller than the pagesize. Additionally, deletion of items has not been
+implemented.
 
@@
 remains below 50\% while still allowing values of various sizes to be inserted.
+In the following description, we will assume that values are of a fixed size.
 
 The implementation of the Bender set consists of two parts: a sparse array
-storing the values in-order, and a binary tree stored in van Emde Boas
-layout (similar to the one proposed by Prokop). Both the tree and the array
-may contain special \emph{empty values}. A depiction of a Bender set storing
-four values (with a capacity for eight) is given in Figure \ref{fig-bender-set}.
+storing the values in order, and a binary tree stored in van Emde Boas
+layout (similar to the one proposed by Prokop) that is used to find values
+efficiently.
+Both the tree and the array may contain special \emph{empty values}.
+A depiction of a Bender set storing four values (with a capacity for eight) is
+given in Figure \ref{fig-bender-set}.
 
@@
 On the highest level, there is a single window of size $C$. Each subsequent level
-has twice as many windows of half that size, and the lowest level has size $log_2 C$
-(rounded up to the next power of 2).
-For each level, a maximum density is given, with the lowest level having density 1,
-the highest level having density $\delta$ (these densities are interpolated linearly
-to obtain the density for the intermediary levels).
+has twice as many windows of half that size, and the lowest level has size
+$log_2 C$ (rounded up to the next power of 2).
+For each level a maximum \emoph{density} is chosen, with the lowest level having
+density 1, the highest level having density $\delta$, and the density for
+the intermediary levels obtained by interpolating linearly between 1 and
+$\delta$.
 
-The top-level density must be a number between $\frac{1}{2}$ and $1$; the optimal value
-depends on practical circumstances such as available memory and the relative cost of
-rebuilding the data structures. Whenever the number of elements stored in
-a window (the window's \emph{population count}) exceeds the size of the window
-multiplied by its maximum density, the window is said to be \emph{overflowing}.
+The top-level density must be a number between $0$ and $1$; the optimal value
+depends on practical circumstances such as available memory and the relative
+cost of rebuilding the data structures.
+Whenever the fraction of non-empty values stored in a window (the window's
+\emph{population count} divided by the size of the window) exceeds its maximum
+density, the window is said to be \emph{overflowing}.
 Overflows are resolved when a higher-level window is \emph{rebalanced},
 meaning that the values in the window are redistributed evenly over the window,
-which will cause some of the values to be moved out from the lower-level window
-which is too full.
+which will cause some of the values to be moved out of the lower-level windows
+which are too full.
 
-Now, to insert a value, the binary tree is used to find a position in the array
+To insert a value, the binary tree is used to find a position in the array
 where the value may be inserted (just before its successor). Then, the windows
@@
 window that can support another element without overflowing is selected, and
-rebalanced (thereby resolving the overflow in all lower level windows).
+then rebalanced (thereby resolving the overflow in all lower level windows).
 
-In the case that all windows overlapping the goal position are overflowing, the
-capacity of the data structure is doubled, a new top level is created and then
-the entire array is rebalanced. This is how the set grows when the number of
-elements stored comes too close to the storage capacity.
+In the case that all windows overlapping the target position would overflow upon
+insertion of the new element, the capacity of the data structure is doubled,
+a new top level is created and then the entire array is rebalanced. This is how
+the set grows when the number of elements stored comes too close to the storage
+capacity. Since changing the capacity causes all values in the array to be moved
+and requires that the entire index tree is recreated, this is an expensive
+operation; however, it occurs only infrequently.
 
@@
 is simply inserted in the middle. In our experiments, this optimization
-increased performance somewhat.
+yielded an reduction in execution time.
 
 One thing that is not specified in the paper by Bender et al, is how the
-population count for windows is kept. One simply option is to keep no such
+population count for windows is kept. The simplest option is to keep no such
 information and simply scan part of the array whenever a population count is
@@
 for the lowest-level windows, and recompute the counts for higher-level windows
-when required. This saves a lot of recomputation, while keeping the additional
+when required. This prevents a lot of recomputation, while keeping the additional
 costs of updating low.
@@
 Measuring the memory usage of a process can be done by retrieving how much of the
-address space has been allocated by a process; typically called the
-\emph{virtual set size}.
-Since a single page can be mapped into memory more than once (either in multiple
-processes, or in the same process) this is not exactly equal to the amount of
-memory allocated for the process, but since the majority of the memory used by
-the data structures is mapped at exactly one location, it is good enough as
-an indication of memory use.
+address space has been allocated by a process; called the \emph{virtual set
+size}. This does not necessarily correspond one-to-one with memory being allocated for the process, but since the majority of the memory used by the
+data structures is mapped at exactly one location, this metric is suitable
+for our experiments.
 
@@
 The simplest is counting the number of seconds elapsed since the start of the
-program (called "wall clock time"). This has the disadvantage that concurrently
-executing processes affect the runtime of the process being measured, so on a
-busy system the reported runtime may vary with different executions.
+program (called \emph{wall clock time}).
+This has the disadvantage that concurrently executing processes affect the runtime
+of the process being measured, so on a busy system the reported runtime may vary
+over different executions.
 
 A different way to measure time is using the statistics the kernel collects
-of how many seconds the CPU spends executing instructions for the process
+of how many seconds the processor spends executing instructions for the process
 (both in user space and in kernel space). These values are independent of
@@
 multiple executions. However, these values do not account for the time the
-system spends waiting (for example, for a page to be swapped in from disk)
-or the time spend by other processes on behalf of the executing process
+system spends waiting (for example, for data to read from disk)
+or the time spent by other processes on behalf of the executing process
 (for example, the kernel swap daemon on Linux). Since these factors may have
@@
 
-Therefore, we decided to measure wall clock time. In order to deal with
-variations across multiple executions, each experiment is run three times,
-and the median value is used for further analysis.
+Therefore, we decided to measure wall clock time. To deal with variations
+across multiple executions, each experiment is run three times, and the median
+value is used for further analysis.
 
@@
 (the return values of the \verb#insert(S,x)# calls) to a file, and then
-running a second time while reading the stored values. As a result, the mock
-set implementation has very little memory and runtime overhead.
+running a second time while reading the stored values. In the second run,
+the mock set implementation does not have to actually store or retrieve
+data and consequently has negligible memory and runtime overhead.
 
@@
 \begin{itemize}
-\item\verb#eratosthenes# is a parallel implementation of the sieve
+\item\textbf{Eratosthenes} is a parallel implementation of the sieve
 of Eratosthenes which is used to find prime numbers upto a given maximum.
@@
 
-\item\verb#leader2# is a leader election algorithm with a configurable number
+\item\textbf{Leader2} is a leader election algorithm with a configurable number
 of processes competing for leadership ($N$). Since a constant number of
@@
 
-\item\verb#peterson_N# is a Peterson's solution to the multi-process mutual
+\item\textbf{Peterson\_N} is Peterson's solution to the multi-process mutual
 exclusion problem (using global variables instead of channels for
@@
 \hline
-\textbf{Case} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
+\textbf{Model} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
 \hline
-eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
-leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
-peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
+Eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
+Leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
+Peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
 \hline
@@
 In Table \ref{tab-cases} an overview is given of the parameters of the models used
-and the properties of the execution of the model. Note that the number of transitions
-relative to the number of iterations indicates the percentage of insertion operations
-that results in a new element being added to the set.
+and the resulting properties of the state space search.
+Note that the number of transitions relative to the number of iterations indicates
+the percentage of insertion operations that results in a new element being added
+to the set, which ranges approximately from 4 to 2.7 non-updating queries per
+insertion.
 
@@
 \section{Conclusion}
-The experimental results clearly show that in terms of both memory usage and execution time,
-the cache-oblivious data structure proposed by Bender, Duan, Iacono and Wu is outperformed
-by traditional data structures in our test scenario.
-The advantage of more cache-friendly behaviour does not appear to be large enough to
-compensate for the increased complexity of the data structure, which results in higher
-memory requirements and increased computational overhead.
-If the data structure does have asymptotic performance benefits, then realistic work loads
-on current hardware systems are not enough to reveal this.
+The experimental results clearly show that in terms of both memory usage and
+execution time, the cache-oblivious data structure proposed by Bender, Duan,
+Iacono and Wu is outperformed by traditional data structures in our test scenario.
+The advantage of more cache-friendly behaviour does not appear to be large enough
+to compensate for the increased complexity of the data structure, which results in
+high ermemory requirements and increased computational overhead.
+If the data structure does have asymptotic performance benefits, then realistic
+work loads on current hardware systems are not enough to reveal this.
+These findings are consistent with earlier results, such as obtained by
+Rahman, Cole and Raman, and Olsen and Skov (see Related Work above).
 
@@
 are entirely without merit. It should be noted that the experiments performed
-for this research are limited in scope: only a single cache-oblivious data
-structure has been examined, in a single test scenario, on a single hardware
+are limited in scope: only a single cache-oblivious data structure has been
+examined, in a single test scenario, on a single hardware
 and software platform. It is possible that in a different setting,
-cache-oblivious data structure do compare favorably to their traditional
-counterparts.
+different cache-oblivious data structures do compare favorably to their traditional
+counterparts. However, since the observed differences in performance are fairly
+large, it is unlikely that small changes are enough to close the performance gap.
 
 We hope that future research will focus not only on the theoretical performance
-of newly developed algorithms and data structure, but also on the performance
-in practice. Theoretical results are invalueble, but to be of greatest
+of newly developed algorithms and data structures, but also on the performance
+in practice. Theoretical results are invaluable, but to be of greatest
 significance newly developed data structures and algorithms should have

### 2008-05-26 search/tests/aggr-results.sh

@@ -24,10 +24,10 @@ aggregate eratosthenes wctime base 01 02 03 04 05 06 07 08 09
 
 aggregate leader2      qsz    base
 aggregate leader2      trans  base
-aggregate leader2      vss    base 01 02 03
-aggregate leader2      wctime base 01 02 03
+aggregate leader2      vss    base 01 02 03 04
+aggregate leader2      wctime base 01 02 03 04
 
 aggregate peterson     qsz    base
 aggregate peterson     trans  base
-aggregate peterson     vss    base 01 02 03
-aggregate peterson     wctime base 01 02 03
+aggregate peterson     vss    base 01 02 03 04
+aggregate peterson     wctime base 01 02 03 04

### 2008-05-26 paper/paper.tex
Fixed some typos. Moved some bibiography entries around.

@@
 Registers  &   32 bytes & 4 bytes & 1 cycle \\
-L1 Cache   &   32 KB    & 64 bytes & 3 cyles \\
+L1 Cache   &   32 KB    & 64 bytes & 3 cycles \\
 L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
@@
 \begin{list}{}{}
-\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existance of a memory hierarchy, and may therefore not perform optimally in practice.
+\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existence of a memory hierarchy, and may therefore not perform optimally in practice.
 \item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
@@
 Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice.
-Altough some research has been done on cache-oblivious data structures, they are seldom used in practice.
+Although some research has been done on cache-oblivious data structures, they are seldom used in practice.
 The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.
@@
 Data is transferred between the two in aligned data blocks of size B.
-In contrast with the hierarchical memory model, the application does not have explicit control over the transfering of blocks between the two memories.
+In contrast with the hierarchical memory model, the application does not have explicit control over the transferring of blocks between the two memories.
 Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program.
-Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and datastructures.
+Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and data structures.
 Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
@@
 
-Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize seperate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory.
+Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory.
 Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.
@@
 accomplish much by itself; instead, the real work is done by the successor
-function and the queue and the set datastructures. Queues can easily be
+function and the queue and the set data structures. Queues can easily be
 implemented efficiently (adding and removing states takes O(1) time).
@@
 Additionally, memory mapped files bypass the C library's allocation functions
-and give the implementor complete control over how data is ordered in (virtual)
+and give the implementer complete control over how data is ordered in (virtual)
 memory.
@@
 function} is used to map key values onto slot numbers. If the slot for a value
-is empty, it may be inserted there. Queries for the existance of an element in
+is empty, it may be inserted there. Queries for the existence of an element in
 the hash table similarly see if the queried value is stored at its designated
@@
 values, and point to a subtree containing only values that fall in between those
-values. The resulting tree is typically very shallow, which is benefical,
+values. The resulting tree is typically very shallow, which is beneficial,
 as the number of pages that need to be retrieved (worst case) is equal to
@@
 implementation does not support this requires stored values to
-be smaller than the pagesize. Additionally, deletion of items has not been
+be smaller than the page size. Additionally, deletion of items has not been
 implemented.
@@
 This tree can be used in a similar way as a binary search tree: by examining the
-two children of a node, we can determine wether a searched-for value lies in the
+two children of a node, we can determine whether a searched-for value lies in the
 left or right subtree.
@@
 \item\textbf{Eratosthenes} is a parallel implementation of the sieve
-of Eratosthenes which is used to find prime numbers upto a given maximum.
+of Eratosthenes which is used to find prime numbers up to a given maximum.
 It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
@@
 TODO: describe parameters of data structures
-Explain results for B-tree (higher page size packs more data), hash table (larger index, less collissions, cache miss anyway)
+Explain results for B-tree (higher page size packs more data), hash table (larger index, less collisions, cache miss anyway)
 
@@
 Iacono and Wu is outperformed by traditional data structures in our test scenario.
-The advantage of more cache-friendly behaviour does not appear to be large enough
+The advantage of more cache-friendly behavior does not appear to be large enough
 to compensate for the increased complexity of the data structure, which results in

### 2008-05-26 paper/paper.tex
More updates to the paper including some pictures of data structures.

@@
-% TODO: ensure all references in paper.bib are cited in the paper!
-% TODO: format checklist (see e-mail)
 \documentclass{acm_proc_article-sp}
@@
 \keywords{cache efficiency, locality of reference, algorithms}
-
-% TODO: more meaningful section titles?
 
@@
 He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.
-
-% Copy and shorten introduction from proposal
-% -> Motivate research into cache-efficient algorithms
-% -> Move "classification" subsection describing three models to front
-% -> Describe each of the three models
-% -> Introduce three data structures:
-%     - note that hash tables have poor locality of reference
-%     - note that btrees have good locality of reference but require pagesize
-% -> Finally, present goal of the research (incorporate "Research Questions" in
-% proposal)
-
-% Marielle suggested:
-%  - give the "standard" memory model an explicit name
-%    (isn't "cache-unaware" enough?)
-%  - move examples of cache oblivious data structures to front
-%  - mention examples of cache unaware and cache aware algorithms
-%    (at least Btree, tries and hash tables are mentioned)
-
-% Note explicitly that cache-oblivious algorithms are relatively complex
-% resulting in higher constant consts, so that the main question is wether
-% the improved locality of reference compensates for the overhead incurred.
 
@@
 
-% TODO: picture?
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{hash-table}
+\caption{A depiction of a separate-chaining hash table}
+\label{fig-hash-table}
+\end{figure}
 
@@
 support removing elements from the hash table, which means we do not have to
-deal with holes that would otherwise occur in the stored file. For our
+deal with holes that would otherwise occur in the stored file.
+Figure \ref{hash-table} shows a hash table (with an index size of four,
+storing three values) and the way it is stored in consecutive memory.
+
+For our
 experiments the FNV-1a hash function \cite{noll2004fnv} is used (modulo the
@@
 
-B-tree pages are ordered in a tree structure. Values in a page are stored
-in order, and pointers are added that originate between a pair of adjacent
-values, and point to a subtree containing only values that fall in between those
-values. The resulting tree is typically very shallow, which is beneficial,
-as the number of pages that need to be retrieved (worst case) is equal to
-the height of the tree.
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{b-tree}
+\caption{A depiction of a B-tree}
+\label{fig-b-tree}
+\end{figure}
 
-When a new value does not fit in the page where it is to be inserted,
-the page is split into two pages; the median value from the old page is
-moved to the parent page, and two new pages (containing the values less
-than respectively greater than the median) are created. When the top-most
-page needs to be split, a new (empty) top-level page is created and the height
-of the B-tree is increased by one. As a result, all leaf nodes in a B-tree are
-at the same depth, and all pages stored are at least half full.
+B-tree pages are ordered in a tree structure. Figure \ref{fig-b-tree} depicts a
+B-tree of height two storing eight values.
+In each page the values are stored in lexicographical order, and the values
+in the first leaf page are lexicographically smaller than ``John'', the values
+in the second leaf page are between ``John'' and ``Philip'' and the values in
+the third leaf page are greater than ``Philip''. Note that since the page
+size is fixed, not all pages can be completely filled.
+
+Since every page can store many values, the resulting tree is typically very
+shallow, which is beneficial as the number of pages that need to be retrieved
+is equal to the height of the tree (worst case).
+
+To insert a value, simply find the leaf page where it belongs. If this page
+does not have enough free space to insert it, the page is split into two pages;
+the median value from the old page is moved to the parent page, and two new pages
+(containing the values less than respectively greater than the median) are
+created. When the top-most page needs to be split, a new (empty) top-level page is
+created and the height of the B-tree is increased by one. As a result, all leaf
+nodes in a B-tree are at the same depth, and all pages stored are at least half
+full, except possibly the root page.
 
@@
 A depiction of a Bender set storing four values (with a capacity for eight) is
-given in Figure \ref{fig-bender-set}.
+given in Figure \ref{fig-bender-set}, with the window population counts for
+the sparse array given on the left (density thresholds not shown) and the index tree for the same array on the right.
 
@@
 \centering
-\includegraphics{bender-set}
-\caption{A depiction of a Bender set}\label{fig-bender-set}
+\includegraphics[width=83mm]{bender-set}
+\caption{A depiction of a Bender set}
+\label{fig-bender-set}
 \end{figure}
@@
 $log_2 C$ (rounded up to the next power of 2).
-For each level a maximum \emoph{density} is chosen, with the lowest level having
+For each level a maximum \emph{density} is chosen, with the lowest level having
 density 1, the highest level having density $\delta$, and the density for
@@
 
+% TODO: add state sizes to table?
+
 In Table \ref{tab-cases} an overview is given of the parameters of the models used
 and the resulting properties of the state space search.
-Note that the number of transitions relative to the number of iterations indicates
-the percentage of insertion operations that results in a new element being added
-to the set, which ranges approximately from 4 to 2.7 non-updating queries per
-insertion.
+Note that the number of transitions relative to the number of iterations gives an
+indication of the ratio of value look-ups (insertions of values that are already
+present in the set) and actual insertions, ranging from approximately from 4 to
+2.7 look-ups per insertion.
 
@@
 \section{Results}
+Each of the tree data structures has some configurable parameters. The hash table
+needs to be parametrized with the size of the index, the B-tree with the size of
+the pages, and the Bender set with the density parameter ($\delta$). To determine
+which parameters to use, various different parameters were first tried on the
+first (and smallest) test case. In this case, a million relatively small states
+need to be stored.
 
-TODO: describe parameters of data structures
-Explain results for B-tree (higher page size packs more data), hash table (larger index, less collisions, cache miss anyway)
+Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
+table. About a million states need to be stored and as expected, the hash table
+with an index size that is close to this number performs best. The hash table
+with a small index (100,000 slots) performs well initially, but gets slower as
+it becomes too full, at which point each slot in the hash table has to store a
+long list of values that map to that slot.
+The hash table with a large index (10,000,000 slots) is a bit slower than the hash
+table with an index size of 1,000,000, possible because of cache misses caused by
+the larger index size.
+As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
+usage by the different hash tables is in the constant size of the index.
+Despite not being the fastest in this case, we will use the hash table with an
+index size of 10,000,000 for further experiments, as other test cases require
+storing well over 1 million states and we want to use a single size that works
+well for all test cases.
 
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-queuesize}
-\caption{Queue size}
-\label{fig-quesize}
-\end{figure}
+Figure \ref{fig-eratosthenes-btree_wctime} and Figure
+\ref{fig-eratosthenes-btree_vss} show the execution time and memory usage
+for the B-tree with page sizes ranging from 1KB to 16KB. The execution platform
+maps memory in 4KB pages, which would suggest that using pages less than 4KB
+makes little sense. Indeed, the B-tree with 1KB pages seems to perform worst,
+but if we take the memory use into account, this is more likely due to the fact
+that fewer items fit on a single page (which means a relatively large portion of
+the page remains unused) than anyting else.
+The 4KB and 16KB pages are relatively close (both in execution time and memory
+usage); we select the 16KB page size for further experiments because state sizes
+are larger in other test cases, in which case the 4KB page size may cause similar
+problems as the 1KB page size here.
 
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-transitions}
-\caption{Transitions}
-\label{fig-transitions}
-\end{figure}
+Finally, in Figure \ref{fig-eratosthenes-bender_wctime} the execution times for
+the Bender set with several different density parameters are given. With a lower
+density, more space is required, but new insertions less frequently cause large windows to be rebalanced.
+Suprisingly, the execution times for the different density values are comparable.
+Figure \ref{fig-eratosthenes-bender_vss} shows that the lower density parameters do
+indeed use a lot more memory. It appears that the advantage of low density is
+negated by the overhead of constructing increasingly large data structures.
+The set with $\delta=0.125$ cannot even finish the test case in the memory
+that is available. Since execution times are similar, we select
+$\delta=0.5$ for future experiments in order to minimize memory use.
 
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
-\caption{Bender set execution time on Eratosthenes}
-\label{fig-eratosthenes-bender_wctime}
-\end{figure}
+% TODO: test with larger density!
 
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
-\caption{Bender set memory usage on Eratosthenes}
-\label{fig-eratosthenes-bender_vss}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
-\caption{Hash table execution time on Eratosthenes}
-\label{fig-eratosthenes-hash_wctime}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
-\caption{Hash table memory usage on Eratosthenes}
-\label{fig-eratosthenes-hash_vss}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
-\caption{B-tree execution time on Eratosthenes}
-\label{fig-eratosthenes-btree_wctime}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
-\caption{B-tree memory usage on Eratosthenes}
-\label{fig-eratosthenes-btree_vss}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-leader2_wctime}
-\caption{Execution times on Leader2}
-\label{fig-leader2_wctime}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-leader2_vss}
-\caption{Memory usage on Leader2}
-\label{leader2_vss}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-peterson_wctime}
-\caption{Execution time on Peterson\_N}
-\label{fig-peterson_wctime}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{fig-peterson_vss}
-\caption{Memory usage on Peterson\_N}
-\label{fig-peterson_vss}
-\end{figure}
-
-% Present results!
-% Tables, graphs, etc.
-% What to include:
-%   queue size over iterations
-%       for each program
-%
-%   insertions per query:
-%       for each program
-%
-%   per program:
-%       wctime over iterations
-%          for each set
-%       memory over iterations
-%          for each set
+Now that we have established the parameters to use for the other test cases.
+The execution times for the three testcases are presented in figures
+\ref{fig-eratosthenes_wctime}, \ref{fig-leader2_wctime} and
+\ref{fig-peterson_wctime}, while the corresponding memory usage is presented in
+figures \ref{fig-eratosthenes_vss}, \ref{fig-leader2_vss} and \ref{fig-peterson_vss} respectively.
 
 \section{Discussion}
-% Explain results, draw conclusions.
-% - relate to insert/test ratio
-% - relate to hardware (not many levels of hierarchy)
+The three test cases paint a very similar picture: in all cases, the hash table
+offers the best performance in terms of both execution time and memory usage,
+followed by the B-tree. In all cases, the Bender set performs worst by a large
+margin.
+
+The difference in execution time between the hash table and B-tree increases as
+the size of the states becomes larger. This is explained by the fact that the
+height of the B-tree depends on the average number of values per page; larger
+values means less values per page, and therefore a deeper tree, and as a
+result more pages to be fetched for a query. The hash table does not have such a
+limitation, as every value in the bucket must be fetched independently, regardless
+of the size of these values.
+
+When it comes to memory usage the B-tree and hash table have similar requirements;
+the B-tree has slightly larger overhead per value stored, but initially the hash
+table uses more memory because of the space allocated for the index.
+
+The performance of the Bender set does not seem to depend on the page size,
+and as a result the difference between the Bender set and the B-tree is smallest
+when the page size is largest. Unfortunately, even then it requires about three
+times as much time and a lot more memory.
+
+The test cases used all have a relatively high percentage of insertions relative
+to value look-ups. This may explain the good performance of the hash table (for
+which insertions are barely more expensive than look-ups) as well as the bad
+performance of the Bender tree (for which insertions are relatively expensive,
+especially when a window is rebalanced). Unfortunately, the insert/look-up
+ratio for our test cases varies too little to derive a meaningful conclusion
+in this regard.
+
+Finally, it should be noted that in our experiments only part of the memory
+hierarchy was used (up to the use of main memory). This still involves several
+levels of processor cache, but it is not a very deep hierarchy. With a deeper
+memory hierarchy, with greater differences in access times between the levels,
+the cache-efficient data structures (the B-tree and the Bender set) should
+perform relatively better.
 
@@
 \bibliography{paper}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=83mm]{fig-queuesize}
+%\caption{Queue size}
+%\label{fig-quesize}
+%\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=83mm]{fig-transitions}
+%\caption{Transitions}
+%\label{fig-transitions}
+%\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
+\caption{Execution time of hash table on Eratosthenes}
+\label{fig-eratosthenes-hash_wctime}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
+\caption{Execution time of B-tree on Eratosthenes}
+\label{fig-eratosthenes-btree_wctime}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
+\caption{Execution time of Bender set on Eratosthenes}
+\label{fig-eratosthenes-bender_wctime}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
+\caption{Memory usage of hash table on Eratosthenes}
+\label{fig-eratosthenes-hash_vss}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
+\caption{Memory usage of B-tree on Eratosthenes}
+\label{fig-eratosthenes-btree_vss}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
+\caption{Memory usage of Bender set on Eratosthenes}
+\label{fig-eratosthenes-bender_vss}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes_wctime}
+\caption{Execution time on Eratosthenes}
+\label{fig-eratosthenes_wctime}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-leader2_wctime}
+\caption{Execution time on Leader2}
+\label{fig-leader2_wctime}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-peterson_wctime}
+\caption{Execution time on Peterson\_N}
+\label{fig-peterson_wctime}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-eratosthenes_vss}
+\caption{Memory usage on Eratosthenes}
+\label{fig-eratosthenes_vss}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-leader2_vss}
+\caption{Memory usage on Leader2}
+\label{fig-leader2_vss}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=83mm]{fig-peterson_vss}
+\caption{Memory usage on Peterson\_N}
+\label{fig-peterson_vss}
+\end{figure}
+
+% What else to include:
+%   queue size over iterations
+%       for each program
+%
+%   insertions per query:
+%       for each program
+%
+
 \end{document}

### 2008-05-31 paper/paper.tex
Updated paper abstract. Updated plot labels.

@@
 
-\title{Evaluation of Cache-Oblivious Data Structures [draft-1]}
+\title{Evaluation of a Cache-Oblivious Data Structure [draft-1]}
 
@@
 consisting of several types of memory with different memory sizes, block
-transfer sizes and access times. To reflect this situation more accurately
-than traditional models, the cache-oblivious model has been proposed, and
-several algorithms and data structures have been developed that perform
-optimally in this model.
+transfer sizes and access times.
+Traditionally, data structures are evaluated in a theoretical model that
+does not take the existance of a memory hierarchy into account. The
+cache-oblivious model has been proposed as a more accurate model.
+Although several data structures have been described in this model
+relatively little empirical performance data is available.
 This paper presents the results of an empirical evaluation of several data
-structures in a realistic scenario, and gives insight into the applicability
-of cache-oblivious data structure in practice.
+structures in a realistic scenario and aims to provide insight into
+the applicability of cache-oblivious data structures in practice.
 \end{abstract}
@@
 density, more space is required, but new insertions less frequently cause large windows to be rebalanced.
-Suprisingly, the execution times for the different density values are comparable.
-Figure \ref{fig-eratosthenes-bender_vss} shows that the lower density parameters do
-indeed use a lot more memory. It appears that the advantage of low density is
+The execution times for the density values of 0.5 and lower are almost the same
+while larger density values are slower.
+Figure \ref{fig-eratosthenes-bender_vss} shows that the lower the density value,
+the more memory is used. It appears that the advantage of low density is
 negated by the overhead of constructing increasingly large data structures.
 The set with $\delta=0.125$ cannot even finish the test case in the memory
-that is available. Since execution times are similar, we select
-$\delta=0.5$ for future experiments in order to minimize memory use.
+that is available. We select $\delta=0.5$ for future experiments, which strikes
+a good balance between execution time and required memory.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{paper}

### 2008-05-31 search/tests/aggr-results.sh

@@ -19,8 +19,8 @@ aggregate() {
 
 aggregate eratosthenes qsz    base
 aggregate eratosthenes trans  base
-aggregate eratosthenes vss    base 01 02 03 04 05 06 07 08 09
-aggregate eratosthenes wctime base 01 02 03 04 05 06 07 08 09
+aggregate eratosthenes vss    base 01 02 03 04 05 06 07 08 09 10 11
+aggregate eratosthenes wctime base 01 02 03 04 05 06 07 08 09 10 11
 
 aggregate leader2      qsz    base
 aggregate leader2      trans  base

### 2008-05-31 search/tests/subtract.py

@@ -49,7 +49,7 @@ for line in file(argv[2]):
     if key not in base:
         print >>stderr, 'WARNING: missing base value for key', key
         continue
-    medi = parse(row[1]) - base[key]
-    mini = parse(row[2]) - base[key]
-    maxi = parse(row[3]) - base[key]
+    medi = max(0, parse(row[1]) - base[key])
+    mini = max(0, parse(row[2]) - base[key])
+    maxi = max(0, parse(row[3]) - base[key])
     print "%19s %19s %19s %19s" % (str(key), str(medi), str(mini), str(maxi))

### 2008-05-31 paper/paper.tex
Updates to the draft paper.

@@
 
-One of the reasons this assumption fails is that modern operating systems employ virtual memory management, which allows (slower) disk-based storage to be dynamically substituted for (fast\-er) main memory when the latter is in short supply.
-Another reason is that as processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
-Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem, but this increases the differences in access time between different types of memory.
+The main cause for this is the increasing difference between processor and memory speeds.
+As processor speeds have increased greatly, the time required to transfer data between processor and main memory has become a bottleneck for many types of computations.
+Hardware architects have added faster (but small) cache memory at various points in the computer architecture to reduce this problem.
+Similar developments have taken place on the boundary between main memory and disk-based storage.
 
@@
 \hline
-Registers  &   32 bytes & 4 bytes & 1 cycle \\
-L1 Cache   &   32 KB    & 64 bytes & 3 cycles \\
-L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
-RAM        &   ~2 GB    & 1 KB     & 50-100 cycles \\
-Disk       & ~300 GB    & 4 KB     & 5,000-10,000 cycles \\
+Registers  &   32~bytes & 4~bytes & 1 cycle \\
+L1 Cache   &   32~KB    & 64~bytes & 3 cycles \\
+L2 Cache   &    1~MB    & 64~bytes & 9 cycles \\
+RAM        &  $\pm$ 2 GB    & 1~KB     & 50-100 cycles \\
+Disk       &  $\pm$ 300 GB    & 4~KB     & 5,000-10,000 cycles \\
 \hline
@@
 Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice.
-Although some research has been done on cache-oblivious data structures, they are seldom used in practice.
-The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.
+Although some research has been done on cache-oblivious data structures, at the time of writing there is no evidence of their use in practice.
+The main goal for this paper is to give some insight in the practical merit of cache-oblivious data structures.
+
+In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious model is of particular interest.
+We will then describe what our goals for this research project were and how our work relates to previous research.
+A large part of the paper will be dedicated to a description of our research methodology.
+Finally, we will present and discuss our results, and draw a conclusion on the applicability of cache-oblivious data structures.
+
+\section{Previous Work}
+The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model.
+We will briefly describe some of these alternatives here.
 
@@
 \subsection{Cache-Oblivious Memory Model}
-A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model, which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
+A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model.
+In this model, there is an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
 Data is transferred between the two in aligned data blocks of size B.
@@
 Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
-This data structure is mostly of theoretic interest, as it is fairly complex to implement and incurs a relatively large overhead in practice.
-
-Later, Bender, Duan, Iacono and Wu simplified the data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for finger searches.
-The latter data structure consists of a sparse array (stored in a linear amount of memory) with a static tree in van Emde Boas layout (named for similarity to a different data structure proposed by van Emde Boas et al.
-\cite{vanemdeboas1976dai}).
+Later, Bender, Duan, Iacono and Wu simplified this data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for additional operations (finger searches in particular).
+This data structure will be explained in detail in Section \ref{sect-bender-set}.
+The authors note that a chief advantage of their data structure over the previously described on is that it is less complex, more easily implementable and therefore more suitable for practical use.
 
@@
 
-\section{Methods}
+From the available publications we can conclude that only a minority of the research on the cache-oblivious memory model compares the practical performance of newly proposed data structures with that of established data structures.
+Contrary to what theoretical analysis suggests, the practical results that are available so far fail to show the superior performance of cache-oblivious data structures.
+Therefore, additional research is needed to determine more precisely to which extent cache-oblivious data structures are useful as a building block for practical work; this paper will provide some insight in this regard.
+
+\section{Research Methods}
 In order to gather empirical data, the research approach must be made more concrete.
@@
 
-Furthermore, we will need to define what practical performance is.
-We will measure two properties: primarily, the total running time of our test application, and secondarily the memory used.
+We also need to define more accurately what we mean by practical performance.
+We will execute our test application and measure two properties: primarily, the total running time, and secondarily the memory used.
 The rationale for selecting these metrics is that if the data structures perform identical functions and plenty of memory is available, the only observable difference in running a program using different data structures will be the run time.
@@
 
-Depending on how states are inserted and extracted from the queue, the search
-may be breadth-first (meaning that all states that are reachable in $N$ steps
+Although variations are possible, our queue operates by a first-in, first-out
+principle, meaning that meaning that all states that are reachable in $N$ steps
 from the initial state are expanded before any states that require more than
-$N$ steps) or depth-first (meaning that states that are generated most recently
-by the successor function are expanded first). We will use breadth-first
-search, which means the queue operates by a first-in first-out principle.
+$N$ steps.
 
@@
 representative of those typically used for formal verification, and a way to
-execute them. For the first part, we have looked at Spin \cite{spin}, a widely
-used model checking tool. Spin uses a custom modeling language (called
-PROMELA) to specify models and is distributed with a collection of example
-models that are suitable for our experiments.
+execute them. For the first part, we have looked at Spin \cite{holzmann2004smc},
+a widely used model checking tool.
+Spin uses a custom modeling language (called PROMELA) to specify models and is
+distributed with a collection of example models that are suitable for our
+experiments.
 
@@
 custom format, a PROMELA compiler is also available to generate the required
-bytecode from the Spin models.\cite{nipsvm} The NIPS VM is preferred to the
+bytecode from the Spin models \cite{nipsvm}. The NIPS VM is preferred to the
 Spin tools because it was designed to be embedded in other programs and as
@@
 The NIPS VM represents program state as a binary string; the size of the state
-depends amongst other on the number of active processes in the program (which may change over the execution of the program). A typical state size is in the order of a few hundred bytes.
+depends amongst other on the number of active processes in the program (which
+may change over the execution of the program).
+A typical state size is in the order of a few hundred bytes.
 
@@
 
+It should be noted that since we only need a limited subset of the functionality
+offered by the set data structures for our test application, we did not
+implement any operations that were not required to perform our experiments.
+However, we did not change the design of the data structures to take advantage
+of this. That means that any additional operations could be implemented without
+structural changes to our existing implementation.
+
 \subsubsection{Hash Table Implementation}
@@
 are a lot of different ways to resolve this collision problem; we use
-\emph{separate chaining} which means that slots do not store values directly,
+\emph{separate chaining}, which means that slots do not store values directly,
 but instead each slot stores a pointer to a singly-linked list of values that
@@
 values that are close together.
-A key property of B-trees is that they do not fix the number of children per node, but instead organize data in pages of a fixed size, each containing
+A key property of B-trees is that they do not fix the number of children per
+node, but instead organize data in pages of a fixed size, each containing
 as many values as will fit. This makes them especially suitable for purposes
@@
 
-To insert a value, simply find the leaf page where it belongs. If this page
-does not have enough free space to insert it, the page is split into two pages;
-the median value from the old page is moved to the parent page, and two new pages
-(containing the values less than respectively greater than the median) are
-created. When the top-most page needs to be split, a new (empty) top-level page is
+New values are inserted into a leaf page which can be determined by traversing
+the tree.
+If this leaf page does not have enough free space to insert the new value,
+the page will have to be \emph{split}: the median value is selected, and the old
+page is replaced by two new pages containing the values less than respectively
+greater than this median value, while the median value itself is moved to the
+parent page.
+When the top-most page needs to be split, a new (empty) top-level page is
 created and the height of the B-tree is increased by one. As a result, all leaf
@@
 
-% TODO: picture and description
-
 B-trees can easily support the insertion of variable-length values, as long
-as individual values fit in the size of a single page. However, there is no
-obvious way to store values that are larger than the size of a single page.
-Production-quality implementations handle this case separately. Our our
-implementation does not support this requires stored values to
-be smaller than the page size. Additionally, deletion of items has not been
-implemented.
+as each individual value fits in a single page. However, values with
+a size larger than the size of a single page must be handled separately.
+Our implementation does not support this and therefore requires all
+stored values to be smaller than the pages.
 
 \subsubsection{Bender Set Implementation}
+\label{sect-bender-set}
 The first implementation challenge for the Bender set is that the description
@@
 is simply inserted in the middle. In our experiments, this optimization
-yielded an reduction in execution time.
+yielded a reduction in execution time.
 
@@
 to efficiently find the successor element for a given value. The tree is
-stored in memory in van Emde Boas lay-out (as described by Prokop) to allow
+stored in memory in van Emde Boas layout (as described by Prokop) to allow
 efficient cache-oblivious traversal.
@@
 address space has been allocated by a process; called the \emph{virtual set
-size}. This does not necessarily correspond one-to-one with memory being allocated for the process, but since the majority of the memory used by the
+size}. This does not necessarily correspond one-to-one with memory being allocated
+for the process, but since the majority of the memory used by the
 data structures is mapped at exactly one location, this metric is suitable
@@
 Therefore, we decided to measure wall clock time. To deal with variations
-across multiple executions, each experiment is run three times, and the median
+across multiple executions, each experiment is run five times, and the median
 value is used for further analysis.
@@
 using the mock set implementation, which means the overhead of the test
-framework is removed from the results.
+framework is removed from the results. Although the extra memory used by
+the test framework is relatively small (less than 10\% in all cases) and
+primarily caused by the queue data structure, the overhead in terms of
+execution time was quite significant: close to 70\% on the worst case.
+This does mean that the set data structure accounts for at least 30\%
+of the execution time of the search algorithm in all cases (and much
+more for the slower data structures).
 
@@
 The experiments where performed on a 64-bit Linux (kernel 2.6) system
-with an Intel Xeon E5320 processor (1.86 GHz, 4MB cache) and 8GB of main
-memory and 2GB swap file. Although the system is a multiprocessor system
+with an Intel Xeon E5320 processor (1.86 GHz, 4~MB cache) and 8~GB of main
+memory and 2~GB swap file. Although the system is a multiprocessor system
 (and the E5320 is a dual-core processor) all code is single-threaded so
 only a single core is used when executing tests.
-
-% Describe data set used; per model:
-%   average state size?
-%   total size?
-%   ratio of insertions/queries?
 
@@
 \begin{center}
-\begin{tabular}{ l l l l }
+\begin{tabular}{ l l r r }
 \hline
@@
 Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
-table. About a million states need to be stored and as expected, the hash table
-with an index size that is close to this number performs best. The hash table
+table.
+The hash table
 with a small index (100,000 slots) performs well initially, but gets slower as
@@
 long list of values that map to that slot.
-The hash table with a large index (10,000,000 slots) is a bit slower than the hash
-table with an index size of 1,000,000, possible because of cache misses caused by
-the larger index size.
 As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
-usage by the different hash tables is in the constant size of the index.
-Despite not being the fastest in this case, we will use the hash table with an
-index size of 10,000,000 for further experiments, as other test cases require
-storing well over 1 million states and we want to use a single size that works
-well for all test cases.
+usage by the different hash tables is in the (fixed) size of the index.
+We will use the hash table with an index size of 10,000,000 for further
+experiments, as it seems to perform best even for a relatively small test case.
 
@@
 \ref{fig-eratosthenes-btree_vss} show the execution time and memory usage
-for the B-tree with page sizes ranging from 1KB to 16KB. The execution platform
-maps memory in 4KB pages, which would suggest that using pages less than 4KB
-makes little sense. Indeed, the B-tree with 1KB pages seems to perform worst,
+for the B-tree with page sizes ranging from 1~KB to 16~KB. The execution platform
+maps memory in 4~KB pages, which would suggest that using pages less than 4KB
+makes little sense. Indeed, the B-tree with 1~KB pages seems to perform worst,
 but if we take the memory use into account, this is more likely due to the fact
@@
 the page remains unused) than anyting else.
-The 4KB and 16KB pages are relatively close (both in execution time and memory
-usage); we select the 16KB page size for further experiments because state sizes
-are larger in other test cases, in which case the 4KB page size may cause similar
-problems as the 1KB page size here.
+The 4~KB and 16~KB pages are relatively close (both in execution time and memory
+usage); we select the 16~KB page size for further experiments because state sizes
+are larger in other test cases, in which case the 4~KB page size may cause similar
+problems as the 1~KB page size here.
 
@@
 
-% TODO: test with larger density!
-
-Now that we have established the parameters to use for the other test cases.
-The execution times for the three testcases are presented in figures
+Now that we have established the parameters to use for the other test cases,
+we can run final experiments on all three test cases.
+The execution times for the these cases are presented in figures
 \ref{fig-eratosthenes_wctime}, \ref{fig-leader2_wctime} and
@@
 margin.
+
+It should be noted that the Bender set is the only data structure that shows
+abrupt changes in both the execution time and memory usage graphs. These
+jumps in the graph occur whenever the capacity of the Bender set is increased;
+this causes the memory allocated for the set to be doubled and the array and
+tree structures to be recreated, which is a relatively expensive operation.
 
@@
 perform relatively better.
+
+\subsection{Implementation Complexity}
+Since all data structures used were implemented from scratch, our research
+also yields some insight in the implementation complexity of the different
+data structures.
 
@@
 %\end{figure}
+
+\section{Results}
 
@@
 
-% What else to include:
-%   queue size over iterations
-%       for each program
-%
-%   insertions per query:
-%       for each program
-%
-
 \end{document}

### 2008-05-31 count-lines.sh
Added tools to count source-lines-of-code.

@@ -1,0 +1,47 @@
+COMMON="
+datastructures/Alloc.h
+datastructures/Alloc.c
+datastructures/config.h
+datastructures/comparison.c
+datastructures/comparison.h
+datastructures/hashing.c
+datastructures/Set.c
+datastructures/Set.h
+datastructures/FileStorage.c
+datastructures/FileStorage.h"
+
+HASHTABLE="
+datastructures/Hash_Set.c"
+
+BTREE="
+datastructures/Btree_Set.c"
+
+BENDERSET="
+datastructures/Bender_Set.c
+datastructures/Bender_Impl.h
+datastructures/Bender_Impl.c"
+
+DEQUE="
+datastructures/Deque.h
+datastructures/File_Deque.c"
+
+SEARCH="
+search/search.h
+search/main.c
+search/search.c
+datastructures/Mock_Set.c
+datastructures/Dummy_Set.c"
+
+report() {
+	LINES=`cat $@ | ./strip-comments.pl | wc -l`
+	echo $LINES
+	TOTAL=$(($TOTAL + $LINES))
+}
+
+echo -n "Common:           "; report $COMMON
+echo -n "Hash table:       "; report $HASHTABLE
+echo -n "B-tree:           "; report $BTREE
+echo -n "Bender set:       "; report $BENDERSET
+echo -n "Deque:            "; report $DEQUE
+echo -n "Search framework: "; report $SEARCH
+echo -n "TOTAL:           "; echo $TOTAL

### 2008-05-31 strip-comments.pl

@@ -1,0 +1,7 @@
+#!/usr/bin/perl
+
+@lines = <>;
+$text = join '', @lines;
+$text =~ s/\/[*].+?[*]\///sg;
+$text =~ s/\n(\s+\n)+/\n/g;
+print $text;

### 2008-05-31 paper/paper.tex
Updates to the paper draft.

@@
 
-When it comes to memory usage the B-tree and hash table have similar requirements;
+With respect to memory usage the B-tree and hash table have similar requirements;
 the B-tree has slightly larger overhead per value stored, but initially the hash
@@
 
-Finally, it should be noted that in our experiments only part of the memory
+\subsection{Implementation Complexity}
+Since all data structures used were implemented from scratch, our research
+also yields some insight in the implementation complexity of the different
+data structures.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{ l r r }
+\hline
+\textbf{Purpose} & \textbf{Lines} & \textbf{Percentage} \\
+\hline
+Common code       & 433 & 18.21\% \\
+Hash table        & 146 &  6.14\% \\
+B-tree            & 299 & 12.57\% \\
+Bender set        & 623 & 26.20\% \\
+Queue             & 184 &  7.74\% \\
+Search Framework  & 693 & 29.14\% \\
+\hline
+\end{tabular}
+\caption{Source lines of code of the test application}
+\label{tab-loc}
+\end{center}
+\end{table}
+
+Table \ref {tab-loc} gives an overview of the source lines of code used for
+the test application, after removing comments and blank lines (which together
+amount for about 25\% of the code). Common code includes allocation functions,
+interface descriptions and comparison and hashing functions. The framework
+includes not just the search algorithm, but also the functionality to report
+various metrics while running a test case.
+
+Although not a perfect metric of implementation complexity, the lines of code
+required for the various components of the test application do give some
+insight in the relative complexity of the data structures. It is apparent why
+hash tables are popular: they perform well while they are relatively easy to
+implement.
+
+\section{Future work}
+[[Measure different data structures]]
+
+[[Measure performance in low memory environments (using disk)]]
+
+[[Mesure cache performance specifically]]
+
+It should be noted that in our experiments only part of the memory
 hierarchy was used (up to the use of main memory). This still involves several
@@
 perform relatively better.
-
-\subsection{Implementation Complexity}
-Since all data structures used were implemented from scratch, our research
-also yields some insight in the implementation complexity of the different
-data structures.
 

### 2008-05-31 search/tests/run-test.sh
Changed number of repetitions from 3 to 7.

@@ -7,7 +7,7 @@ MODELS=../models/
 # External file declares:
 MODEL=                                  # base name of model to use
 SET=                                    # set configuration 
-REPEAT=3                                # number of iterations
+REPEAT=7                                # number of iterations
 MAXVSS=unlimited                        # maximum virtual set size
 INTERVAL=10000                          # reporting interval
 MAXIT=10000000                          # maximum number of iterations

### 2008-05-31 paper/paper.tex
Updates to the draft paper.

@@
 
-\textit{\small 9$^{th}$ Twente Student Conference on IT, Enschede, June, 2008}
+\textit{\small 9$^{th}$ Twente Student Conference on IT, Enschede, June 23$^{rd}$, 2008}
 
@@
 \section{Related Work}
+\label{sect-related-work}
 Many data structures and algorithms have been analyzed in the cache-oblivious model.
@@
 
-Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory.
+Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory.
 Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.
@@
 deal with holes that would otherwise occur in the stored file.
-Figure \ref{hash-table} shows a hash table (with an index size of four,
+Figure \ref{fig-hash-table} shows a hash table (with an index size of four,
 storing three values) and the way it is stored in consecutive memory.
@@
 The experiments where performed on a 64-bit Linux (kernel 2.6) system
-with an Intel Xeon E5320 processor (1.86 GHz, 4~MB cache) and 8~GB of main
-memory and 2~GB swap file. Although the system is a multiprocessor system
-(and the E5320 is a dual-core processor) all code is single-threaded so
+with an Intel Xeon E5335 processor (2~GHz, 4~MB cache) and 8~GB of main
+memory (no swap file). Although the system is a multiprocessor system
+(and the E5335 is a dual-core processor) all code is single-threaded so
 only a single core is used when executing tests.
@@
 usage by the different hash tables is in the (fixed) size of the index.
-We will use the hash table with an index size of 10,000,000 for further
-experiments, as it seems to perform best even for a relatively small test case.
+Although for this case the hash table with an index size of 1,000,000 is
+slightly faster, we will use an index size of 10,000,000 for further
+experiments, as these cases require more values to be stored.
 
@@
 
-\section{Future work}
-[[Measure different data structures]]
-
-[[Measure performance in low memory environments (using disk)]]
-
-[[Mesure cache performance specifically]]
-
+\section{Future Work}
 It should be noted that in our experiments only part of the memory
 hierarchy was used (up to the use of main memory). This still involves several
-levels of processor cache, but it is not a very deep hierarchy. With a deeper
+levels of processor cache, but it is not a very deep hierarchy. In a deeper
 memory hierarchy, with greater differences in access times between the levels,
 the cache-efficient data structures (the B-tree and the Bender set) should
-perform relatively better.
+perform better. Specifically, using disk-based storage as the lowest level of
+storage seems a logical extension of our research.
+
+In our experiments we did not measure cache efficiency specifically; instead,
+we measured total execution time only, which is affected by several different
+factors of which cache efficiency is only one.
+To better understand how different factors influence performance, it would
+be interesting to measure cache efficiency seperately and reporton actual
+cache hits and misses on different boundaries of the memory hierarchy.
+
+Finally, we only evaluated a single data structure in a single test
+environment (even though we used more than one model to perform experiments).
+To draw more general conclusions about the practical merits of cache-oblivious
+data structures, it will be necessary to perform experiments at a larger scale,
+comparing multiple cache-oblivious data structures and using scenarios that
+differ more.
 
 \section{Conclusion}
-The experimental results clearly show that in terms of both memory usage and
-execution time, the cache-oblivious data structure proposed by Bender, Duan,
-Iacono and Wu is outperformed by traditional data structures in our test scenario.
+The experimental results clearly show that the cache-oblivious data structure
+proposed by Bender, Duan, Iacono and Wu is outperformed by traditional data
+structures in our test scenario, in terms of both execution time and memory use.
 The advantage of more cache-friendly behavior does not appear to be large enough
 to compensate for the increased complexity of the data structure, which results in
-high ermemory requirements and increased computational overhead.
+higher memory requirements and increased computational overhead.
 If the data structure does have asymptotic performance benefits, then realistic
@@
 These findings are consistent with earlier results, such as obtained by
-Rahman, Cole and Raman, and Olsen and Skov (see Related Work above).
+Rahman, Cole and Raman, and Olsen and Skov (see Section \ref{sect-related-work}).
 
-Of course, this does not prove definitely that cache-oblivious data structures
-are entirely without merit. It should be noted that the experiments performed
-are limited in scope: only a single cache-oblivious data structure has been
-examined, in a single test scenario, on a single hardware
-and software platform. It is possible that in a different setting,
-different cache-oblivious data structures do compare favorably to their traditional
-counterparts. However, since the observed differences in performance are fairly
-large, it is unlikely that small changes are enough to close the performance gap.
+Of course, this does not prove conclusively that cache-oblivious data structures
+are entirely without merit. Our experiments are limited in scope: only a single
+cache-oblivious data structure has been examined, in a single test scenario,
+on a single hardware
+and software platform.
+It is possible that in a different setting, different cache-oblivious data
+structures do compare favorably to their traditional counterparts.
+However, since the observed differences in performance are fairly large,
+it is unlikely that small changes are enough to close the performance gap.
 
@@
 %\end{figure}
-
-\section{Results}
 

### 2008-05-31 paper/paper.tex
Updated the paper (updated results included).

@@
 
-Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms \cite{demaine2002coa}.
-He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.
+The lack of unrealistic requirements combined with the promise of cache-efficiency accross multiple levels of the memory hierarchy, makes it an interesting model for the development of low-maintenance, high-performance data structures for use in real-world applications.
 
@@
 Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.
+
+Demaine gives an introduction into the cache-oblivious memory model and an overview of a selection of cache-oblivious data structures and algorithms \cite{demaine2002coa}.
+He also motivates the simplifications made in the cache-oblivious memory model, such as the assumption of full cache associativity and an optimal replacement policy.
 
@@
 to efficiently find the successor element for a given value. The tree is
-stored in memory in van Emde Boas layout (as described by Prokop) to allow
-efficient cache-oblivious traversal.
+stored in memory in van Emde Boas layout to allow cache-oblivious
+traversal. This layout is named after the data structure described in
+\cite{vanemdeboas1976dai} and determines an order in which the nodes of a
+tree can be stored in a contiguous block of memory, in such a way that traversing
+a tree with $N$ nodes from root to leaf requires only $O(\log_B N + 1)$ pages to be
+fetched from memory.
 
-In the tree, the leaf nodes (when numbered from left to right) store the same values
+In van Emde Boas lay-out, which is defined recursively, a tree with 
+$N$ nodes is divided in half vertically, creating a small subtree at the
+top and a number of subtrees (approximately $\sqrt{N}$) below. Each subtree
+has a size of approximately $\sqrt{N}$ nodes and will be stored in a
+contiguous block of memory.
+
+In Figure \ref{fig-van-emde-boas} a binary tree is shown, with the nodes
+labeled with their positions according to the van Emde Boas layout.
+In this example, only two levels of recursion are needed. If each page
+stores three nodes, the path from node 1 to node 12 (or any other leaf node)
+requires visits only two pages.
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{vanemdeboas}
+\caption{A binary tree in van Emde Boas layout}
+\label{fig-van-emde-boas}
+\end{figure}
+
+In the index tree used for the Bender set, the leaf nodes store the same values
 as the sparse array (including empty values).
@@
 Therefore, we decided to measure wall clock time. To deal with variations
-across multiple executions, each experiment is run five times, and the median
+across multiple executions, each experiment is run seven times, and the median
 value is used for further analysis.
@@
 usage by the different hash tables is in the (fixed) size of the index.
-Although for this case the hash table with an index size of 1,000,000 is
-slightly faster, we will use an index size of 10,000,000 for further
-experiments, as these cases require more values to be stored.
+We will use the hash table with an index size of 10,000,000 for further
+experiments, as it performed best in this case, and seems suitable for other
+cases (which require storing more values) as well.
 
@@
 times as much time and a lot more memory.
+
+The increased memory requirements of the Bender set can in part be explained by
+our implementation of variable-length values, which wastes some space by storing
+them in fixed-size slots. This overhead should be less than 25\% on average
+(and at most 50\% in the worst case).
 

### 2008-06-01 paper/paper.tex
Created final draft version.

@@
 
-\title{Evaluation of a Cache-Oblivious Data Structure [draft-1]}
+\title{Evaluation of a Cache-Oblivious Data Structure [draft-2]}
 
@@
 However, like every model, it is a simplification of reality.
-One of the simplifications is the assumption that storing and retrieving data in memory is a constant-time operation (i.e.
-the uniform memory model).
+One of the simplifications is the assumption that data can be stored or retrieved at any location in a constant amount of time, which is why we will call this the \emph{uniform memory model}.
 Advancements in software and hardware design over the last two de\-cades have caused this assumption to be increasingly detached from reality.
@@
 \centering
-\includegraphics{memhier}
+\includegraphics[width=55mm]{memhier}
 \caption{Schematic depiction of the memory hierarchy}\label{fig-memhier}
@@
 \begin{list}{}{}
-\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existence of a memory hierarchy, and may therefore not perform optimally in practice.
-\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
-\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (explained below) which does not allow para\-metri\-zation with properties of the cache layout.
+\item \textbf{Cache-unaware} data structures and algorithms are designed for
+the traditional uniform memory model, with no consideration of the existence
+of a memory hierarchy.
+\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and
+algorithms are designed to perform optimally in the external memory model
+(described below) but require parametrization with some or all properties of
+the cache layout (such as block size or cache memory size).
+\item \textbf{Cache-oblivious} data structures and algorithms are designed to
+perform optimally in the cache-oblivious model (explained below) which does
+not allow para\-metri\-zation with properties of the cache layout.
 \end{list}
 
-Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice.
-Although some research has been done on cache-oblivious data structures, at the time of writing there is no evidence of their use in practice.
+A great number of cache-unaware and cache-aware data structures has been developed and these are also widely used in practice.
+Although some research has been done on cache-oblivious data structures, there is currently no evidence that they are used in practice.
+Consequently, it is unclear if they are suitable for practical use at all.
 The main goal for this paper is to give some insight in the practical merit of cache-oblivious data structures.
 
-In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious model is of particular interest.
+In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious mod\-el is of particular interest.
 We will then describe what our goals for this research project were and how our work relates to previous research.
@@
 \section{Previous Work}
-The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model.
-We will briefly describe some of these alternatives here.
+The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of the alternatives.
 
@@
 Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
-Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory addresses is described by a non-decreasing function of these addresses, which means that accessing data at higher addresses can be slower than at lower addresses.
-This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where and that data at all locations can be accessed directly, which is usually not the case in practice.
-As a result, the applicability of their model to the design and evaluation of \emph{practical} algorithms is limited.
+Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this
+shortcoming by introducing a hierarchical memory model in which the cost of
+accessing values at different memory addresses is described by a non-decreasing
+function of these addresses, which means that accessing data at higher
+addresses can be slower than at lower addresses.
+This is a very general model that can be applied to the real memory hierarchy,
+but it assumes that the application has full control over which data is placed
+where, which is usually not the case in practice.
+As a result, the applicability of their model to the design and evaluation of
+practical algorithms is limited.
 
@@
 
-The lack of unrealistic requirements combined with the promise of cache-efficiency accross multiple levels of the memory hierarchy, makes it an interesting model for the development of low-maintenance, high-performance data structures for use in real-world applications.
-
-\section{Contributions}
+The cache-oblivious model is very similar to the real memory hierarchy,
+which means that algorithms designed for this model can
+easily be implemented in practice. This property, combined with the
+promise of cache-efficiency across multiple levels of the memory hierarchy,
+makes it a promising model for the development of low-maintenance,
+high-performance data structures for use in real-world applications.
+
+\section{Research Goals}
 Several cache-oblivious data structures and algorithms have been proposed.
@@
 This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice.
-To determine if cache-oblivious data structures have practical merit, empirical performance data is required, which is scarce as existing research has focused mainly on theoretical analysis.
-This paper addresses the question by reporting on the performance of newly implemented (but previously described) cache-oblivious data structures and their traditional counterparts (both cache-aware and cache-unaware data structures).
+To determine if cache-oblivious data structures have practical merit, empirical performance data is required, which is scarce, as existing research has focused mainly on theoretical analysis.
+This paper addresses the question by reporting on the performance of a newly
+implemented (but previously described) cache-oblivious data structure and two
+of its traditional counterparts (both cache-aware and cache-unaware data
+structures).
 
@@
 This data structure will be explained in detail in Section \ref{sect-bender-set}.
-The authors note that a chief advantage of their data structure over the previously described on is that it is less complex, more easily implementable and therefore more suitable for practical use.
+The authors note that a chief advantage of their data structure over the
+previously described one is that it is less complex, more easily implementable
+and therefore more suitable for practical use.
 
@@
 
-Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory.
-Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.
+Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize
+separate-chaining hash tables for cache efficiency by storing the linked lists
+that contain the contents for a bucket in contiguous memory. Their experiments
+show a performance gain over traditional methods, especially when the hash table
+is heavily loaded.
 
@@
 We also need to define more accurately what we mean by practical performance.
-We will execute our test application and measure two properties: primarily, the total running time, and secondarily the memory used.
-The rationale for selecting these metrics is that if the data structures perform identical functions and plenty of memory is available, the only observable difference in running a program using different data structures will be the run time.
-Memory use is of secondary interest because in practice memory may be limited, which would preclude the use of certain data structures that require a lot of memory to function.
+Our experiments are performed by running a test application (which will be
+described in detail below) and measuring two properties: primarily the
+execution time, and secondarily the memory in use.
+The rationale for selecting these metrics is that if the data structures perform
+identical functions and enough memory is available, the only observable
+difference in running a program using different data structures will be the
+execution time.
+Memory use is of secondary interest because in practice memory may be limited,
+which would preclude the use of data structures that require a large amount of
+memory to function.
 
 \subsection{State Space Search}
-The test application that we use to gather performance data is an implementation
-of a state space search algorithm.
+The test application that we used to gather performance data implements a state
+space search algorithm.
 This is a suitable scenario for two reasons.
@@
 State space search can be used to verify the correctness of software programs.
-For this purpose, programs are first modeled using a formal language, that is also used to specify properties of the program that should hold during its
+For this purpose, programs are first modeled using a formal language, that is
+also used to specify properties of the program that should hold during its
 execution. An executing program can be in a (possibly infinite) amount of states,
@@
 algorithm to generate all of these states, in order to check that desired
-properties hold in all of them. For this approach it is often required that
-the state space be finite (for obvious reasons).
-
-\begin{program}
+properties hold in all of them. This approach usually requires the state state
+space to be finite, although exhausting the entire state space is not necessary
+for our experiments.
+
+\begin{program}[t]
 \begin{verbatim}
@@
 
-Although variations are possible, our queue operates by a first-in, first-out
-principle, meaning that meaning that all states that are reachable in $N$ steps
+Although other behaviour is possible, our queue operates by a first-in,
+first-out principle, meaning that all states that are reachable in $N$ steps
 from the initial state are expanded before any states that require more than
@@
 accomplish much by itself; instead, the real work is done by the successor
-function and the queue and the set data structures. Queues can easily be
+function and the queue and set data structures. Queues can easily be
 implemented efficiently (adding and removing states takes O(1) time).
@@
 \subsection{Experimental Framework}
-For our experimental framework, we need a collection of formal models that are
+For our experimental framework, we needed a collection of formal models that are
 representative of those typically used for formal verification, and a way to
@@
 
-For the execution of these models we will use the NIPS VM \cite{weber2007evm}, a
+For the execution of these models we used the NIPS VM \cite{weber2007evm}, a
 high-performance virtual machine for state space generation that is easily
@@
 The NIPS VM represents program state as a binary string; the size of the state
-depends amongst other on the number of active processes in the program (which
-may change over the execution of the program).
+depends (amongst others) on the number of active processes in the program,
+which may change over the execution of the program.
 A typical state size is in the order of a few hundred bytes.
@@
 Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
-\item \textbf{Cache-aware}: B-trees.
-The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
+\item \textbf{Cache-aware}: B-trees. The B-tree is the standard choice for
+storing (ordered) data on disk, and depends on a page size being selected that
+corresponds with the size of data blocks that can be efficiently transferred.
 \item \textbf{Cache-oblivious}: the data structure proposed by Bender, Duan, Iacono and Wu.
@@
 our knowledge there are no high-quality implementations publicly available.
-The Bender set will therefore have to be implemented from scratch.
+The Bender set therefore had to be implemented from scratch.
 
@@
 undesirable to use existing libraries for a fair comparison, for two reasons.
-First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations.
+First, many existing implementations support additional operations (such as locking, synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations.
 Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not.
@@
 \subsubsection{Common Implementation Details}
-For memory management, the data structures make use of memory mapped files.
-The B-tree and hash table implementations use a single file for all data storage;
-the Bender tree uses several files (as discussed below) and some main memory.
-
-The use of memory mapped files means that the operating system is in charge of
-when and which pages are transferred between main memory and disk storage; the
-application assumes no control over this policy. This is in accordance with the
-cache-oblivious model, that assumes no control over the cache is possible.
-Additionally, memory mapped files bypass the C library's allocation functions
-and give the implementer complete control over how data is ordered in (virtual)
-memory.
+All data structures were implemented in C using a common interface.
+For memory management, the data structures make use of a single memory mapped
+data region, bypassing the C library's allocation functions and giving the
+implementer complete control over how data is arranged in memory.
+
+In principle, this also means the operating system has control over when and
+which data pages are transfered from main memory to disk storage and back.
+However, our experiments (which were performed on a system without a swap file)
+were limited to using main memory only.
 
@@
 However, we did not change the design of the data structures to take advantage
-of this. That means that any additional operations could be implemented without
-structural changes to our existing implementation.
+of the reduced functionality. That means that additional operations could be
+implemented without structural changes to our existing implementation.
 
@@
 index array (the \emph{index}) with a fixed number of slots. A \emph{hash
-function} is used to map key values onto slot numbers. If the slot for a value
+function} is used to map values onto slot numbers. If the slot for a value
 is empty, it may be inserted there. Queries for the existence of an element in
@@
 which is problematic if each slot can only store one value. There
-are a lot of different ways to resolve this collision problem; we use
+are many different ways to resolve this collision problem; we use
 \emph{separate chaining}, which means that slots do not store values directly,
@@
 relatively good performance when the number of values stored exceeds the size
-of the index array \cite{sedgewick1998ac}.
-
-\begin{figure}
+of the index array \cite{sedgewick1998ac}. Figure \ref{fig-hash-table} shows a
+hash table (with an index size of four, storing three values) and the way it
+is stored in consecutive memory.
+
+\begin{figure}[t]
 \centering
@@
 deal with holes that would otherwise occur in the stored file.
-Figure \ref{fig-hash-table} shows a hash table (with an index size of four,
-storing three values) and the way it is stored in consecutive memory.
 
@@
 B-trees are similar to other search tree data structures in the sense that
-they store ordered values hierarchically in such a way that each subtree store
-values that are close together.
-A key property of B-trees is that they do not fix the number of children per
-node, but instead organize data in pages of a fixed size, each containing
-as many values as will fit. This makes them especially suitable for purposes
-of on-disk storage where reading and writing data in a few large blocks is
-relatively cheap compared to accessing several smaller blocks.
-
-\begin{figure}
+they store ordered values hierarchically in such a way that every subtree stores
+consecutive values.
+A key property of B-trees is that (unlike most in-memory tree structures) they
+do not fix the number of children per node, but instead organize data in pages
+of a fixed size, each containing as many values as will fit.
+This makes them especially suitable for purposes of on-disk storage where
+reading and writing data in a few large blocks is relatively cheap compared to
+accessing several smaller blocks.
+
+\begin{figure}[b]
 \centering
@@
 the third leaf page are greater than ``Philip''. Note that since the page
-size is fixed, not all pages can be completely filled.
+size is fixed, not all pages are completely filled.
 
@@
 The first implementation challenge for the Bender set is that the description
-given by Bender et al. assumes that all values stored in the set are of a
+given by Bender et al assumes that all values stored in the set are of a
 fixed size, which is not the case in our experimental framework. To work
@@
 its size is rounded up to a power of two and it is inserted in the
-corresponding set instance. This ensures that the amount of spaces wasted
+corresponding set instance. This ensures that the amount of space wasted
 remains below 50\% while still allowing values of various sizes to be inserted.
-In the following description, we will assume that values are of a fixed size.
-
-The implementation of the Bender set consists of two parts: a sparse array
-storing the values in order, and a binary tree stored in van Emde Boas
-layout (similar to the one proposed by Prokop) that is used to find values
+The following description will be of a single Bender set instance, and it will
+therefore be assumed that values do have a fixed size.
+
+A Bender set has a capacity $C$ that is a power of two. Its implementation
+consists of two parts: a sparse array storing the values in order, and a binary
+tree stored in van Emde Boas layout that is used to search for values
 efficiently.
 Both the tree and the array may contain special \emph{empty values}.
-A depiction of a Bender set storing four values (with a capacity for eight) is
-given in Figure \ref{fig-bender-set}, with the window population counts for
-the sparse array given on the left (density thresholds not shown) and the index tree for the same array on the right.
-
-\begin{figure}
-\centering
-\includegraphics[width=83mm]{bender-set}
-\caption{A depiction of a Bender set}
-\label{fig-bender-set}
-\end{figure}
-
-A Bender set is created with a capacity $C$ that is a power of two.
+
 Initially, the array stores only empty values and is partitioned into
@@
 $\delta$.
+
+A depiction of a Bender set storing four values (with a capacity for eight) is
+given in Figure \ref{fig-bender-set}, with the window population counts for
+the sparse array given on the left (density thresholds not shown) and the index
+tree on the right.
+
+\begin{figure}
+\centering
+\includegraphics[width=83mm]{bender-set}
+\caption{A depiction of a Bender set}
+\label{fig-bender-set}
+\end{figure}
 
@@
 
-To insert a value, the binary tree is used to find a position in the array
-where the value may be inserted (just before its successor). Then, the windows
-overlapping this position are considered from bottom to top, and the first
-window that can support another element without overflowing is selected, and
-then rebalanced (thereby resolving the overflow in all lower level windows).
-
-In the case that all windows overlapping the target position would overflow upon
-insertion of the new element, the capacity of the data structure is doubled,
-a new top level is created and then the entire array is rebalanced. This is how
-the set grows when the number of elements stored comes too close to the storage
-capacity. Since changing the capacity causes all values in the array to be moved
-and requires that the entire index tree is recreated, this is an expensive
-operation; however, it occurs only infrequently.
+To insert a value, the index tree is used to find the position of the smallest
+existing value that is greater than the new value (i.e. the value's successor);
+the new value will be inserted right before its successor.
+Then, the windows overlapping the goal position are considered from bottom to
+top. The lowest (smallest) window that can support another element without
+overflowing is selected, and then rebalanced (thereby resolving the overflow in
+the overlapping lower level windows).
+
+If all windows, including the topmost window spanning the entire array,
+would overflow upon insertion of the new element, the capacity of the data
+structure must be increased to make room for more elements.
+When this happens, the capacity is doubled, a new top level is
+created and then the entire array is rebalanced.
+Since increasing the capacity causes all values in the array to be moved
+and the entire index tree to be recreated, this is an expensive operation;
+however, it only occurs infrequently.
 
@@
 
-One thing that is not specified in the paper by Bender et al, is how the
+A detail that is not specified in the paper by Bender et al, is how the
 population count for windows is kept. The simplest option is to keep no such
@@
 $N$ nodes is divided in half vertically, creating a small subtree at the
-top and a number of subtrees (approximately $\sqrt{N}$) below. Each subtree
-has a size of approximately $\sqrt{N}$ nodes and will be stored in a
-contiguous block of memory.
+top and a number of subtrees (approximately $\sqrt{N}$) at the bottom.
+Each subtree has a size of approximately $\sqrt{N}$ nodes and is stored in
+van Emde Boas layout in a contiguous block of memory; these blocks are then
+stored consecutively.
 
@@
 labeled with their positions according to the van Emde Boas layout.
-In this example, only two levels of recursion are needed. If each page
-stores three nodes, the path from node 1 to node 12 (or any other leaf node)
-requires visits only two pages.
+In this example, only two levels of recursion are needed. If we suppose that
+every page stores three nodes, then the highlighted path from node 1 to 12
+visits only two pages.
 
@@
 This tree can be used in a similar way as a binary search tree: by examining the
-two children of a node, we can determine whether a searched-for value lies in the
-left or right subtree.
+two children of a node, we can determine whether a searched-for value belongs
+in the left or right subtree.
+See the right side of Figure \ref{fig-bender-set} for an example.
 
 Note that the \emph{structure} of the tree is static and only changes when the
-capacity of the set is increased (in this case, the tree is recreated from scratch).
+capacity of the set is increased (in which case the tree is recreated from scratch).
 Of course, the values stored in the tree have to be updated when the corresponding
@@
 multiple executions. However, these values do not account for the time the
-system spends waiting (for example, for data to read from disk)
+system spends waiting (for example, for data to be read from disk)
 or the time spent by other processes on behalf of the executing process
@@
 
-Therefore, we decided to measure wall clock time. To deal with variations
-across multiple executions, each experiment is run seven times, and the median
-value is used for further analysis.
+Therefore, we decided to measure wall clock time and deal with variations
+across multiple executions by running each experiment seven times and using
+the median value for further analysis.
 
@@
 \subsubsection{System Configuration}
-The experiments where performed on a 64-bit Linux (kernel 2.6) system
+The experiments where performed on a 64-bit Linux system (kernel version 2.6.18)
 with an Intel Xeon E5335 processor (2~GHz, 4~MB cache) and 8~GB of main
@@
 tested. A new process is created for every prime number found, and as a result
-the state space can only by exhaustively searched for relatively small values
+the state space can only be exhaustively searched for relatively small values
 of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
@@
 indication of the ratio of value look-ups (insertions of values that are already
-present in the set) and actual insertions, ranging from approximately from 4 to
+present in the set) and actual insertions, ranging approximately from 4 to
 2.7 look-ups per insertion.
@@
 As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
-usage by the different hash tables is in the (fixed) size of the index.
-We will use the hash table with an index size of 10,000,000 for further
+usage of the different hash tables is in the (fixed) size of the index.
+We will use the hash table with an index size of 10 million slots for further
 experiments, as it performed best in this case, and seems suitable for other
@@
 \ref{fig-eratosthenes-btree_vss} show the execution time and memory usage
-for the B-tree with page sizes ranging from 1~KB to 16~KB. The execution platform
-maps memory in 4~KB pages, which would suggest that using pages less than 4KB
+for the B-tree with page sizes ranging from 1~kilobyte to 16~kilobyte.
+The execution platform
+maps memory in 4~KB pages, which would suggest that using pages less than 4~KB
 makes little sense. Indeed, the B-tree with 1~KB pages seems to perform worst,
-but if we take the memory use into account, this is more likely due to the fact
-that fewer items fit on a single page (which means a relatively large portion of
-the page remains unused) than anyting else.
-The 4~KB and 16~KB pages are relatively close (both in execution time and memory
-usage); we select the 16~KB page size for further experiments because state sizes
-are larger in other test cases, in which case the 4~KB page size may cause similar
-problems as the 1~KB page size here.
+but if we take the memory use into account, this is most likely due to the fact
+that fewer items fit on a single page, which means a relatively large portion of
+the page remains unused.
+The difference between 4~KB and 16~KB pages is relatively small (both in
+execution time and memory usage); we select the 16~KB page size for further
+experiments because state sizes are larger in other test cases, in which case
+the 4~KB page size may cause similar problems as the 1~KB page size here.
 
@@
 that is available. We select $\delta=0.5$ for future experiments, which strikes
-a good balance between execution time and required memory.
+a good balance between execution time and memory requirements.
 
@@
 we can run final experiments on all three test cases.
-The execution times for the these cases are presented in figures
-\ref{fig-eratosthenes_wctime}, \ref{fig-leader2_wctime} and
-\ref{fig-peterson_wctime}, while the corresponding memory usage is presented in
-figures \ref{fig-eratosthenes_vss}, \ref{fig-leader2_vss} and \ref{fig-peterson_vss} respectively.
+The execution times for these cases are presented in figures
+\ref{fig-eratosthenes_wctime}--\ref{fig-peterson_wctime}
+and the corresponding memory usage is presented in
+figures \ref{fig-eratosthenes_vss}--\ref{fig-peterson_vss}.
 
@@
 when the page size is largest. Unfortunately, even then it requires about three
-times as much time and a lot more memory.
+times as much time (and 5--6 times as much memory).
 
@@
 our implementation of variable-length values, which wastes some space by storing
-them in fixed-size slots. This overhead should be less than 25\% on average
-(and at most 50\% in the worst case).
-
-The test cases used all have a relatively high percentage of insertions relative
-to value look-ups. This may explain the good performance of the hash table (for
+them in fixed-size slots. This overhead should be around 25\% on average.
+
+The test cases used all have a relatively high ratio of value insertions to
+look-ups. This may explain the good performance of the hash table (for
 which insertions are barely more expensive than look-ups) as well as the bad
 performance of the Bender tree (for which insertions are relatively expensive,
-especially when a window is rebalanced). Unfortunately, the insert/look-up
-ratio for our test cases varies too little to derive a meaningful conclusion
-in this regard.
+especially when a window is rebalanced). Unfortunately, our experiments do not
+give enough data to determine how the relative performance of the data
+structures changes when this ratio changes.
 
 \subsection{Implementation Complexity}
-Since all data structures used were implemented from scratch, our research
-also yields some insight in the implementation complexity of the different
-data structures.
-
-\begin{table}
+Since all data structures included in the experiments were implemented from
+scratch, our research also yields some insight in the implementation complexity
+of the different data structures.
+
+Table \ref {tab-loc} gives the number of lines of source code used to implement
+various part of the test application, after removing comments and blank lines
+(which amount to about 25\% of the code).
+Common code includes allocation functions, interface descriptions and comparison
+and hashing functions. The framework includes not just the search algorithm,
+but also the functionality to report various metrics while running a test case.
+
+\begin{table}[t]
 \begin{center}
@@
 
-Table \ref {tab-loc} gives an overview of the source lines of code used for
-the test application, after removing comments and blank lines (which together
-amount for about 25\% of the code). Common code includes allocation functions,
-interface descriptions and comparison and hashing functions. The framework
-includes not just the search algorithm, but also the functionality to report
-various metrics while running a test case.
-
 Although not a perfect metric of implementation complexity, the lines of code
 required for the various components of the test application do give some
-insight in the relative complexity of the data structures. It is apparent why
-hash tables are popular: they perform well while they are relatively easy to
-implement.
+insight in the relative complexity of the data structures. It is clear why
+hash tables are a popular choice: they are easy to implement yet perform very
+well. The Bender set not only uses more lines of code, but in our experience
+also required a greater amount of effort to implement correctly.
 
@@
 To better understand how different factors influence performance, it would
-be interesting to measure cache efficiency seperately and reporton actual
+be interesting to measure cache efficiency separately and report on actual
 cache hits and misses on different boundaries of the memory hierarchy.
@@
 cache-oblivious data structure has been examined, in a single test scenario,
-on a single hardware
-and software platform.
-It is possible that in a different setting, different cache-oblivious data
-structures do compare favorably to their traditional counterparts.
+on a single platform.
+In a different setting, different cache-oblivious data structures might compare
+favorably to their traditional counterparts.
 However, since the observed differences in performance are fairly large,
@@
 
-We hope that future research will focus not only on the theoretical performance
-of newly developed algorithms and data structures, but also on the performance
-in practice. Theoretical results are invaluable, but to be of greatest
-significance newly developed data structures and algorithms should have
-demonstrable practical merit as well.
+We hope that future research on cache-oblivious data structures will not focus
+on theoretical performance alone, but will also compare performance in practice
+with existing alternatives.
+Although theoretical results are invaluable, newly developed data structures and
+algorithms should, preferably, have demonstrable practical merit as well.
 

### 2008-06-01 paper/paper.tex
Spell checked.

@@
 Traditionally, data structures are evaluated in a theoretical model that
-does not take the existance of a memory hierarchy into account. The
+does not take the existence of a memory hierarchy into account. The
 cache-oblivious model has been proposed as a more accurate model.
@@
 
-In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious mod\-el is of particular interest.
+In the following pages we will give an overview of the different available memory models and explain why the cache-obliv\-i\-ous mod\-el is of particular interest.
 We will then describe what our goals for this research project were and how our work relates to previous research.
@@
 \section{Previous Work}
-The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of the alternatives.
+The cache-oblivious memory model is not the only or the first mod\-el that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of the alternatives.
 
@@
 In their experimental results the cache-oblivious tree performs worse than the (non-oblivious) alternatives.
-Nevertheless, they conclude that \textit{``cache-obli\-vious data structures may have significant practical importance''}.
+Nevertheless, they conclude that \textit{``cache-oblivious data structures may have significant practical importance''}.
 
@@
 
-Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-obli\-vious priority deque.
-Their main result is that although the cache-obli\-vious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
+Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-oblivious priority deque.
+Their main result is that although the cache-obliv\-i\-ous data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.
 
@@
 
-Although other behaviour is possible, our queue operates by a first-in,
+Although other behavior is possible, our queue operates by a first-in,
 first-out principle, meaning that all states that are reachable in $N$ steps
@@
 In principle, this also means the operating system has control over when and
-which data pages are transfered from main memory to disk storage and back.
+which data pages are transferred from main memory to disk storage and back.
 However, our experiments (which were performed on a system without a swap file)
@@
 program (called \emph{wall clock time}).
-This has the disadvantage that concurrently executing processes affect the runtime
-of the process being measured, so on a busy system the reported runtime may vary
+This has the disadvantage that concurrently executing processes affect the timing
+of the process being measured, so on a busy system the reported time may vary
 over different executions.

### 2008-06-02 paper/paper.tex
Changed plots and figures so all plots fit on a single page.

@@
 \usepackage{url}
+\usepackage{subfigure}
 
@@
 \end{verbatim}
-\caption{Pseudo-code for a simple state search algorithm.}
+\caption{Pseudo-code for a simple state search algorithm}
 \label{prog-search}
@@
 \includegraphics[width=83mm]{hash-table}
-\caption{A depiction of a separate-chaining hash table}
+\caption{Depiction of a separate-chaining hash table}
 \label{fig-hash-table}
@@
 \includegraphics[width=83mm]{b-tree}
-\caption{A depiction of a B-tree}
+\caption{Depiction of a B-tree}
 \label{fig-b-tree}
@@
 \includegraphics[width=83mm]{bender-set}
-\caption{A depiction of a Bender set}
+\caption{Ddepiction of a Bender set}
 \label{fig-bender-set}
@@
 we can run final experiments on all three test cases.
-The execution times for these cases are presented in figures
-\ref{fig-eratosthenes_wctime}--\ref{fig-peterson_wctime}
-and the corresponding memory usage is presented in
-figures \ref{fig-eratosthenes_vss}--\ref{fig-peterson_vss}.
+The execution times for these cases are presented in Figure \ref{fig-wctime}
+and the corresponding memory usage is presented in Figure \ref{fig-vss}.
 
@@
 
-%\begin{figure}
-%\centering
-%\includegraphics[width=83mm]{fig-queuesize}
-%\caption{Queue size}
-%\label{fig-quesize}
-%\end{figure}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=83mm]{fig-transitions}
-%\caption{Transitions}
-%\label{fig-transitions}
-%\end{figure}
-
-\begin{figure}[p]
+\begin{figure*}[p]
 \centering
-\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
-\caption{Execution time of hash table on Eratosthenes}
+\subfigure[B-tree]{
     \label{fig-eratosthenes-hash_wctime}
-\end{figure}
-
-\begin{figure}[p]
+    \includegraphics[width=56mm]{fig-eratosthenes-hash_wctime}}
+\subfigure[Hash table]{
+    \label{fig-eratosthenes-btree_wctime}
+    \includegraphics[width=56mm]{fig-eratosthenes-btree_wctime}}
+\subfigure[Bender set]{
+    \label{fig-eratosthenes-bender_wctime}
+    \includegraphics[width=56mm]{fig-eratosthenes-bender_wctime}}
+\caption{Execution times per data structure on Eratosthenes}
+\label{fig-eratosthenes-wctime}
+\end{figure*}
+
+\begin{figure*}[p]
 \centering
-\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
-\caption{Execution time of B-tree on Eratosthenes}
-\label{fig-eratosthenes-btree_wctime}
-\end{figure}
-
-\begin{figure}[p]
+\subfigure[B-tree]{
+    \label{fig-eratosthenes-hash_vss}
+    \includegraphics[width=56mm]{fig-eratosthenes-hash_vss}}
+\subfigure[Hash table]{
+    \label{fig-eratosthenes-btree_vss}
+    \includegraphics[width=56mm]{fig-eratosthenes-btree_vss}}
+\subfigure[Bender set]{
+    \label{fig-eratosthenes-bender_vss}
+    \includegraphics[width=56mm]{fig-eratosthenes-bender_vss}}
+\caption{Memory usage per data structure on Eratosthenes}
+\label{fig-eratosthenes-vss}
+\end{figure*}
+
+\begin{figure*}[p]
 \centering
-\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
-\caption{Execution time of Bender set on Eratosthenes}
-\label{fig-eratosthenes-bender_wctime}
-\end{figure}
-
-\begin{figure}[p]
+\subfigure[Eratosthenes]{
+    \label{fig-eratosthenes_wctime}
+    \includegraphics[width=56mm]{fig-eratosthenes_wctime}}
+\subfigure[Leader2]{
+    \label{fig-leader2_wctime}
+    \includegraphics[width=56mm]{fig-leader2_wctime}}
+\subfigure[Peterson\_N]{
+    \label{fig-peterson_wctime}
+    \includegraphics[width=56mm]{fig-peterson_wctime}}
+\caption{Execution times per test case}
+\label{fig-wctime}
+\end{figure*}
+
+\begin{figure*}[p]
 \centering
-\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
-\caption{Memory usage of hash table on Eratosthenes}
-\label{fig-eratosthenes-hash_vss}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
-\caption{Memory usage of B-tree on Eratosthenes}
-\label{fig-eratosthenes-btree_vss}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
-\caption{Memory usage of Bender set on Eratosthenes}
-\label{fig-eratosthenes-bender_vss}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes_wctime}
-\caption{Execution time on Eratosthenes}
-\label{fig-eratosthenes_wctime}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-leader2_wctime}
-\caption{Execution time on Leader2}
-\label{fig-leader2_wctime}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-peterson_wctime}
-\caption{Execution time on Peterson\_N}
-\label{fig-peterson_wctime}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-eratosthenes_vss}
-\caption{Memory usage on Eratosthenes}
+\subfigure[Eratosthenes]{
     \label{fig-eratosthenes_vss}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-leader2_vss}
-\caption{Memory usage on Leader2}
+    \includegraphics[width=56mm]{fig-eratosthenes_vss}}
+\subfigure[Leader2]{
     \label{fig-leader2_vss}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=83mm]{fig-peterson_vss}
-\caption{Memory usage on Peterson\_N}
+    \includegraphics[width=56mm]{fig-leader2_vss}}
+\subfigure[Peterson\_N]{
     \label{fig-peterson_vss}
-\end{figure}
+    \includegraphics[width=56mm]{fig-peterson_vss}}
+\caption{Memory usage per test case}
+\label{fig-vss}
+\end{figure*}
 

### 2008-06-06 paper/paper.tex
Some final fixes.
Small formatting changes to the paper.

@@
 
-\title{Evaluation of a Cache-Oblivious Data Structure [draft-2]}
+\title{Evaluation of a Cache-Oblivious Data Structure}
 
@@
 A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model.
-In this model, there is an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory.
-Data is transferred between the two in aligned data blocks of size B.
+In this model, there is an infinitely large external memory and an internal memory of size $M$ which operates as a cache for the external memory.
+Data is transferred between the two in aligned data blocks of size $B$.
 In contrast with the hierarchical memory model, the application does not have explicit control over the transferring of blocks between the two memories.
 Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program.
-Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and data structures.
-Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).
+Additionally (and in contrast with the external memory model) the values of parameters like $M$ and $B$ are not known to the application, so they cannot be used explicitly when defining algorithms and data structures.
+Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of $M$ and $B$ (and other parameters relevant to the problem).
 
@@
 used in the performance evaluation have been implemented from scratch.
-
+\clearpage
 \subsubsection{Set Operations}
@@
 \includegraphics[width=83mm]{bender-set}
-\caption{Ddepiction of a Bender set}
+\caption{Depiction of a Bender set}
 \label{fig-bender-set}

### 2008-06-06 search/search.c
Applied patch by Michael Weber that allows reporting on Mac OS X.

@@ -1,8 +1,16 @@
 #include "search.h"
 #include <assert.h>
 #include <stdio.h>
-#include <nips_vm/nipsvm.h>
 #include <sys/time.h>
+#include <unistd.h>
+#if defined(__APPLE__)
+#include <mach/task.h>
+#include <mach/mach_init.h>
+#include <mach/shared_memory_server.h>
+#elif defined(__linux__)
+#include <asm/param.h>                  /* HZ */
+#endif
+#include <nips_vm/nipsvm.h>
 
 const int PAGESIZE = 4096;
 
@@ -112,29 +120,59 @@ void report(FILE *fp, SearchContext *sc)
 {
     FILE *stat;
     int res;
-    unsigned long utime, stime, vsize;
+    unsigned long vsize;
+    double utime, stime;
     long rss;
-
+#if defined(__linux__)
+    unsigned long lutime, lstime;
     stat = fopen("/proc/self/stat", "rt");
     assert(stat != NULL);
     res = fscanf( stat, "%*d %*s %*c %*d %*d %*d %*d %*d %*u "
                         "%*lu %*lu %*lu %*lu "
                         "%lu %lu %*ld %*ld %*ld %*ld %*ld %*ld %*llu "
                         "%lu %ld",
-                  &utime, &stime, &vsize, &rss );
+                  &lutime, &lstime, &vsize, &rss );
+    rss *= PAGESIZE;
+    utime = lutime / (double)HZ;
+    stime = lstime / (double)HZ;
     assert(res == 4);
     fclose(stat);
+#elif defined(__APPLE__)
+    task_t task = MACH_PORT_NULL;
+    if (task_for_pid (current_task(), getpid(), &task) != KERN_SUCCESS) {
+        perror ("task_for_pid");
+        return;
+    }
+    struct task_basic_info task_basic_info;
+    mach_msg_type_number_t task_basic_info_count = TASK_BASIC_INFO_COUNT;
+    task_info(task, TASK_BASIC_INFO, (task_info_t)&task_basic_info,
+              &task_basic_info_count);
+    rss = task_basic_info.resident_size / 1024UL;
+    unsigned long int vmsize_bytes = task_basic_info.virtual_size;
+    if (vmsize_bytes > SHARED_TEXT_REGION_SIZE + SHARED_DATA_REGION_SIZE)
+        vmsize_bytes -= SHARED_TEXT_REGION_SIZE + SHARED_DATA_REGION_SIZE;
+    vsize = vmsize_bytes / 1024UL;
 
[…]

### 2008-06-06 datastructures/Alloc.c
Replaced mapping of /dev/zero by an anonymous mapping to increase portability.

@@ -50,7 +50,7 @@ void *Allocator_mmap(Alloc *a, void *old_data, size_t size)
     if (old_data == NULL)
     {
         assert(size > 0);
-        if (!FS_create(&a->fs, "/dev/zero"))
+        if (!FS_create(&a->fs, NULL))
             return NULL;
 
         return FS_resize(&a->fs, NULL, size);

### 2008-06-06 datastructures/FileStorage.c

@@ -17,10 +17,18 @@ bool FS_create(FileStorage *fs, const char *path)
 
     assert(ALLOC_CHUNK_SIZE > 0);
 
-    /* Open the file */
-    fd = open(path, O_CREAT | O_RDWR, 0666);
-    if (fd < 0)
-        return false;
+    if (path == NULL)
+    {
+        /* Map anonymously */
+	fd = -1;
+    }
+    else
+    {
+        /* Open the file */
+        fd = open(path, O_CREAT | O_RDWR, 0666);
+        if (fd < 0)
+            return false;
+    }
 
     /* Initialize FileStorage structure */
     fs->size       = 0;
@@ -34,7 +42,8 @@ void FS_destroy(FileStorage *fs, void *data)
 {
     if (data != NULL)
         munmap(data, fs->size);
-    close(fs->fd);
+    if (fs->fd != -1)
+        close(fs->fd);
 }
 
 void *FS_resize(FileStorage *fs, void *data, size_t size)
@@ -66,13 +75,12 @@ void *FS_reserve(FileStorage *fs, void *data, size_t size)
 
     /* Change file size */
     assert(sizeof(off_t) >= sizeof(size_t));
-    res = ftruncate(fs->fd, (off_t)new_size);
-    /* truncation may fail for special files like /dev/zero; in that case,
-       this not an error. (If it is, then mmap/mremap will fail later on) */
-    /*
-    if (res != 0)
-        return NULL;
-    */
+    if (fs->fd != -1)
+    {
+        res = ftruncate(fs->fd, (off_t)new_size);
+        if (res != 0)
+            return NULL;
+    }
 
     /* Remap memory */
     if (HAVE_MREMAP && data != NULL)
@@ -84,7 +92,8 @@ void *FS_reserve(FileStorage *fs, void *data, size_t size)
     {
         if (data != NULL)
             munmap(data, old_size);
-        data = mmap( NULL, new_size, PROT_READ|PROT_WRITE, MAP_PRIVATE,
+        data = mmap( NULL, new_size, PROT_READ|PROT_WRITE,
+	             fs->fd == -1 ? (MAP_PRIVATE|MAP_ANONYMOUS) : MAP_SHARED,
                      fs->fd, (off_t)0 );
         /* FIXME: if this fails, the old mapping is destroyed! */
     }

### 2008-06-06 search/main.c

@@ -136,7 +136,7 @@ int main(int argc, char *argv[])
     }
 
     /* Create deque data structure */
-    params.queue = File_Deque_create("/dev/zero");
+    params.queue = File_Deque_create(NULL);
     if (params.queue == NULL)
     {
         perror("Could not create deque");

### 2008-06-06 datastructures/config.h
Modified config.h so HAVE_MREMAP is set by default on Linux only.

@@ -1,3 +1,11 @@
 #define _GNU_SOURCE
-#define HAVE_MREMAP 1
+
+#ifndef HAVE_MREMAP
+#  ifdef __linux__
+#    define HAVE_MREMAP 1	
+#  else
+#    define HAVE_MREMAP 0
+#  endif
+#endif 
+
 #define ALLOC_CHUNK_SIZE 4096

### 2008-06-06 datastructures/FileStorage.c
Fixed mmap() code so it works on FreeBSD.

@@ -85,7 +85,9 @@ void *FS_reserve(FileStorage *fs, void *data, size_t size)
     /* Remap memory */
     if (HAVE_MREMAP && data != NULL)
     {
+#if HAVE_MREMAP
         data = mremap(data, old_size, new_size, MREMAP_MAYMOVE);
+#endif
         /* FIXME: If this fails, does the old mapping still exist? */
     }
     else
@@ -93,7 +95,7 @@ void *FS_reserve(FileStorage *fs, void *data, size_t size)
         if (data != NULL)
             munmap(data, old_size);
         data = mmap( NULL, new_size, PROT_READ|PROT_WRITE,
-	             fs->fd == -1 ? (MAP_PRIVATE|MAP_ANONYMOUS) : MAP_SHARED,
+                     fs->fd == -1 ? (MAP_PRIVATE|MAP_ANON) : MAP_SHARED,
                      fs->fd, (off_t)0 );
         /* FIXME: if this fails, the old mapping is destroyed! */
     }
